### 2019-05-01 main.tex
initial install

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
%\geometry{landscape}                		% Activate for rotated page geometry
%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage{amssymb}

%SetFonts
\usepackage[T1]{fontenc}
\usepackage{lmodern}

%SetFonts

\title{Fast algebraic filtering of surfaces from 3D medical images with Julia}
\author{Miroslav Ji\v{r}\'ik and Alberto Paoluzzi}
\date{}							% Activate to display a given date or no date

\begin{document}
\maketitle

\begin{abstract}
In this paper we introduce a novel algebraic \textsc{lar-surf} filter, well founded on algebraic topology methods, to extract and smooth the boundary surface of any subset of voxels arising from the segmentation of a 3D medical image. The input is defined as a \emph{chain}, i.e.~as a vector from a linear space of 3-chains, represented in coordinates as a sparse Boolean array. The output is produced as the result of the mapping via the linear boundary operator $\partial_3:C_3 \to C_2$ between linear spaces of 3- and 2-chains.  In particular, when the input set of voxels is either not (4-)connected, or contains one or more empty regions inside, \textsc{lar-surf} generates a non connected set of closed surfaces, i.e.~a set of 2-cycles. The only data structures used are sparse arrays with one or two indices, i.e.~sparse vectors and matrices. This work is based on LAR (Linear Algebraic Representation) methods, and is implemented in Julia language, natively supporting parallel computing on hybrid architectures.
\end{abstract}

\tableofcontents

%\section{}
%\subsection{}

\include{intro}
\include{background}
\include{filter}
\include{examples}
\include{conclusion}

\end{document}

### 2019-05-01 src/filter.tex
added src dir

\section{Filter design and implementation}\label{sec:filter}

\subsection{Block-parametric design}\label{sec:aaaa}

\subsection{Julia implementation}\label{sec:aaaa}

\subsection{Code optimization}\label{sec:aaaa}

\subsection{Performance analysis}\label{sec:aaaa}

### 2019-05-01 src/main.tex
added src dir

--- 2019-05-01 main.tex
@@
 In this paper we introduce a novel algebraic \textsc{lar-surf} filter, well founded on algebraic topology methods, to extract and smooth the boundary surface of any subset of voxels arising from the segmentation of a 3D medical image.
-The input is defined as a \emph{chain}, i.e.~as a vector from a linear space of 3-chains, represented in coordinates as a sparse Boolean array.
+The input is defined as a \emph{chain}, i.e.~as a vector from a linear space of 3-chains, represented in coordinates as a sparse Boolean vector.
 The output is produced as the result of the mapping via the linear boundary operator $\partial_3:C_3 \to C_2$ between linear spaces of 3- and 2-chains.
-In particular, when the input set of voxels is either not (4-)connected, or contains one or more empty regions inside, \textsc{lar-surf} generates a non connected set of closed surfaces, i.e.~a set of 2-cycles.
-The only data structures used are sparse arrays with one or two indices, i.e.~sparse vectors and matrices.
+In particular, when the input set of voxels is either not (4-)connected, or contains one or more empty regions inside, \textsc{lar-surf} generates a non connected set of closed surfaces, i.e.~a set of 2-cycles---using the language of algebraic topology.
+The only data structures used by this approach are sparse arrays with one or two indices, i.e.~sparse vectors and matrices.
 This work is based on LAR (Linear Algebraic Representation) methods, and is implemented in Julia language, natively supporting parallel computing on hybrid architectures.

### 2019-05-09 src/background.tex
working on algorithm design

\section{Background}\label{sec:background}

\subsection{Linear Algebraic Representation}\label{sec:aaaa}

A \emph{representation scheme} for solid modeling is a mapping between a space of mathematical models and a space of symbolic representations, often generated by a formal grammar.
Solid pointsets (i.e., `$r$-sets') are defined~\cite{requicha:80} as compact (bounded and closed) regular and semianalytic subsets of the $d$-space. A large number of representation schemes were defined in the past forty years, including the two main classes of (a) \emph{boundary representations} (`$B$-reps'), where the solid model is represented through a representation of its boundary elements, i.e.~faces, edges and vertices, and (b) \emph{decompositive/enumerative representations}, that are a decomposition of either the object or the embedding space, respectively, into a well-defined cellular complex. In particular, a boundary representation provides a cellular decomposition of the object's boundary into cells of dimension zero (vertices), one (edges), and two (faces). Medical imaging is a kind of enumerative representation of organs and tissues of interest, as subsets of 3D volume elements (voxels) from the 3D image.

\subsection{Multiindices from Cartesian indices}\label{sec:aaaa}

\subsection{Taubin Smoothing}\label{sec:aaaa}

### 2019-05-09 src/filter.tex
working on algorithm design

@@
 
-\subsection{Block-parametric design}\label{sec:aaaa}
+\subsection{Block-parametric design}\label{sec:block}
 
-\subsection{Julia implementation}\label{sec:aaaa}
+\subsubsection{Block decomposition}\label{sec:bbbb}
 
-\subsection{Code optimization}\label{sec:aaaa}
+We assume that medical devices produce 3D images with lateral dimensions that are integer multiples of some powers of two, like 128, 256, 512, etc.
+Any cuboidal portion of image is completely determined by the Cartesian indices of its voxel of lowest and highest indices, and denoted as $Image([\ell_x, \ell_y, \ell_z],[h_x, h_y, h_z])$.
 
-\subsection{Performance analysis}\label{sec:aaaa}
+For the sake of simplicity, we assume a common size on the three image axes, and the corresponding image portion, called \emph{block} $B$, as a function of its element of the  lowest single block  coordinates $i,j,k\in\N$ and block dimension $n$:
+\[
+B(i,j,k,n) := Image([in,jn,kn],[i(n+1),j(n+1),k(n+1)]) 
+\]
+
+\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+   \centering
+   \includegraphics[width=2in]{figs/blocks} 
+   \caption{A possible block partitioning of a radiologic image. The evidenced 2D block is $B([2,1,64]) = Image([128,64],[172,128)]$}
+   \label{fig:blocks}
+\end{figure}
+
+Figure~\ref{fig:blocks} shows the block decomposition in a 2D case, with positive integer $u,v,w$ lateral sizes of image. Note that block sides do not necessarily correspond to image edges.
+
+\subsubsection{Block operator }\label{sec:bbbb}
+
+\paragraph{Chain coordinates }\label{sec:bbbb} 
+We are going to treat each image block independently from each other.
+Hence we map each image subspace $B(i,j,n)$ to the linear \emph{chain} space $C_2$ of dimension $n\times n$, using coordinate vectors $c_{h,k}\in \B^{n\times n} := \{0,1\}^{n\times n}$, where the basis element $c = c_{h,k} \in C_2$ is mapped via Cartesian-to-linear map to the Boolean vector
+\[
+Image(h,k) \mapsto c_{h,k} := [0 \cdots 0\ 1\ 0 \cdots 0] \in \B^{n\times n}
+\]
+for each $0\leq h,k \leq n$, and where the (single) unit element is in position $nk + h \leq n\times n$.
+
+Therefore, each pixel (or voxel) in a block image will be seen as a basis Boolean vector in $C_2$, and each subset of image elements, as the corresponding Boolean vector in $C_2$, with many ones as the cardinality of the subset.
+
+\paragraph{Boundary operator }\label{sec:bbbb} For a fixed block size $n$, the boundary operator $\partial_d : C_d\to C_{d-1}$, with $d\in\{2,3\}$, will be constructed once and for all using the algorithm given in~\cite{}, and inlined in the generated boundary extraction code.
+
+It is easy to see that the operator's matrix $[\partial_d]$ is \emph{very sparse}, since it contains $2\times d$ non-zero elements (ones) for each column (of length $n^d$), i.e.~4 ones and 6 ones for the 2D and 3D case, respectively.
+In fact the matrix of a linear operator between linear spaces contains by columns the basis element of the domain space, represented in the target space.
+In our case, the former is an image element (2-cube or 3-cube), represented as the chain of its boundary---i.e.
+either a 1-cycle of 4 edges, or  a 2-cycle of 6 faces, respectively.
+
+The number of rows of $[\partial_d]$ equates the dimension of the linear space $C_{d-1}$, i.e.~the number of $(d-1)$-cells---elementary $(d-1)$-chains---in the cellular partition of the image.
+To compute their number, we act in two steps.
+(a) First we map one-to-one the $n^d$ $d$-cells with $d$ adjacent $(d-1)$-cells, so getting $d\,n^d$ distinct basis elements of $C_{d-1}$.
+(b) Then we complete this bases by adjoining $n^{d-1}$ boundary elements for each of the $d$ dimensions of the image, so providing further $d\,n^{d-1}$ basis elements for $C_{d-1}$.
+The dimension of $C_{d-1}$, and therefore the number of rows of $[\partial_d]$ matrix is $d\,(n^{d-1}+n^{d}) = d\,n\,(1+n)^{d-1}$.
+The number of column equates the number of basis elements of $C_d$, i.e.~the number $n^d$ of block elements.
+
+\paragraph{Sparsity and size of boundary matrix }\label{sec:bbbb}
+
+As we have seen, we have $2d$ non-zero elements for each column of $[\partial_d]$, so that their total number is $2d\,n^d$.
+The number of matrix element is $d\,n\,(1+n)^{d-1} \times n^d$, giving a ratio of
+\[
+\frac{\mbox{non-zero\ elements}}{\mbox{total\ elements}} = 
+\frac{2d\times n^d}{d\,n\,(1+n)^{d-1} \times n^d} =
+\frac{2}{n+n^d}
+\]
+Using sparse matrices in CSC (Compressed Sparse Column) format we get a storage size:
+\[
+mem([\partial_d]_{n^d}) = 2\times \#\mbox{nzero} + \#\mbox{columns} = 2\times 2d\,n^d + n^d = (4d+1)n^d.
+\]
+In conclusion, for block size $n=64$, the matrix $[\partial_d]$ requires for 2D images $9\times 64^2=36,864$ memory elements, and for 3D images $13\times 64^3=3,407,872$ memory elements.
+Counting the bytes for the standard implementation of a sparse binary matrix (1 byte for values and 8 bytes for indices) we get $(18d+8)n^d$ bytes, giving $176$\,KB for 2D and $15.872$\,MB for 3D.
+
+\subsubsection{Block boundary mapping}\label{sec:bbbb}
+
+\subsubsection{Block parallelism}\label{sec:bbbb}
+
+\subsubsection{Reduced model}\label{sec:bbbb}
+
+\subsection{Julia implementation}\label{sec:implementation}
+
+\subsection{Code optimization}\label{sec:optimization}
+
+\subsection{Performance analysis}\label{sec:analysis}

### 2019-05-09 src/filter.tex
update filter.tex file

@@
 
-For the sake of simplicity, we assume a common size on the three image axes, and the corresponding image portion, called \emph{block} $B$, as a function of its element of the  lowest single block  coordinates $i,j,k\in\N$ and block dimension $n$:
+For the sake of simplicity, we assume a common size on the three image axes, and the corresponding image portion $B$, called \emph{block}, as a function of its element of the  lowest  block  coordinates $i,j,k\in\N$ and of block dimension $n$:
 \[
@@
 
-Figure~\ref{fig:blocks} shows the block decomposition in a 2D case, with positive integer $u,v,w$ lateral sizes of image. Note that block sides do not necessarily correspond to image edges.
+Figure~\ref{fig:blocks} shows the block decomposition in a 2D case, with positive integer $u,v$ lateral sizes of image. Note that block sides do not necessarily correspond to image edges.
 
@@
 
+Here we refer directly to the 3D case.
+Let us call \emph{segment} the bulk content $S$ of interest within the input 3D image of size $(u,v,w)$. We aim to compute the segment boundary $\partial_3 S$. 
+First we set the size $n$ of the block, in order to decompose the input $Image(u,v,w)$ into a fair number 
+\[
+M = \ceil{u/n} \times \ceil{v/n} \times \ceil{w/n}
+\] of blocks. 
+Then we consider each image portion $c_{i,j,k} = S\cap B(i,j,k,n)$ and compute its (binary) coordinate representation  $[c]_{i,j,k}\in C_3(n,n,n)$.
+This one is a sparse binary vector of length $n^3$.
+Then assemble the $M$ representations $c$ of segment portions into a sparse binary matrix $\T{S}$, of dimension $n^d \times M$.
+Finally compute a matrix $\T{B}$ of boundary portions of $S$, represented by columns as chain coordinate vectors in $C_2$:
+\[
+\T{B} = [\partial_3(n,n,n)]\, \T{S}.
+\]
+
+Of course, the $\T{B}$ sparse matrix has the same column number $M$ of $\T{S}$, because each column contains the boundary representation of the corresponding $S\cap B_{i,j,k}$, and the number of rows of the operator, equal to the dimension of the linear space $C_2$.
+
+\paragraph{Embedding}
+A final computational step is needed, in order to embed the 2-chains in Euclidean space $\E^3$ and to assemble the whole resulting surface.
+In particular, we need to compute the \emph{embedding function} $\mu : C_0 \to \E^3$, where $C_0$ is the space of 0-chains, one-to-one corresponding to the vertices of the extracted surface.
+The simplest solution is to associate  four 0-cells to each 2-cell of the extracted surface, i.e.~to each non-zero entry in every column of \T{B}.
+The $\mu$ function  can be computed by identifying, via  element position in the column, a triple of integer values $0\leq x\leq u, 0\leq y\leq v, 0\leq z\leq w$ for each vertex of the 2-cell.
+The mapping can be implemented using a dictionary, thet will store the inverse coordinate transformation used at the beginning, i.e.~the one from linear to Cartesian coords, in order of not duplicating the output vertices.
+
+\paragraph{Assembling the surface portions}
+
 \subsubsection{Block parallelism}\label{sec:bbbb}

### 2019-05-10 src/filter.tex
filter design draft finished

@@
 \[
-M = \ceil{u/n} \times \ceil{v/n} \times \ceil{w/n}
+M = \ceil{u/n} \times \ceil{v/n} \times \ceil{w/n} = \frac{uvw}{n^3}
 \] of blocks. 
@@
 \[
-\T{B} = [\partial_3(n,n,n)]\, \T{S}.
+\T{B} = [\partial_3(n)]\, \T{S}.
 \]
-
-Of course, the $\T{B}$ sparse matrix has the same column number $M$ of $\T{S}$, because each column contains the boundary representation of the corresponding $S\cap B_{i,j,k}$, and the number of rows of the operator, equal to the dimension of the linear space $C_2$.
+where the boundary matrix has dimension $d\,n\,(1+n)^{d-1} \times n^d$.
+Of course, the $\T{B}$ sparse matrix has the same column number $M$ of $\T{S}$, because each column contains the boundary representation of the corresponding $S\cap Box_{i,j,k}{i,j,k}$, and the number of rows of the operator, equal to the dimension of the linear space $C_2$.
 
@@
 The simplest solution is to associate  four 0-cells to each 2-cell of the extracted surface, i.e.~to each non-zero entry in every column of \T{B}.
-The $\mu$ function  can be computed by identifying, via  element position in the column, a triple of integer values $0\leq x\leq u, 0\leq y\leq v, 0\leq z\leq w$ for each vertex of the 2-cell.
-The mapping can be implemented using a dictionary, thet will store the inverse coordinate transformation used at the beginning, i.e.~the one from linear to Cartesian coords, in order of not duplicating the output vertices.
+The $\mu$ function  can be computed by identifying, via  element position in the column, a triple of integer values $0\leq x\leq u$, $0\leq y\leq v$, and $0\leq z\leq w$ for each vertex of the 2-cell.
+The mapping can be implemented using a dictionary, that will store the inverse coordinate transformation used at the beginning, i.e.~the one from linear to Cartesian coords, in order of not duplicating the output vertices.
 
-\paragraph{Assembling the surface portions}
+\paragraph{Surface assembling}
 
-\subsubsection{Block parallelism}\label{sec:bbbb}
+All boundary surface subsets $B_{i,j,k}(S) = \partial_3 S \cap \mbox{Box}_{i,j,k}$, provided by  columns of $\T{S}$, are embedded in the same coordinate space. In formal terms: 
+\[
+\texttt{Lar}(S) := (\texttt{Geom}(S), \texttt{Top}(S)) = (\texttt{V}, \texttt{CV}),
+\]
+where, with respect to the \emph{chain complex} $C_3\to C_2\to C_1\to C_0$ induced by the input image $Im$ and segment $S_{i,j,k}$, we get
+\begin{align}
+\texttt{Geom} &:= \mu(C_0(_{i,j,k})) = \texttt{V},
+\\
+\texttt{Top} &:= C_3(S) = \T{S} \mapsto \texttt{CV}.
+\end{align}
+and
+\begin{align}
+\texttt{Lar}(B_{i,j,k}) &:= (\texttt{Geom}, \texttt{Top}) = (\texttt{W}, \texttt{FW}),
+\\
+\texttt{Geom} &:= \mu(C_0(B_{i,j,k})) = \texttt{W} \subset \texttt{V},
+\\
+\texttt{Top} &:= C_2(B_{i,j,k}) = \T{B}_{i,j,k} \mapsto \texttt{FW} \subset \texttt{FV}.
+\end{align}
 
-\subsubsection{Reduced model}\label{sec:bbbb}
+A translation transformation applied to each vertex subset $\texttt{W}_{i,j,k}$ with translation  vector $\v{t} = [i,j,k]$ will therefore move it in the final space position, so finally giving
+\[
+\texttt{Lar}(B) = \oplus_{i,j,k}\texttt{Lar}(\partial_3 S_{i,j,k})) = \oplus_{i,j,k}(\texttt{W}, \texttt{FW}) .
+\]
+
+\subsubsection{Block-level parallelism}\label{sec:bbbb}
+
+In the computational pipeline introduced in this paper, several steps can be efficiently performed in parallel at image-block level, depending on the embarassingly data parallel nature of the problem.
+In particular, little effort is needed to separate the problem into a number of parallel tasks $S_{i,j,k}$, using multiarray slicing.
+The granularity of parallelism, depending on the block size $n$, is further enforced by the computation of a single boundary matrix $[\partial_d(n)]$, in turn depending on $n$, so that the initial communication cost of broadcasting the matrix to nodes can be carefully controlled, and finely tuned depending on the system architecture.
+The whole approach is appropriate  for SIMD hybrid architectures of CPUs and GPUs, since only the initial block setup of boundary matrix and image slices, as well the final collection of computed surface portions, require inter-process communication.
 

### 2019-05-10 src/implementation.tex
splitted filter.tex file into design.tex and implementation.tex

\subsection{Julia implementation}\label{sec:implementation}

\subsubsection{Code optimization}\label{sec:optimization}

\subsubsection{Performance analysis}\label{sec:analysis}

### 2019-05-25 src/background.tex
some background materials

@@
 
-\subsection{Linear Algebraic Representation}\label{sec:aaaa}
+\subsection{Representation Schemes}\label{sec:aaaa}
 
@@
 
+\subsection{Linear Algebraic Representation}\label{sec:aaaa}
+
+The \emph{Linear Algebraic Representation} (\textsc{lar}) aims to represent the \emph{chain complex} generated by a piecewise-linear geometric complex embedded either in 2D or in 3D.
+In few words, it gets a minimal characterization of geometry and topology of a cellular complex, i.e.~the embedding mapping $\mu : C_0 \to \E^d$ of 0-cells (vertices), as well a description of $(d-1)$-cells as subsets of vertices, and is able to return the whole chain complex
+\[ 
+C_\bullet = (C_p, \partial_p) := 
+C_3 \ 
+\substack{
+\delta_2 \\
+\longleftarrow \\[-1mm]
+\longrightarrow \\
+\partial_3 
+}
+\ C_2 \ 
+\substack{
+\delta_1 \\
+\longleftarrow \\[-1mm]
+\longrightarrow \\
+\partial_2 
+}
+\ C_1 \ 
+\substack{
+\delta_0 \\
+\longleftarrow \\[-1mm]
+\longrightarrow \\
+\partial_1 
+}
+\ C_0 .
+\] 
+and, in particular, any basis for linear chain spaces $C_p$, and any linear
+boundary/coboundary map \(\partial_p\) and
+\(\delta_p=\partial_{p+1}^\top\) between them.
+The domain of \textsc{lar} is the set of chain complexes generated by cell $d$-complexes ($2\leq d\leq 3$).
+The computer representations of \textsc{lar} are sparse binary matrices to represent both the operators and the chain bases.
+Note that in algebraic topology a $p$-chain is defined as a linear combination of $p$-cells with scalars from a field.
+When the scalar coefficients are from $\{-1, 0, +1\}$, a chain may represent \emph{any (oriented) subset of cells} from the cellular complex.
+
+We may therefore get the $(p-1)$-boundary $\partial_p \gamma_p$ of \emph{any} $p$-chain $\gamma_p$, by multiplication of the coordinate representation $[\partial_p]$ of the boundary operator times the coordinate representation $[\gamma_p]$ of the chain in terms of such scalars, i.e.~by a  matrix-vector product $ [\partial_p] [\gamma_p] $.
+
+It is easy to understand that the \textsc{lar} representation scheme is very expressive, i.e.~that it  has a large domain,  including collections of: line segments, quads, triangles, polygons, meshes;  pixels, voxels, volume images; B-reps, enumerative and decompositive representations of solids.
+In this paper we apply \textsc{lar} methods to computation of boundary representations of solid models from segmentation (labeling) of 3D medical images.
+
+\subsubsection{Construction of $\partial_d$ boundary matrix}
+
+Once fixed an ordering for all the cells (vertices, edges, pixels, and voxels), i.e.~for 0-, 1-, 2-, and 3-elements of a cell partitioning of a 3D image, i.e., once fixed the $p$-bases for linear spaces $C_p$ of $p$-chains $(0\leq p\leq d)$, we call $M_p = (m_{i,j})$ the \emph{characteristic matrix} of the $p$-basis, expressed as subsets of 0-cells, where  we have that $m_{i,j}=1$ iff the $j$-th $0$-cell belongs to the $i$-th $p$-cell, and $m_{i,j}=0$ otherwise.
+
+Note that, by computing the (sparse) matrix product $(M_{p-1} M_{p}^t) = (n_{i,j})$, with $n_{i,j} = \sum_{k} m_{i,k}m_{k,j}$, we get for each $n_{ij}$ the \emph{number of vertices} shared by $c_{p-1}^i$ and $c_{p}^j$.
+When this number equates the cardinality of $c_{p-1}^i$, this elementary chain is contained on the boundary of $c_{p}^j$.
+In a 3D image, with cubic 3-cells and squared 2-cells, everywhere we get $n_{i,j}=4$, we may state $c_{2}^i\subset\partial c_{3}^j$.
+In this case, by looking in each $j$ column of $M_{2} M_{3}^t = (n_{i,j})$, we have exactly \emph{six rows} where  $n_{i,j} = 4$.
+
+Finally, consider the linear graded boundary operator $\partial_p : C_p \to C_{p-1}$.
+As such, it contains by columns the representation of domain basis elements, expressed as linear combination of the basis elements of the range space.
+Therefore, the operator matrix $[\partial_d]$ is readily obtained by setting $[\partial_d](i,j)=1$ iff $n_{i,j}=4$ and $[\partial_d](i,j)=0$ otherwise.
+Of course, it will contain six non-zero elements for column.
+It may be worth to remember that every 3-cell (voxel) has exactly six 2-faces.
+
+It is possible to show that all the interesting relation of incidence/adjacency between cells of different dimensions can be both computed and efficiently queried by pairwise computing some products, with one of terms possibly transposed, of the two boundary and coboundary operator matrices $[\partial_p]$ and $[\delta_p] = [\partial_p^\top] = [\partial_p]^t$.
+We may also show that such matrices are \emph{very sparse}, with their sparseness growing rapidly with the dimensions (see Section~\ref{sec:block}).
+The pattern of non-zeros in matrix $[\partial_3]$ corresponding to a brick of shape $(4,4,4)$ is given in Figure~\ref{fig:boundary_matrix_4x4x4}.
+
+\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+   \centering
+   \includegraphics[width=0.75\linewidth]{figs/boundary_matrix_4x4x4.pdf} 
+   \caption{A binary image of the coboundary operator  $\delta_2 = \partial^\top : C_2 \to C_3$, built for small 3D images with shape $(4,4,4)$.
+Note that the number of rows equates the cardinality $4\times 4\times 4 = 64$ of the voxel set; the number of columns is $d\,n\,(1+n)^{d-1} = 3\times 4\times 25 = 300$.
+Of course, the number of non-zeros per row (cardinality of single voxel facets) is six.}
+   \label{fig:boundary_matrix_4x4x4}
+\end{figure}
+
 \subsection{Multiindices from Cartesian indices}\label{sec:aaaa}
 
+In order to utilize the topological algebra shortly recalled in this paper, we need to explicitly sort the cells of the various dimensions into linearly ordered sequences, possibly according to the linear order their information is linearly accommodated in computer storage.
+
 \subsection{Taubin Smoothing}\label{sec:aaaa}

### 2019-05-25 src/design.tex
splitted filter.tex file into design.tex and implementation.tex
some background materials

--- 2019-05-10 src/filter.tex
@@
-\section{Filter design and implementation}\label{sec:filter}
+\section{Block-parametric design}\label{sec:filter}
 
-\subsection{Block-parametric design}\label{sec:block}
-
-\subsubsection{Block decomposition}\label{sec:bbbb}
+\subsection{Block decomposition}\label{sec:bbbb}
 
 We assume that medical devices produce 3D images with lateral dimensions that are integer multiples of some powers of two, like 128, 256, 512, etc.
-Any cuboidal portion of image is completely determined by the Cartesian indices of its voxel of lowest and highest indices, and denoted as $Image([\ell_x, \ell_y, \ell_z],[h_x, h_y, h_z])$.
+Any cuboidal portion of image is completely determined by the Cartesian indices of its voxel of lowest and highest indices, and extracted by multidimensional array \emph{slicing} as $Image([\ell_x$:$h_x, \ell_y$:$h_y, \ell_z$:$h_z])$.
 
@@
 \[
-B(i,j,k,n) := Image([in,jn,kn],[i(n+1),j(n+1),k(n+1)]) 
+B(i,j,k,n) := Image([in\mbox{:}in+n, jn\mbox{:}jn+n, kn\mbox{:}kn+n]) 
 \]
@@
    \centering
-   \includegraphics[width=2in]{figs/blocks} 
-   \caption{A possible block partitioning of a radiologic image. The evidenced 2D block is $B([2,1,64]) = Image([128,64],[172,128)]$}
+   \includegraphics[width=0.5\linewidth]{figs/blocks} 
+   \caption{A possible block partitioning of a radiologic image. The evidenced 2D block, of size $n^d=64^2$, is sliced as $B([2,1,64]) = Image([128:172],[64:128)]$}
    \label{fig:blocks}
@@
 
-Figure~\ref{fig:blocks} shows the block decomposition in a 2D case, with positive integer $u,v$ lateral sizes of image. Note that block sides do not necessarily correspond to image edges.
+Figure~\ref{fig:blocks} shows the block decomposition in a 2D image, with positive integer $u,v$ lateral sizes of image. Note that block sides do not necessarily correspond to image edges.
 
-\subsubsection{Block operator }\label{sec:bbbb}
+\subsection{Block operator }\label{sec:block}
 
@@
 
-\subsubsection{Block boundary mapping}\label{sec:bbbb}
+\subsection{Block boundary mapping}\label{sec:bbbb}
 
@@
 
-\subsubsection{Block-level parallelism}\label{sec:bbbb}
+\subsection{Block-level parallelism}\label{sec:bbbb}
 
@@
 The whole approach is appropriate  for SIMD hybrid architectures of CPUs and GPUs, since only the initial block setup of boundary matrix and image slices, as well the final collection of computed surface portions, require inter-process communication.
-
-\subsection{Julia implementation}\label{sec:implementation}
-
-\subsection{Code optimization}\label{sec:optimization}
-
-\subsection{Performance analysis}\label{sec:analysis}

### 2019-06-11 src/background.tex
posted contents for implementation section

@@
 In particular, a boundary representation provides a cellular decomposition of the object's boundary into cells of dimension zero (vertices), one (edges), and two (faces).
-Medical imaging is a kind of enumerative representation of organs and tissues of interest, as subsets of 3D volume elements (voxels) from the 3D image.
+Medical imaging can be classified as \emph{enumerative representation} of cellular decompositions of organs and tissues of interest, in particular, as subsets \emph{of 3D volume elements} (voxels) from the 3D image.
 

### 2019-06-11 src/implementation.tex
posted contents for implementation section

@@
 \subsection{Julia implementation}\label{sec:implementation}
+
+The computer code is implemented in Julia language~\cite{} according to the workflow described below, whose stages are parallelized and/or optimized in various ways.
+
+\paragraph{Workflow setup} The functions in this preliminary step include:
+\begin{enumerate}
+
+\item input of 3D medical image $\mathcal{I}$ dimensions $\ell_1, \ell_2, \ell_3$, such that: $\mathcal{I} = [\ell_1]\times[\ell_2]\times[\ell_3]$, where $[\ell_k] = (1,2,\ldots,\ell_k)$;
+
+\item analysis of resources available in the computational environment, including operating system, type and number of compute nodes (processors, cores, GPUs), number of cores per node, RAM and caches amounts;
+
+\item depending on the above, best decision for $\emph{size}$ of 3D image blocks $B$ (bricks); defaults to: $\emph{size}=64$; hence the number of bricks will be $n=\ceil{\ell_1/size}\times\ceil{\ell_2/size}\times\ceil{\ell_3/size}$.
+Hence default value: $n=256$, for images $\mathcal{I} = 512\times 512\times 256$;
+
+\item computation of a Julia's sparse boundary matrix $[\partial_B]$, returning a value of type \texttt{SparseMatrixCSC\{Int8\}\{Int64\}}, where \texttt{Int8} and \texttt{Int64} are the types for values and indices, respectively, stored by Compressed Sparse Column (CSC) format; the average $[\partial_B]$ (for $\emph{size}$ = 64) is about
+
+\item creation of a either a local or distributed channel to implement a producer/consumer model of parallel/distributed computation, depending on available resources;
+
+\item distribution of matrix $[\partial_B]$, of size $45$ MB, to all available nodes/cores (workers) using the macro \texttt{@eveywhere}.
+
+\end{enumerate}
+
+The number of non-zeros within the sparse matrix $[\partial_{64^3}]$ is $\mathtt{nnz} = 4 792 266$, for a memory size of $9\times \mathtt{nnz}+8*262144 \simeq 45$ MB.
+The memory size of the sparse matrix is computed as $8+1$ bytes for non-zero element (i.e.
+6 per row), plus 8 bytes per column.
+
+\paragraph{Job enqueuing} Communication and data synchronization may be managed through \emph{Channels}, which are the FIFO conduits that provide producer/consumer communication.
+Overall execution time can be improved if other tasks can be run while a task is being executed, or while waiting for an external service/function to complete.
+The single work items follows:
+\begin{enumerate}
+
+\item extraction of the block views from image arrays depending on 3 indices;
+
+\item transformation of each job from global to local coordinates;
+
+\item further transform of each forground voxel from Cartesian to linear coordinates, using the suitable Julia's library functions.
+
+\item enqueing of the job (a sequence of integer positions for the non-zeros image elements aligned in a memory buffer).
+\end{enumerate}
+
+\paragraph{3-Chain encoding} 
+Each block $B$ of the 3D image must by converted into the coordinate representation of a vector of the linear space $C_3$ of 3-chains.
+In coordinates local to $B$, once fixed an ordering from Cartesian to linear coords, this vector is represented by a binary array of length $64^3=262144$,  with a non-zero value (i.e.~$1$) for each foreground voxel in the interesting image portion in $B$.
+Therefore, the coded portion of the segment $\mathcal{S}$ inside $B$, denoted as $\mathcal{S}(B)$, results with a space occupancy of about $262$ KB if encoded as a full array (i.e.~including the zero values), When encoded as a sparse vector, its space occupancy will correspondingly decrease.
+
+\begin{enumerate}
+
+\item each job encoding produces either a full or sparse binary vector. With full arrays depending by one index, we get 262 KB per job;
+
+\item special format for sparse CSC (Compressed Sparse Column) vectors can be used, since the \emph{value} data for non-zeros do not need storage.
+Hence only a single 1-array of \texttt{Int64} row positions (with total length equal to the number of non-zeros in the block, with $8\times\mathtt{nnz}$ KB storage) is needed;
+
+\item prepare subsequnces of such data vectors (non-zero row indices in linear indices), in order to feed efficiently the available processor threads.
+In case of presence of one/more GPUs, a smaller size of the block---and hence of the boundary matrix and the encoded 3-chain vectors---and then much higher vector numbers, are preferable for speed.
+
+\end{enumerate}
+
+\paragraph{SpMM Multiplication} 
+According to the current literature~\cite{} it is more convenient to execute SpMV (sparse matrix-vector) multiplications than SpMSpV (sparse matrix-sparse vector) multiplications.
+Since we have 256 such jobs (one multiplication per block) to perform in the standard setting of the algorithm\footnote{
+Size of block $64^3$; size of image $512^2\times 256$.
+}, or even more, in case of either smaller blocks or image greater than the standard ones, this stage must be evidently parallelized and carefully tuned, possibly by using the available GPUs.
+\begin{enumerate}
+
+\item Various multiplication algorithms are being experimented, using several packages for sparse algebra and/or customized implementations;
+
+\item the total speed of this stage will strongly depend on both the hardware available, on the granularity of blocks, and on the choice between dense/sparse storage of encoded 3-chains;
+
+\item anyway, the compute elements or threads will be feed without solution of continuity in a \emph{dataflow} process.
+This parallel operation is, according to our preliminary experiments, the critical one of the whole workflow, in the sense that any $\Delta T$ (either positive or negative) in this stage will contribute to the total time $T$.
+
+\end{enumerate}
+
+\paragraph{2-Chain decoding} 
+Each multiplication of $[\partial_B] : C_3 \to C_2$, times a 3-chain $\nu\in C_3$, produces a 2-chain  $\sigma\in C_2$, i.e.~the coordinate representation of the boundary vector $\sigma\in C_2$.
+The inverse of the coding algorithm is executed in the present stage.
+This process can also be partially superimposed in time with the previous one, depending on the size of the memory buffer used to feed the CPU cores or the GPUs.
+The elementary steps follow:
+
+\begin{enumerate}
+
+\item conversion from position of ones (or non-zeros) in the 2-chain to row linear indices;
+
+\item conversion from linear indices to Cartesian indices in coordinates local to the $B$ block, using the appropriate library functions;
+
+\item conversion from each Cartesian index value to a suitably oriented (i.e.~with proper attitude) geometry quadrilateral (or pair of triangles) in local coordinates.
+
+\end{enumerate}
+
+A Julia's vectorized pipeline seems the more appropriate implementation model for the job of each worker.
+
+\paragraph{Assembling and artifact filtering} 
+The results of the previous stages can be described a a collection of sets of geometric quads, each one encoded as an array of quadruples of integer indices, pointing to the linear array of grid vertices associated to the standard image block $B$.
+In other words, \emph{all quads} of \textbf{each job} are now given in the \textbf{same} \emph{local coordinates}.
+Other than to put each partial surface $\mathcal{S}(B) = (\texttt{V}_B, \sum_{\sigma\in B} \texttt{FV}_\sigma)$ in the global coordinate system of the image, the present stage must eliminate the redundant boundary features possibly generated at the edges of the partial surface $\mathcal{S}(B)$ within each block $B$ such that $B \cap \mathcal{I} \not= \emptyset$.
+\begin{enumerate}
+
+\item translate each array \texttt{FV}, of type \texttt{Lar.Cells}, by summing to each linear vertex index the linearized offset of the Cartesian coordinates $(n,m,p)(B)$ of the vertex with lowest coordinates of its block $B$.
+
+\item remove both instances of double quads generated by \texttt{Lar} software at the block boundaries (see Figure~\ref{fig:blockboundary}).
+They are artifacts generated by the decomposition of the whole image into a number of blocks of tractable size.
+
+\item 
+a smart strategy of removal of such artifacts can be used, which does not require any sorting nor searching on the assembled array of quads.
+It will consist of arranging each block with all three dimensions increased by one, so that each 2-adjacent pair of blocks will covering each other for a full side of blocks of depth one.
+The details of this artifact filtering will be elucidated in Section~\ref{sec:covering}.
+
+\end{enumerate}
+
+\paragraph{Smoothing} 
+The final smoothing of the generated surfaces cannot be performed block-wise, since this would introduce smoothing artifacts at the block boundaries.
+Anyway, the Taubin smoothing~\cite{} can be performed in parallel, since for each vertex in the final surface (except eventually the ones on the image $\mathcal{I}$ boundaries) it essentially consist in computing a new position as a proper average of its neighborhood vertices, i.e.~by applying a discrete Laplacian operator.
+Some appropriate set workers may so be assigne the task of generating iteratively a new position for the vertices they take cure of.
+In particular, we have:
+\begin{enumerate}
+
+\item Job enqueuing, by writing sets of integers (global linear indices of vertices) in array buffers of appropriate type \texttt{Channel};
+
+\item iterated vectorized computation of proper averages of closed vertices;
+
+\item job dequeuing, by recovering finished tasks from a channel and assembling the results into the embedding function $\mathtt{V}: C_0 \to \E^3$ providing an array of type \texttt{Lar.Points} of \texttt{Float64 $\times$ 3}, with vertex coordinates by column.
+\end{enumerate}
 

### 2019-06-11 src/implementation.tex
posted contents for implementation section

--- 2019-05-10 src/implementation.tex
@@
 \subsection{Julia implementation}\label{sec:implementation}
+
+The computer code is implemented in Julia language~\cite{} according to the workflow described below, whose stages are parallelized and/or optimized in various ways.
+
+\paragraph{Workflow setup} The functions in this preliminary step include:
+\begin{enumerate}
+
+\item input of 3D medical image $\mathcal{I}$ dimensions $\ell_1, \ell_2, \ell_3$, such that: $\mathcal{I} = [\ell_1]\times[\ell_2]\times[\ell_3]$, where $[\ell_k] = (1,2,\ldots,\ell_k)$;
+
+\item analysis of resources available in the computational environment, including operating system, type and number of compute nodes (processors, cores, GPUs), number of cores per node, RAM and caches amounts;
+
+\item depending on the above, best decision for $\emph{size}$ of 3D image blocks $B$ (bricks); defaults to: $\emph{size}=64$; hence the number of bricks will be $n=\ceil{\ell_1/size}\times\ceil{\ell_2/size}\times\ceil{\ell_3/size}$.
+Hence default value: $n=256$, for images $\mathcal{I} = 512\times 512\times 256$;
+
+\item computation of a Julia's sparse boundary matrix $[\partial_B]$, returning a value of type \texttt{SparseMatrixCSC\{Int8\}\{Int64\}}, where \texttt{Int8} and \texttt{Int64} are the types for values and indices, respectively, stored by Compressed Sparse Column (CSC) format; the average $[\partial_B]$ (for $\emph{size}$ = 64) is about
+
+\item creation of a either a local or distributed channel to implement a producer/consumer model of parallel/distributed computation, depending on available resources;
+
+\item distribution of matrix $[\partial_B]$, of size $45$ MB, to all available nodes/cores (workers) using the macro \texttt{@eveywhere}.
+
+\end{enumerate}
+
+With $size=64$, the number of non-zeros within the sparse matrix $[\partial_{64^3}]$ is $\mathtt{nnz} = 4 792 266$, for a memory size of $9\times \mathtt{nnz}+8*262144 \simeq 45$ MB.
+The memory size of the sparse matrix is computed by considering $8+1$ bytes for non-zero element (which are exactly 6 per row), plus 8 bytes per each index of column start.
+
+\paragraph{Job enqueuing} Communication and data synchronization may be managed through \emph{Channels}, which are the FIFO conduits that may provide producer/consumer communication.
+Overall execution time can be improved if other tasks can be run while a task is being executed, or while waiting for an external service/function to complete.
+The single work items of this stage follow:
+\begin{enumerate}
+
+\item extraction of the block views, depending on 3 Cartesian indices, from image arrays;
+
+\item transform  each job \emph{from global} $[\ell_1]\times[\ell_2]\times[\ell_3]$ to \emph{local coordinates} $[n]\times[n]\times[n]$;
+
+\item further transform of each \emph{foreground voxel} $\nu\in\mathcal{S}\subseteq\mathcal{I}$ from Cartesian to linear coordinates, using the suitable Julia's library functions.
+
+\item enqueing the job (as a sequence of integer positions for the non-zeros image elements aligned in a memory buffer of proper \texttt{Channel} type).
+\end{enumerate}
+
+\paragraph{3-Chain encoding} 
+The interesting part of the \emph{Image} $\mathcal{I}$ is called \emph{Segment} $\mathcal{S}$.
+The goal of the whole \emph{workflow} is to extract a \emph{boundary model} of $\mathcal{S}$ from $\mathcal{I}$.
+The portion of $\mathcal{S}$ inside $B$, will be denoted as $\mathcal{S}(B)$.
+Each block $B$ of the 3D image must by converted into the \emph{coordinate representation} of a vector $\nu\in C_3$ in the linear space  of 3-chains.
+
+In coordinates local to $B$, once fixed an ordering from Cartesian to linear coords, this vector is represented by a \emph{binary array} of length $size^3$.
+With $size=64$, we have $64^3=262144$,  with a non-zero value (i.e.~$1$) for each foreground voxel in $\mathcal{S}(B)$.
+Therefore, the coded segment portion $\mathcal{S}(B)$, results with a space occupancy of about $262$ KB if encoded as a full array (i.e.~including the zero values).
+Whether encoded as a sparse vector, its space occupancy will correspondingly decrease.
+
+\begin{enumerate}
+
+\item each encoding task produces either a full or sparse binary vector. With full or sparse arrays depending by one index, we get either 262 KB or less per job, correspondingly;
+
+\item special format for sparse CSC (Compressed Sparse Column) vectors can be used, since the \emph{value} data for non-zeros does not need storage.
+Hence only a single 1-array of \texttt{Int64} row positions (with total length equal to the number of non-zeros in the block, with $8\times\mathtt{nnz}$ kB storage) is needed;
+
+\item prepare subsequences of such data vectors (non-zero linear row indices), in order to feed efficiently the available processor threads.
+In case of presence of one/more GPUs, a smaller size of the block---and hence of the boundary matrix and the encoded 3-chain vectors---and then much higher vector numbers, are preferable for speed.
+
+\end{enumerate}
+
+\paragraph{SpMM Multiplication} 
+According to the current literature~\cite{} it is more convenient to execute SpMV (sparse matrix-vector) multiplications than SpMSpV (sparse matrix-sparse vector) multiplications.
+Since we have 256 such jobs (one multiplication per block) to perform in the standard setting of the algorithm\footnote{
+Size of block $64^3$; size of image $512^2\times 256$.
+}, or more in case of either smaller blocks or image greater than the standard one, this stage must be evidently parallelized and carefully tuned, possibly by using the GPU, if available.
+\begin{enumerate}
+
+\item Various multiplication algorithms are being experimented, using several packages for sparse linear algebra and/or custom implementations;
+
+\item the total speed of this stage will strongly depend on the hardware available, on the granularity of blocks, and on the choice between dense/sparse storage of encoded 3-chains;
+
+\item anyway, the compute elements or threads will be feed without solution of continuity in a \emph{dataflow} process.
+This parallel operation is, according to our preliminary experiments, the critical one of the whole workflow, in the sense that any $\Delta T$ (either positive or negative) in this stage will contribute to the total time $T$.
+
+\end{enumerate}
+
+\paragraph{2-Chain decoding} 
+Each multiplication of $[\partial_B] : C_3 \to C_2$, times a 3-chain $\nu\in C_3$, produces a 2-chain  $\sigma\in C_2$, i.e.~the \emph{coordinate representation} of the \emph{boundary vector} $\sigma\in C_2$.
+The inverse of the coding algorithm is executed in the present stage.
+This process can also be partially superimposed in time with the previous ones, depending on the size of the memory buffers used to feed the CPU cores or the GPUs and get their results.
+Some elementary steps follow:
+
+\begin{enumerate}
+
+\item conversion from position of ones (or non-zeros) in the 2-chain to linear indices of rows;
+
+\item conversion from linear indices to Cartesian indices in coordinates local to the $B$ block, using the appropriate library functions;
+
+\item conversion from each Cartesian index value to a suitably oriented (i.e.~with proper attitude) geometry quadrilateral (or pair of triangles) in local coordinates.
+
+\end{enumerate}
+
+A Julia's vectorized pipeline dataflow seems the more appropriate implementation model for the job of each worker.
+
+\paragraph{Assembling and artifact filtering} 
+The results of the previous stages can be described as a \emph{collection of sets} of \emph{geometric quadrilaterals (quads)}, each one encoded as an array of quadruples of integer indices, pointing to the linear array of grid vertices associated to the standard image block $B$.
+In other words, \emph{all quads} of \textbf{each job} are now given in the \textbf{same} \emph{local coordinates}.
+Besides to put each partial surface $\mathcal{S}(B) = (\texttt{V}_B, \texttt{FV}_\sigma)$ in the global coordinate system of the image, the present stage must eliminate the redundant boundary features possibly generated at the edges of the partial surface $\mathcal{S}(B)$ within each block $B$ such that $B \cap \mathcal{I} \not= \emptyset$:
+\begin{enumerate}
+
+\item translate each array $\mathtt{FV}_\sigma$, of type \texttt{Lar.Cells}, by summing to each vertex index the linearized offset of the Cartesian coordinates $(n,m,p)(B)$ of the $B$'s \emph{reference vertex}, i.e.~the one with (all) lowest \emph{Cartesian coordinates} within the $B$ block.
+
+\item remove both instances of \emph{double quads} generated by \texttt{Lar} software at the block boundaries (see Figure~\ref{fig:blockboundary}).
+They are artifacts generated by the decomposition of the whole image into a number of blocks of tractable size.
+
+\item 
+a smart strategy of removal of such artifacts may be used, which does not require any sorting nor searching on the assembled array of quads.
+It will consist in arranging each block with all three dimensions decreased-increased by one, so that each 2-adjacent pair of blocks will be covering each other for a full side extent of blocks of depth one.
+The details of this \emph{artifact filtering} will be elucidated in Section~\ref{sec:covering}.
+
+\end{enumerate}
+
+\paragraph{Smoothing} 
+The final smoothing of the generated surfaces cannot be performed block-wise, since this would introduce smoothing artifacts at the block boundaries.
+Anyway, the Taubin smoothing~\cite{} can be performed in parallel, since for each vertex in the final surface (except eventually the ones on the image $\mathcal{I}$ boundaries) it essentially consist in computing a new position as a proper average of its neighborhood vertices, i.e.~by applying a discrete Laplacian operator.
+Some appropriate set workers may so be assigne the task of generating iteratively a new position for the vertices they take cure of.
+In particular, we have:
+\begin{enumerate}
+
+\item Job enqueuing, by writing sets of integers (global linear indices of vertices) in array buffers of appropriate type \texttt{Channel};
+
+\item iterated vectorized computation of proper averages of closed vertices;
+
+\item job dequeuing, by recovering finished tasks from a channel and assembling the results into the embedding function $\mathtt{V}: C_0 \to \E^3$ providing an array of type \texttt{Lar.Points} of \texttt{Float64 $\times$ 3}, with vertex coordinates by column.
+\end{enumerate}
 

### 2019-06-12 src/implementation.tex
crumbs

@@
 
-\item extraction of the block views, depending on 3 Cartesian indices, from image arrays;
+\item extraction, from image arrays of the block views, depending on 3 Cartesian indices;
 
-\item transform  each job \emph{from global} $[\ell_1]\times[\ell_2]\times[\ell_3]$ to \emph{local coordinates} $[n]\times[n]\times[n]$;
+\item transform  each block \emph{from global} $[\ell_1]\times[\ell_2]\times[\ell_3]$ to \emph{local coordinates} $[n]\times[n]\times[n]$;
 

### 2019-06-12 src/implementation.tex
posted Appendix

@@
 
-With $size=64$, the number of non-zeros within the sparse matrix $[\partial_{64^3}]$ is $\mathtt{nnz} = 4 792 266$, for a memory size of $9\times \mathtt{nnz}+8*262144 \simeq 45$ MB.
+With $size=64$, the number of non-zeros within the sparse matrix $[\partial_{64^3}]$ is $\mathtt{nnz} = 4 792 266$, for a memory size of $9\times \mathtt{nnz}+8\times 262144 \simeq 45$ MB.
 The memory size of the sparse matrix is computed by considering $8+1$ bytes for non-zero element (which are exactly 6 per row), plus 8 bytes per each index of column start.
@@
 
-\subsubsection{Code optimization}\label{sec:optimization}
-
 \subsubsection{Performance analysis}\label{sec:analysis}

### 2019-06-12 src/main.tex
working on algorithm design
filter design draft finished
splitted filter.tex file into design.tex and implementation.tex
some background materials
posted Appendix

@@
 \usepackage{amssymb}
+\usepackage{amsmath}
+\usepackage{hyperref}
 
@@
 \usepackage{lmodern}
+
+\input{macros}
 
@@
 \include{background}
-\include{filter}
+\include{design}
+\include{implementation}
 \include{examples}
@@
 
+\appendix
+\section{Appendix}
+\subsection{Symbol list}
+
+\begin{description}
+\item[$\mathcal{I}$]  three-dimensional medical image
+\\[-8mm]
+\item[$\ell_1, \ell_2, \ell_3$]  dimensions of image
+\\[-8mm]
+\item[$\mathcal{S}$]  segment: subset of voxels
+\\[-8mm]
+\item[$\mathcal{B}$] 3D image block 
+\\[-8mm]
+\item[$size$] lateral dimension of cubic block $\mathcal{B}$ 
+\\[-8mm]
+\item[$n$]	number of blocks $\mathcal{B}$ (jobs) in $\mathcal{I}$
+\\[-8mm]
+\item[[$\partial_\mathcal{B}$] ]  boundary matrix for block $\mathcal{B}$
+\\[-8mm]
+\item[$C_p$] linear (vector) space of $p$-chains
+\\[-8mm]
+\item[$\nu\in C_p$] $p$-chain
+\\[-8mm]
+\item[[$\nu$]] coordinate representation (binary vector) of  $\nu$
+\\[-8mm]
+\item[$\E^3$] Euclidean 3-space
+\end{description}
+
+\subsection{Definitions}
+\begin{description}
+\item[CSC]  Compressed Sparse Column format for sparse matrices
+\\[-8mm]
+\item[Global coordinates]  Integer linear coordinates of $\mathcal{I}$
+\\[-8mm]
+\item[Local coordinates] Integer linear coordinates of $\mathcal{B}$
+\\[-8mm]
+\item[Cartesian coordinates] Integer triples $(i,j,k)$ one-to-one with voxels 
+\\[-8mm]
+\item[Voxels] individual elements in 3D image (3-cells)
+\\[-8mm]
+\item[$p$-chain] formal linear combination of $p$-cells with coefficients in $\{0,1\}$
+\\[-8mm]
+\item[Coord. repr.]  binary vector (for $p$-chains) or binary matrix (for chain operators)
+\\[-8mm]
+\item[Quad]	geometric quadrilateral; convex polygon with four vertices 
+\\[-8mm]
+\item[Foreground voxel] individual element of a segment $\mathcal{S}$
+\\[-8mm]
+\item[Segment]	subset of voxels resulting from image segmentation
+\end{description}
+
 \end{document}

### 2019-06-23 src/examples.tex
tables added

\section{Examples}\label{sec:examples}

The 
For an experiments the we used the dataset Ircad1b.

\begin{table}
\input{input/dataset_ircad}
\end{table}
\begin{table}
\input{input/dataset_ircad_desc}
\end{table}

### 2019-06-30 src/background.tex
my last update

@@
 A \emph{representation scheme} for solid modeling is a mapping between a space of mathematical models and a space of symbolic representations, often generated by a formal grammar.
-Solid pointsets (i.e., `$r$-sets') are defined~\cite{requicha:80} as compact (bounded and closed) regular and semianalytic subsets of the $d$-space.
-A large number of representation schemes were defined in the past forty years, including the two main classes of (a) \emph{boundary representations} (`$B$-reps'), where the solid model is represented through a representation of its boundary elements, i.e.~faces, edges and vertices, and (b) \emph{decompositive/enumerative representations}, that are a decomposition of either the object or the embedding space, respectively, into a well-defined cellular complex.
-In particular, a boundary representation provides a cellular decomposition of the object's boundary into cells of dimension zero (vertices), one (edges), and two (faces).
+Solid pointsets (i.e., `$r$-sets') are defined~\cite{Requicha:1980:RRS:356827.356833} as compact (bounded and closed) regular and semianalytic subsets of the $d$-space.
+A large number of representation schemes were defined in the past forty years, including the two main classes of (a) \emph{boundary representations} (`$B$-reps'), where the solid model is represented through a representation of its boundary elements, i.e.~faces, edges and vertices, and (b) \emph{decompositive/enumerative representations}, that are a decomposition of either the object or the embedding space, respectively, into a well-defined \emph{cellular complex}.
+In particular, a boundary representation provides a cellular decomposition of the object's boundary into \emph{cells} of dimension zero (vertices), one (edges), and two (faces).
 Medical imaging can be classified as \emph{enumerative representation} of cellular decompositions of organs and tissues of interest, in particular, as subsets \emph{of 3D volume elements} (voxels) from the 3D image.
@@
 
-The \emph{Linear Algebraic Representation} (\textsc{lar}) aims to represent the \emph{chain complex} generated by a piecewise-linear geometric complex embedded either in 2D or in 3D.
+The \emph{Linear Algebraic Representation} (\textsc{lar})~\cite{Dicarlo:2014:TNL:2543138.2543294} aims to represent the \emph{chain complex}~\cite{TSAS} generated by a piecewise-linear geometric complex embedded either in 2D or in 3D.
 In few words, it gets a minimal characterization of geometry and topology of a cellular complex, i.e.~the embedding mapping $\mu : C_0 \to \E^d$ of 0-cells (vertices), as well a description of $(d-1)$-cells as subsets of vertices, and is able to return the whole chain complex

### 2019-06-30 src/design.tex
posted contents for implementation section
my last update

@@
 
-For the sake of simplicity, we assume a common size on the three image axes, and the corresponding image portion $B$, called \emph{block}, as a function of its element of the  lowest  block  coordinates $i,j,k\in\N$ and of block dimension $n$:
+For the sake of simplicity, we assume a common size on the three image axes, and the corresponding image portion $\mathcal{B}$, called \emph{block}, as a function of its element of the  lowest  block  coordinates $i,j,k\in\N$ and of block dimension $n$:
 \[
-B(i,j,k,n) := Image([in\mbox{:}in+n, jn\mbox{:}jn+n, kn\mbox{:}kn+n]) 
+\mathcal{B}(i,j,k,n) := Image([in\mbox{:}in+n, jn\mbox{:}jn+n, kn\mbox{:}kn+n]) 
 \]
@@
    \includegraphics[width=0.5\linewidth]{figs/blocks} 
-   \caption{A possible block partitioning of a radiologic image. The evidenced 2D block, of size $n^d=64^2$, is sliced as $B([2,1,64]) = Image([128:172],[64:128)]$}
+   \caption{A possible block partitioning of a radiologic image. The evidenced 2D block, of size $n^d=64^2$, is sliced as $\mathcal{B}([2,1,64]) = Image([128:172],[64:128)]$}
    \label{fig:blocks}
@@
 We are going to treat each image block independently from each other.
-Hence we map each image subspace $B(i,j,n)$ to the linear \emph{chain} space $C_2$ of dimension $n\times n$, using coordinate vectors $c_{h,k}\in \B^{n\times n} := \{0,1\}^{n\times n}$, where the basis element $c = c_{h,k} \in C_2$ is mapped via Cartesian-to-linear map to the Boolean vector
+Hence we map each image subspace $\mathcal{B}(i,j,n)$ to the linear \emph{chain} space $C_2$ of dimension $n\times n$, using coordinate vectors $c_{h,k}\in \\mathcal{B}^{n\times n} := \{0,1\}^{n\times n}$, where the basis element $c = c_{h,k} \in C_2$ is mapped via Cartesian-to-linear map to the binary vector
 \[
-Image(h,k) \mapsto c_{h,k} := [0 \cdots 0\ 1\ 0 \cdots 0] \in \B^{n\times n}
+Image(h,k) \mapsto c_{h,k} := [0 \cdots 0\ 1\ 0 \cdots 0] \in \\mathcal{B}^{n\times n}
 \]
@@
 
-Therefore, each pixel (or voxel) in a block image will be seen as a basis Boolean vector in $C_2$, and each subset of image elements, as the corresponding Boolean vector in $C_2$, with many ones as the cardinality of the subset.
+Therefore, each pixel (or voxel) in a block image will be seen as a basis binary vector in $C_2$, and each subset of image elements, as the corresponding binary vector in $C_2$, with many ones as the cardinality of the subset.
 
@@
 
+\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+   \centering
+   \includegraphics[width=0.4\linewidth]{figs/grid.pdf} 
+   \caption{example caption}
+   \label{fig:example}
+\end{figure}
+
 \subsection{Block boundary mapping}\label{sec:bbbb}
@@
 \[
-M = \ceil{u/n} \times \ceil{v/n} \times \ceil{w/n} = \frac{uvw}{n^3}
+M = \ceil{u/n} \times \ceil{v/n} \times \ceil{w/n} \simeq \frac{uvw}{n^3}
 \] of blocks. 
-Then we consider each image portion $c_{i,j,k} = S\cap B(i,j,k,n)$ and compute its (binary) coordinate representation  $[c]_{i,j,k}\in C_3(n,n,n)$.
+Then we consider each image portion $c_{i,j,k} = S\cap \mathcal{B}(i,j,k,n)$ and compute its (binary) coordinate representation  $[c]_{i,j,k}\in C_3(n,n,n)$.
 This one is a sparse binary vector of length $n^3$.
 Then assemble the $M$ representations $c$ of segment portions into a sparse binary matrix $\T{S}$, of dimension $n^d \times M$.
-Finally compute a matrix $\T{B}$ of boundary portions of $S$, represented by columns as chain coordinate vectors in $C_2$:
+Finally compute a matrix $\T{\mathcal{B}}$ of boundary portions of $S$, represented by columns as chain coordinate vectors in $C_2$:
 \[
-\T{B} = [\partial_3(n)]\, \T{S}.
+\T{\mathcal{B}} = [\partial_3(n)]\, \T{S}.
 \]
 where the boundary matrix has dimension $d\,n\,(1+n)^{d-1} \times n^d$.
-Of course, the $\T{B}$ sparse matrix has the same column number $M$ of $\T{S}$, because each column contains the boundary representation of the corresponding $S\cap Box_{i,j,k}{i,j,k}$, and the number of rows of the operator, equal to the dimension of the linear space $C_2$.
+Of course, the $\T{\mathcal{B}}$ sparse matrix has the same column number $M$ of $\T{S}$, because each column contains the boundary representation of the corresponding $S\cap Box_{i,j,k}{i,j,k}$, and the number of rows of the operator, equal to the dimension of the linear space $C_2$.
 
@@
 In particular, we need to compute the \emph{embedding function} $\mu : C_0 \to \E^3$, where $C_0$ is the space of 0-chains, one-to-one corresponding to the vertices of the extracted surface.
-The simplest solution is to associate  four 0-cells to each 2-cell of the extracted surface, i.e.~to each non-zero entry in every column of \T{B}.
+The simplest solution is to associate  four 0-cells to each 2-cell of the extracted surface, i.e.~to each non-zero entry in every column of $\T{\mathcal{B}}$.
 The $\mu$ function  can be computed by identifying, via  element position in the column, a triple of integer values $0\leq x\leq u$, $0\leq y\leq v$, and $0\leq z\leq w$ for each vertex of the 2-cell.
@@
 
-All boundary surface subsets $B_{i,j,k}(S) = \partial_3 S \cap \mbox{Box}_{i,j,k}$, provided by  columns of $\T{S}$, are embedded in the same coordinate space. In formal terms: 
+All boundary surface subsets $\mathcal{B}_{i,j,k}(S) = \partial_3 S \cap \mbox{Box}_{i,j,k}$, provided by  columns of $\T{S}$, are embedded in the same coordinate space. In formal terms: 
 \[
@@
 \\
-\texttt{Top} &:= C_2(B_{i,j,k}) = \T{B}_{i,j,k} \mapsto \texttt{FW} \subset \texttt{FV}.
+\texttt{Top} &:= C_2(B_{i,j,k}) = \T{\mathcal{B}}_{i,j,k} \mapsto \texttt{FW} \subset \texttt{FV}.
 \end{align}
@@
 \[
-\texttt{Lar}(B) = \oplus_{i,j,k}\texttt{Lar}(\partial_3 S_{i,j,k})) = \oplus_{i,j,k}(\texttt{W}, \texttt{FW}) .
+\texttt{Lar}(\mathcal{B}) = \oplus_{i,j,k}\texttt{Lar}(\partial_3 S_{i,j,k})) = \oplus_{i,j,k}(\texttt{W}, \texttt{FW}) .
 \]

### 2019-06-30 src/implementation.tex
ircad experiment added
my last update

@@
-\subsection{Julia implementation}\label{sec:implementation}
+\section{Julia implementation}\label{sec:implementation}
 
 The computer code is implemented in Julia language~\cite{} according to the workflow described below, whose stages are parallelized and/or optimized in various ways.
+
+\subsection{Parallel workflow}\label{sec:implementation}
 
@@
 
-\item depending on the above, best decision for $\emph{size}$ of 3D image blocks $B$ (bricks); defaults to: $\emph{size}=64$; hence the number of bricks will be $n=\ceil{\ell_1/size}\times\ceil{\ell_2/size}\times\ceil{\ell_3/size}$.
+\item depending on the above, best decision for $\emph{size}$ of 3D image blocks $\mathcal{B}$ (bricks); defaults to: $\emph{size}=64$; hence the number of bricks will be $n=\ceil{\ell_1/size}\times\ceil{\ell_2/size}\times\ceil{\ell_3/size}$.
 Hence default value: $n=256$, for images $\mathcal{I} = 512\times 512\times 256$;
 
-\item computation of a Julia's sparse boundary matrix $[\partial_B]$, returning a value of type \texttt{SparseMatrixCSC\{Int8\}\{Int64\}}, where \texttt{Int8} and \texttt{Int64} are the types for values and indices, respectively, stored by Compressed Sparse Column (CSC) format; the average $[\partial_B]$ (for $\emph{size}$ = 64) is about
+\item computation of a Julia's sparse boundary matrix $[\partial_B]$, returning a value of type \texttt{SparseMatrixCSC\{Int8\}\{Int64\}}, where \texttt{Int8} and \texttt{Int64} are the types for values and indices, respectively, stored by Compressed Sparse Column (CSC) format; the average $[\partial_B]$ (for $\emph{size}$ = 64) is about 45 MB;
 
-\item creation of a either a local or distributed channel to implement a producer/consumer model of parallel/distributed computation, depending on available resources;
+\item creation of either a local or distributed channel to implement a producer/consumer model of parallel/distributed computation, depending on available resources;
 
-\item distribution of matrix $[\partial_B]$, of size $45$ MB, to all available nodes/cores (workers) using the macro \texttt{@eveywhere}.
+\item distribution of matrix $[\partial_B]$, of default size $45$ MB, to all available nodes/cores (workers) using the macro \texttt{@eveywhere}.
 
@@
 
-With $size=64$, the number of non-zeros within the sparse matrix $[\partial_{64^3}]$ is $\mathtt{nnz} = 4 792 266$, for a memory size of $9\times \mathtt{nnz}+8\times 262144 \simeq 45$ MB.
-The memory size of the sparse matrix is computed by considering $8+1$ bytes for non-zero element (which are exactly 6 per row), plus 8 bytes per each index of column start.
+The memory occupancy of the sparse matrix is computed by considering $8+1$ bytes for non-zero element (which are exactly 6 per row), plus 8 bytes per each index of column start.  
+With $size=64$, the number of non-zeros within the sparse matrix $[\partial_{64^3}]$ is $\mathtt{nnz} = 4 792 266$, for a storage size of $9\times \mathtt{nnz}+8\times 262144 \simeq 45$ MB.
 
@@
 The goal of the whole \emph{workflow} is to extract a \emph{boundary model} of $\mathcal{S}$ from $\mathcal{I}$.
-The portion of $\mathcal{S}$ inside $B$, will be denoted as $\mathcal{S}(B)$.
-Each block $B$ of the 3D image must by converted into the \emph{coordinate representation} of a vector $\nu\in C_3$ in the linear space  of 3-chains.
+The portion of $\mathcal{S}$ inside $\mathcal{B}$, will be denoted as $\mathcal{S}(B)$.
+Each block $\mathcal{B}$ of the 3D image must by converted into the \emph{coordinate representation} of a vector $\nu\in C_3$ in the linear space  of 3-chains.
 
-In coordinates local to $B$, once fixed an ordering from Cartesian to linear coords, this vector is represented by a \emph{binary array} of length $size^3$.
+In coordinates local to $\mathcal{B}$, once fixed an ordering from Cartesian to linear coords, this vector is represented by a \emph{binary array} of length $size^3$.
 With $size=64$, we have $64^3=262144$,  with a non-zero value (i.e.~$1$) for each foreground voxel in $\mathcal{S}(B)$.
@@
 
-\item conversion from linear indices to Cartesian indices in coordinates local to the $B$ block, using the appropriate library functions;
+\item conversion from linear indices to Cartesian indices in coordinates local to the $\mathcal{B}$ block, using the appropriate library functions;
 
@@
 \paragraph{Assembling and artifact filtering} 
-The results of the previous stages can be described as a \emph{collection of sets} of \emph{geometric quadrilaterals (quads)}, each one encoded as an array of quadruples of integer indices, pointing to the linear array of grid vertices associated to the standard image block $B$.
+The results of the previous stages can be described as a \emph{collection of sets} of \emph{geometric quadrilaterals (quads)}, each one encoded as an array of quadruples of integer indices, pointing to the linear array of grid vertices associated to the image block $\mathcal{B}$.
 In other words, \emph{all quads} of \textbf{each job} are now given in the \textbf{same} \emph{local coordinates}.
-Besides to put each partial surface $\mathcal{S}(B) = (\texttt{V}_B, \texttt{FV}_\sigma)$ in the global coordinate system of the image, the present stage must eliminate the redundant boundary features possibly generated at the edges of the partial surface $\mathcal{S}(B)$ within each block $B$ such that $B \cap \mathcal{I} \not= \emptyset$:
+Besides to put each partial surface $\mathcal{S}(B) = (\texttt{V}_B, \texttt{FV}_\sigma)$ in the global coordinate system of the image, the present stage must eliminate the redundant boundary features possibly generated at the edges of the partial surface $\mathcal{S}(B)$ within each block $\mathcal{B}$ such that $B \cap \mathcal{I} \not= \emptyset$:
 \begin{enumerate}
 
-\item translate each array $\mathtt{FV}_\sigma$, of type \texttt{Lar.Cells}, by summing to each vertex index the linearized offset of the Cartesian coordinates $(n,m,p)(B)$ of the $B$'s \emph{reference vertex}, i.e.~the one with (all) lowest \emph{Cartesian coordinates} within the $B$ block.
+\item translate each array $\mathtt{FV}_\sigma$, of type \texttt{Lar.Cells}, by summing to each vertex index the linearized offset of the Cartesian coordinates $(n,m,p)(B)$ of the $\mathcal{B}$'s \emph{reference vertex}, i.e.~the one with (all) lowest \emph{Cartesian coordinates} within the $\mathcal{B}$ block.
 
@@
 It will consist in arranging each block with all three dimensions decreased-increased by one, so that each 2-adjacent pair of blocks will be covering each other for a full side extent of blocks of depth one.
-The details of this \emph{artifact filtering} will be elucidated in Section~\ref{sec:covering}.
+The details of this \emph{artifact filtering} are elucidated in Section~\ref{sec:covering}.
 
@@
 
-\subsubsection{Performance analysis}\label{sec:analysis}
+\subsection{Performance analysis}\label{sec:analysis}

### 2019-06-30 src/main.tex
tables added
my last update

@@
 %SetFonts
+\usepackage[utf8x]{inputenc}
 \usepackage[T1]{fontenc}
@@
 
-\title{Fast algebraic filtering of surfaces from 3D medical images with Julia}
-\author{Miroslav Ji\v{r}\'ik and Alberto Paoluzzi}
+\title{Algebraic filtering of surfaces from medical images with Julia}
+\author{Miroslav Jirik and Alberto Paoluzzi}
 \date{}							% Activate to display a given date or no date
@@
 \begin{abstract}
-In this paper we introduce a novel algebraic \textsc{lar-surf} filter, well founded on algebraic topology methods, to extract and smooth the boundary surface of any subset of voxels arising from the segmentation of a 3D medical image.
-The input is defined as a \emph{chain}, i.e.~as a vector from a linear space of 3-chains, represented in coordinates as a sparse Boolean vector.
-The output is produced as the result of the mapping via the linear boundary operator $\partial_3:C_3 \to C_2$ between linear spaces of 3- and 2-chains.
+In this paper we introduce a novel algebraic \textsc{lar-surf} filter, well founded on algebraic topology methods, to extract and smooth the boundary surface of any subset of voxels arising from  segmentation of a 3D medical image.
+The input is defined as a \emph{chain}, i.e.~as a vector from a linear space of 3-chains, represented in coordinates as a sparse binary vector.
+The output is produced by a linear mapping between spaces of 3- and 2-chains through the boundary operator $\partial_3:C_3 \to C_2$.
 In particular, when the input set of voxels is either not (4-)connected, or contains one or more empty regions inside, \textsc{lar-surf} generates a non connected set of closed surfaces, i.e.~a set of 2-cycles---using the language of algebraic topology.
@@
 \item[$\mathcal{I}$]  three-dimensional medical image
-\\[-8mm]
+\\[-7mm]
 \item[$\ell_1, \ell_2, \ell_3$]  dimensions of image
-\\[-8mm]
-\item[$\mathcal{S}$]  segment: subset of voxels
-\\[-8mm]
+\\[-7mm]
+\item[$\mathcal{S}$]  segment: a subset of voxels from image segmentation
+\\[-7mm]
 \item[$\mathcal{B}$] 3D image block 
-\\[-8mm]
+\\[-7mm]
 \item[$size$] lateral dimension of cubic block $\mathcal{B}$ 
-\\[-8mm]
+\\[-7mm]
 \item[$n$]	number of blocks $\mathcal{B}$ (jobs) in $\mathcal{I}$
-\\[-8mm]
+\\[-7mm]
 \item[[$\partial_\mathcal{B}$] ]  boundary matrix for block $\mathcal{B}$
-\\[-8mm]
+\\[-7mm]
 \item[$C_p$] linear (vector) space of $p$-chains
-\\[-8mm]
+\\[-7mm]
 \item[$\nu\in C_p$] $p$-chain
-\\[-8mm]
+\\[-7mm]
 \item[[$\nu$]] coordinate representation (binary vector) of  $\nu$
-\\[-8mm]
+\\[-7mm]
 \item[$\E^3$] Euclidean 3-space
@@
 \begin{description}
+
+\item[Boundary model] Closed manifold surface of the boundary of a solid model
+\\[-7mm]
 \item[CSC]  Compressed Sparse Column format for sparse matrices
-\\[-8mm]
+\\[-7mm]
 \item[Global coordinates]  Integer linear coordinates of $\mathcal{I}$
-\\[-8mm]
+\\[-7mm]
 \item[Local coordinates] Integer linear coordinates of $\mathcal{B}$
-\\[-8mm]
+\\[-7mm]
 \item[Cartesian coordinates] Integer triples $(i,j,k)$ one-to-one with voxels 
-\\[-8mm]
-\item[Voxels] individual elements in 3D image (3-cells)
-\\[-8mm]
-\item[$p$-chain] formal linear combination of $p$-cells with coefficients in $\{0,1\}$
-\\[-8mm]
-\item[Coord. repr.]  binary vector (for $p$-chains) or binary matrix (for chain operators)
-\\[-8mm]
-\item[Quad]	geometric quadrilateral; convex polygon with four vertices 
-\\[-8mm]
-\item[Foreground voxel] individual element of a segment $\mathcal{S}$
-\\[-8mm]
-\item[Segment]	subset of voxels resulting from image segmentation
+\\[-7mm]
+\item[Voxels] Individual elements in 3D image (3-cells)
+\\[-7mm]
+\item[$p$-chain] Formal linear combination of $p$-cells with coefficients in $\{0,1\}$
+\\[-7mm]
+\item[Coord. repr.]  Binary vector (for $p$-chains) or binary matrix (for chain operators)
+\\[-7mm]
+\item[Quad]	Geometric quadrilateral; convex polygon with four vertices 
+\\[-7mm]
+\item[Foreground voxel] Individual element of a segment $\mathcal{S}$
+\\[-7mm]
+\item[Segment]	Subset of voxels resulting from image segmentation
 \end{description}
 
+% Bibliography
+\bibliographystyle{acm}
+%\bibliography{TOMS2017.bib,proposal.bib,cadj2017.bib,bib.bib,schemes.bib}
+\bibliography{TSAS-2019.bib}
+
 \end{document}

### 2019-07-06 src/examples.tex
new examples

@@
 \section{Examples}\label{sec:examples}
+
+The size of boundary matrix is critical parameter of LAR-SURF method.
+To determine optimal size of boundary matrix the experiment on artifical data was performed (Fig.
+\ref{fig:bm_size_tesla}).
+Size of experimental data is set to $512\times512\times512$ and it is derived from typical size of Computed Tomography medical images.
+Computation is done on Tesla DGX-1 machine
+
+\begin{figure}
+\includegraphics[width=0.99\textwidth]{figs/bm_size_tesla.pdf} 
+%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
+\caption{Time requirements of LAR-SURF filter used on artifical with different size of boundary matrix}
+\label{fig:bm_size_tesla}
+\end{figure}
 
@@
 \end{table}
+
+We performed experiment on Ircad dataset to compare time requirements and used t-test  check the  Based on $\alpha=0.99$ the with the $p=\input{input/ircad_pvalue}$ , 
+$s=\input{input/ircad_statistics}$
+
+\begin{figure}
+\includegraphics[width=0.99\textwidth]{figs/ircad_comparison.pdf} 
+%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
+\caption{Time requirements of LAR-SURF filter and Marching Cubes on Ircad dataset. Error bars shows the 95\% confidence interval}
+\label{fig:ircad_comparison}
+\end{figure}

### 2019-07-07 src/implementation.tex
new examples added and text with statistic

--- 2019-06-12 src/implementation.tex
@@
 \end{enumerate}
+
+\begin{figure}[htbp]
+\includegraphics[width=0.99\textwidth]{figs/schema_horizontal.pdf} 
+%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
+\caption{Workflow of LAR-SURF algorithm}
+\label{fig:schema}
+\end{figure}
 
@@
 \subsubsection{Performance analysis}\label{sec:analysis}
+
+The size of boundary matrix is critical parameter of LAR-SURF method.
+To determine optimal size of boundary matrix the experiment on artifical data was performed (Fig.
+\ref{fig:bm_size_tesla}).
+Size of experimental data is set to $512\times512\times512$ and it is derived from typical size of Computed Tomography medical images.
+Computation is done on Tesla DGX-1 machine.
+
+\begin{figure}
+\centering
+\includegraphics[width=0.6\textwidth]{figs/bm_size_tesla.pdf} 
+%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
+\caption{Time requirements of LAR-SURF filter used on artifical with different size of boundary matrix}
+\label{fig:bm_size_tesla}
+\end{figure}
+
+To compare time requirements of LAR-SURF with Marching Cubes implemented in Python we performed experiment on Ircadb dataset \cite{ircad}.
+Dataset contain 20 Computed Tomography images (see table \ref{tab:ircad1}) with xy-resolution from 0.56 mm to 0.87 mm and z-resolution from 1.0 mm to 4.0 mm.
+The number of slices is each series varies from 74 to 260 and size of each slice is $512\times512$.
+Dataset contain manually segmented liver, portal vein and other structures.
+We performed surface extraction of liver with Marching Cubes and LAR-SURF.
+Time required for computation can be seen on figure \ref{fig:ircad_comparison}.
+
+Based on t-test with $\alpha=0.99$, $p=\input{input/ircad_pvalue}$ and 
+$s=\input{input/ircad_statistics}$ it can be shown that the mean of time consumed by LAR-SURF is significantly lower from time consumed by Marching Cubes.
+
+\begin{figure}
+\centering
+\includegraphics[width=0.99\textwidth]{figs/ircad_comparison.pdf} 
+%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
+\caption{Time requirements of LAR-SURF filter and Marching Cubes on Ircad dataset. Error bars shows the 95\% confidence interval}
+\label{fig:ircad_comparison}
+\end{figure}

### 2019-07-08 src/background.tex

@@
 \(\delta_p=\partial_{p+1}^\top\) between them.
-The domain of \textsc{lar} is the set of chain complexes generated by cell $d$-complexes ($2\leq d\leq 3$).
-The computer representations of \textsc{lar} are sparse binary matrices to represent both the operators and the chain bases.
+The \emph{domain} of \textsc{lar} is the set of \textbf{chain complexes} generated by cell $d$-complexes ($2\leq d\leq 3$).
+The computer \emph{representations} of \textsc{lar} are \textbf{sparse binary matrices} to represent both the operators and the chain bases.
 Note that in algebraic topology a $p$-chain is defined as a linear combination of $p$-cells with scalars from a field.

### 2019-11-10 src/background.tex
paper joined and partially revised

@@
 
-A \emph{representation scheme} for solid modeling is a mapping between a space of mathematical models and a space of symbolic representations, often generated by a formal grammar.
+A \emph{representation scheme} for solid modeling is a mapping between a space of mathematical models and a space of symbolic representations, like generated by a formal grammar.
 Solid pointsets (i.e., `$r$-sets') are defined~\cite{Requicha:1980:RRS:356827.356833} as compact (bounded and closed) regular and semianalytic subsets of the $d$-space.
@@
 
-The \emph{Linear Algebraic Representation} (\textsc{lar})~\cite{Dicarlo:2014:TNL:2543138.2543294} aims to represent the \emph{chain complex}~\cite{TSAS} generated by a piecewise-linear geometric complex embedded either in 2D or in 3D.
+The \emph{Linear Algebraic Representation} (\textsc{lar})~\cite{Dicarlo:2014:TNL:2543138.2543294} aims to represent the \emph{chain complex}~\cite{TSAS} generated by a piecewise-linear \emph{geometric complex} embedded either in 2D or in 3D.
 In few words, it gets a minimal characterization of geometry and topology of a cellular complex, i.e.~the embedding mapping $\mu : C_0 \to \E^d$ of 0-cells (vertices), as well a description of $(d-1)$-cells as subsets of vertices, and is able to return the whole chain complex
@@
 \substack{
-\delta_2 \\
+\delta_2 \\x
 \longleftarrow \\[-1mm]
@@
 boundary/coboundary map \(\partial_p\) and
-\(\delta_p=\partial_{p+1}^\top\) between them.
+\(\delta_p=\partial_{p-1}^\top\) between them.
 The \emph{domain} of \textsc{lar} is the set of \textbf{chain complexes} generated by cell $d$-complexes ($2\leq d\leq 3$).
@@
 
-We may therefore get the $(p-1)$-boundary $\partial_p \gamma_p$ of \emph{any} $p$-chain $\gamma_p$, by multiplication of the coordinate representation $[\partial_p]$ of the boundary operator times the coordinate representation $[\gamma_p]$ of the chain in terms of such scalars, i.e.~by a  matrix-vector product $ [\partial_p] [\gamma_p] $.
+We may therefore get the $(p-1)$-boundary $\partial_p c_p$ of \emph{any} $p$-chain $c_p$, by multiplication of the coordinate representation $[\partial_p]$ of the boundary operator times the coordinate representation $[c_p]$ of the chain in terms of such scalars, i.e.~by a  matrix-vector product $ [\partial_p] [c_p] $.
 
-It is easy to understand that the \textsc{lar} representation scheme is very expressive, i.e.~that it  has a large domain,  including collections of: line segments, quads, triangles, polygons, meshes;  pixels, voxels, volume images; B-reps, enumerative and decompositive representations of solids.
+It is possible to show that the \textsc{lar} representation scheme is very expressive, i.e.~that it  has a large domain,  including collections of: line segments, quads, triangles, polygons, meshes;  pixels, voxels, volume images; B-reps, enumerative and decompositive representations of solids.
 In this paper we apply \textsc{lar} methods to computation of boundary representations of solid models from segmentation (labeling) of 3D medical images.
+To display a triangulation of  boundary faces  in their proper position in space, the information required is contained in the \emph{geometric chain complex} (GCC):
+\[
+\mu: C_0\to\E^3,\ (\delta_0, \delta_1, \delta_2)
+\qquad\equiv\qquad
+\texttt{(geometry, topology) = (W, (EV, FE, CF))}
+\]
+The GCC allows to transform the (possibly non connected) boundary 2-cycle of surfaces as a standard B-rep~\cite{shapiroSM:202}. 
+The \texttt{geometry} is given by
+the embedding matrix \texttt{W} of vertices (0-cells), and \texttt{topology} is
+given by the three sparse matrices \texttt{CF}, \texttt{FE}, \texttt{EV} of coboundaries
+{i.e.},~$\delta_3, \delta_2, \delta_1$, of  chain complex describing a   
+space arrangement~\cite{paoluzzi2019finite}.
+Note that ordered pairs of letters from \texttt{V,E,F,C}, correspond to \emph{\emph{\texttt{V}}ertices$\to$\emph{\texttt{E}}dges$\to$\emph{\texttt{F}}aces$\to$\emph{\texttt{C}}ells} into the 
+\emph{\emph{\texttt{C}}olumn$\to$\emph{\texttt{R}}ow} order of matrix maps of operators.
 
-\subsubsection{Construction of $\partial_d$ boundary matrix}
+\subsubsection{Construction of boundary matrix $\partial_d$}
 
-Once fixed an ordering for all the cells (vertices, edges, pixels, and voxels), i.e.~for 0-, 1-, 2-, and 3-elements of a cell partitioning of a 3D image, i.e., once fixed the $p$-bases for linear spaces $C_p$ of $p$-chains $(0\leq p\leq d)$, we call $M_p = (m_{i,j})$ the \emph{characteristic matrix} of the $p$-basis, expressed as subsets of 0-cells, where  we have that $m_{i,j}=1$ iff the $j$-th $0$-cell belongs to the $i$-th $p$-cell, and $m_{i,j}=0$ otherwise.
+Once fixed an ordering for all the cells (vertices, edges, pixels, and voxels), i.e.~for 0-, 1-, 2-, and 3-elements of a cell partition \texttt{V,E,F,C} of a 3D image, i.e., once fixed the $p$-bases for linear spaces $C_p$ of $p$-chains $(0\leq p\leq 3)$, we call $M_p = (m_{i,j})$ the \emph{characteristic matrix} of the $p$-basis, expressed as subsets of 0-cells, where  we have that $m_{i,j}=1$ iff the $j$-th $0$-cell belongs to the boundary of $i$-th $p$-cell, and $m_{i,j}=0$ otherwise.
 
@@
 When this number equates the cardinality of $c_{p-1}^i$, this elementary chain is contained on the boundary of $c_{p}^j$.
-In a 3D image, with cubic 3-cells and squared 2-cells, everywhere we get $n_{i,j}=4$, we may state $c_{2}^i\subset\partial c_{3}^j$.
+In a 3D image, with cubic 3-cells and squared 2-cells in between, everywhere we get $n_{i,j}=4$, we may state $c_{2}^i\subset\partial c_{3}^j$.
 In this case, by looking in each $j$ column of $M_{2} M_{3}^t = (n_{i,j})$, we have exactly \emph{six rows} where  $n_{i,j} = 4$.
@@
 Of course, it will contain six non-zero elements for column.
-It may be worth to remember that every 3-cell (voxel) has exactly six 2-faces.
+It may be worth to remember that every 3-cell (voxel) of 3D image has exactly six 2-faces.
 
-It is possible to show that all the interesting relation of incidence/adjacency between cells of different dimensions can be both computed and efficiently queried by pairwise computing some products, with one of terms possibly transposed, of the two boundary and coboundary operator matrices $[\partial_p]$ and $[\delta_p] = [\partial_p^\top] = [\partial_p]^t$.
-We may also show that such matrices are \emph{very sparse}, with their sparseness growing rapidly with the dimensions (see Section~\ref{sec:block}).
+It is possible to show that all the interesting relations of incidence/adjacency between cells of different dimensions can be both computed and efficiently queried by pairwise computing some matrix products, with one of terms possibly transposed, using only the two boundary and coboundary operator matrices $[\partial_p]$ and $[\delta_p]$, and where $[\delta_p] = [\partial_p^\top]$.
+We~may also show that such matrices are \emph{very sparse}, with their sparseness growing rapidly with the dimensions (see Section~\ref{sec:block}).
 The pattern of non-zeros in matrix $[\partial_3]$ corresponding to a brick of shape $(4,4,4)$ is given in Figure~\ref{fig:boundary_matrix_4x4x4}.
@@
    \includegraphics[width=0.75\linewidth]{figs/boundary_matrix_4x4x4.pdf} 
-   \caption{A binary image of the coboundary operator  $\delta_2 = \partial^\top : C_2 \to C_3$, built for small 3D images with shape $(4,4,4)$.
+   \caption{A binary image of the coboundary operator  $\delta_2 = \partial_3^\top : C_2 \to C_3$, built for a small 3D image with shape $(4,4,4)$.
 Note that the number of rows equates the cardinality $4\times 4\times 4 = 64$ of the voxel set; the number of columns is $d\,n\,(1+n)^{d-1} = 3\times 4\times 25 = 300$.
-Of course, the number of non-zeros per row (cardinality of single voxel facets) is six.}
+Of course, the number of non-zeros per row (cardinality of single voxel facets) is six, whereas the number of non-zeros per column is two, but on boundary facets.}
    \label{fig:boundary_matrix_4x4x4}

### 2019-11-10 src/design.tex
paper joined and partially revised

@@
 We assume that medical devices produce 3D images with lateral dimensions that are integer multiples of some powers of two, like 128, 256, 512, etc.
-Any cuboidal portion of image is completely determined by the Cartesian indices of its voxel of lowest and highest indices, and extracted by multidimensional array \emph{slicing} as $Image([\ell_x$:$h_x, \ell_y$:$h_y, \ell_z$:$h_z])$.
+Any cuboidal portion of image is completely determined by the Cartesian indices of its voxels of lowest and highest indices, and extracted by multidimensional array \emph{slicing} as $Image([\ell_x$:$h_x, \ell_y$:$h_y, \ell_z$:$h_z])$.
 
-For the sake of simplicity, we assume a common size on the three image axes, and the corresponding image portion $\mathcal{B}$, called \emph{block}, as a function of its element of the  lowest  block  coordinates $i,j,k\in\N$ and of block dimension $n$:
+For the sake of simplicity, we assume a common size on the three image axes, and the corresponding image portion $\B$, called \emph{block}, as a function of its element of the  lowest  block  coordinates $i,j,k\in [1:n]$ and of block dimension $n\in\N$:
 \[
-\mathcal{B}(i,j,k,n) := Image([in\mbox{:}in+n, jn\mbox{:}jn+n, kn\mbox{:}kn+n]) 
+\B(i,j,k,n) := Image([in:in+n, jn:jn+n, kn:kn+n]) 
 \]
@@
    \includegraphics[width=0.5\linewidth]{figs/blocks} 
-   \caption{A possible block partitioning of a radiologic image. The evidenced 2D block, of size $n^d=64^2$, is sliced as $\mathcal{B}([2,1,64]) = Image([128:172],[64:128)]$}
+   \caption{A possible block partitioning of a radiologic image. The evidenced 2D block, of size $n^d=64^2$, is sliced by $\B([2,1,64]) = Image([128:172],[64:128)]$}
    \label{fig:blocks}
@@
 
-Figure~\ref{fig:blocks} shows the block decomposition in a 2D image, with positive integer $u,v$ lateral sizes of image. Note that block sides do not necessarily correspond to image edges.
+Figure~\ref{fig:blocks} shows the block decomposition in a 2D image, with positive integers $(u,v)$ giving the lateral sizes of image.
+Note that block sides do not necessarily correspond to image edges.
 
@@
 We are going to treat each image block independently from each other.
-Hence we map each image subspace $\mathcal{B}(i,j,n)$ to the linear \emph{chain} space $C_2$ of dimension $n\times n$, using coordinate vectors $c_{h,k}\in \\mathcal{B}^{n\times n} := \{0,1\}^{n\times n}$, where the basis element $c = c_{h,k} \in C_2$ is mapped via Cartesian-to-linear map to the binary vector
+Hence we map each image subspace $\B(i,j,n)$ to the linear \emph{chain} space $C_2$ of dimension $n\times n$, using coordinate vectors $c_{h,k}\in \\B^{n\times n} := \{0,1\}^{n\times n}$, where the basis element $c = c_{h,k} \in C_2$ is mapped via Cartesian-to-linear map to the binary vector
 \[
-Image(h,k) \mapsto c_{h,k} := [0 \cdots 0\ 1\ 0 \cdots 0] \in \\mathcal{B}^{n\times n}
+Image(h,k) \mapsto c_{h,k} := [0 \cdots 0\ 1\ 0 \cdots 0] \in \B^{n\times n}
 \]
@@
 \] of blocks. 
-Then we consider each image portion $c_{i,j,k} = S\cap \mathcal{B}(i,j,k,n)$ and compute its (binary) coordinate representation  $[c]_{i,j,k}\in C_3(n,n,n)$.
+Then we consider each image portion $c_{i,j,k} = S\cap \B(i,j,k,n)$ and compute its (binary) coordinate representation  $[c]_{i,j,k}\in C_3(n,n,n)$.
 This one is a sparse binary vector of length $n^3$.
 Then assemble the $M$ representations $c$ of segment portions into a sparse binary matrix $\T{S}$, of dimension $n^d \times M$.
-Finally compute a matrix $\T{\mathcal{B}}$ of boundary portions of $S$, represented by columns as chain coordinate vectors in $C_2$:
+Finally compute a matrix $\T{\B}$ of boundary portions of $S$, represented by columns as chain coordinate vectors in $C_2$:
 \[
-\T{\mathcal{B}} = [\partial_3(n)]\, \T{S}.
+\T{\B} = [\partial_3(n)]\, \T{S}.
 \]
 where the boundary matrix has dimension $d\,n\,(1+n)^{d-1} \times n^d$.
-Of course, the $\T{\mathcal{B}}$ sparse matrix has the same column number $M$ of $\T{S}$, because each column contains the boundary representation of the corresponding $S\cap Box_{i,j,k}{i,j,k}$, and the number of rows of the operator, equal to the dimension of the linear space $C_2$.
+Of course, the $\T{\B}$ sparse matrix has the same column number $M$ of $\T{S}$, because each column contains the boundary representation of the corresponding $S\cap Box_{i,j,k}{i,j,k}$, and the number of rows of the operator, equal to the dimension of the linear space $C_2$.
 
@@
 In particular, we need to compute the \emph{embedding function} $\mu : C_0 \to \E^3$, where $C_0$ is the space of 0-chains, one-to-one corresponding to the vertices of the extracted surface.
-The simplest solution is to associate  four 0-cells to each 2-cell of the extracted surface, i.e.~to each non-zero entry in every column of $\T{\mathcal{B}}$.
+The simplest solution is to associate  four 0-cells to each 2-cell of the extracted surface, i.e.~to each non-zero entry in every column of $\T{\B}$.
 The $\mu$ function  can be computed by identifying, via  element position in the column, a triple of integer values $0\leq x\leq u$, $0\leq y\leq v$, and $0\leq z\leq w$ for each vertex of the 2-cell.
@@
 
-All boundary surface subsets $\mathcal{B}_{i,j,k}(S) = \partial_3 S \cap \mbox{Box}_{i,j,k}$, provided by  columns of $\T{S}$, are embedded in the same coordinate space. In formal terms: 
+All boundary surface subsets $\B_{i,j,k}(S) = \partial_3 S \cap \mbox{Box}_{i,j,k}$, provided by  columns of $\T{S}$, are embedded in the same coordinate space. In formal terms: 
 \[
@@
 \\
-\texttt{Top} &:= C_2(B_{i,j,k}) = \T{\mathcal{B}}_{i,j,k} \mapsto \texttt{FW} \subset \texttt{FV}.
+\texttt{Top} &:= C_2(B_{i,j,k}) = \T{\B}_{i,j,k} \mapsto \texttt{FW} \subset \texttt{FV}.
 \end{align}
@@
 \[
-\texttt{Lar}(\mathcal{B}) = \oplus_{i,j,k}\texttt{Lar}(\partial_3 S_{i,j,k})) = \oplus_{i,j,k}(\texttt{W}, \texttt{FW}) .
+\texttt{Lar}(\B) = \oplus_{i,j,k}\texttt{Lar}(\partial_3 S_{i,j,k})) = \oplus_{i,j,k}(\texttt{W}, \texttt{FW}) .
 \]

### 2019-11-10 src/paper.tex
paper joined and partially revised

\documentclass[11pt, oneside]{amsart}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   		% ... or a4paper or a5paper or ...
%\geometry{landscape}                		% Activate for rotated page geometry
%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{hyperref}

%SetFonts
\usepackage[utf8x]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

\nonstopmode
\input{macros}

% rules for for inputed tables
% the
\usepackage{siunitx}
\usepackage{booktabs}

%alternative rules for tables
%\newcommand{\toprule}{\hline}
%\newcommand{\midrule}{\hline}
%\newcommand{\bottomrule}{\hline}
%\newcommand{\specialrule}{}

%SetFonts

\title{Algebraic filtering of surfaces from medical images with Julia}
\author{Miroslav Jirik and Alberto Paoluzzi}
\date{}							% Activate to display a given date or no date

\begin{document}
\maketitle

\begin{abstract}
In this paper we introduce a novel algebraic \textsc{lar-surf} filter, well founded on algebraic topology methods, to extract and smooth the boundary surface of any subset of voxels arising from  segmentation of a 3D medical image. The input is defined as a \emph{chain}, i.e.~as a vector from a linear space of 3-chains, represented in coordinates as a sparse binary vector. The output is produced by a linear mapping between spaces of 3- and 2-chains through the boundary operator $\partial_3:C_3 \to C_2$.  In particular, when the input set of voxels is either not (4-)connected, or contains one or more empty regions inside, \textsc{lar-surf} generates a non connected set of closed surfaces, i.e.~a set of 2-cycles---using the language of algebraic topology. The only data structures used by this approach are sparse arrays with one or two indices, i.e.~sparse vectors and matrices. This work is based on LAR (Linear Algebraic Representation) methods, and is implemented in Julia language, natively supporting parallel computing on hybrid architectures.
\end{abstract}

\tableofcontents

%\section{}
%\subsection{}

%\input{intro}
\section{Introduction}\label{sec:intro}

Isosurface extraction to produce geometric models of surfaces from volumetric data is important in many aplications. It is often used for indirect visualization of the medical data or for flow modeling \cite{Rohan2018a}. 
In this paper we discuss an approach based on basic algebraic topology and linear algebra, using linear spaces $C_p$ of chains (of cells) of dimension $0 \leq p \leq 3$ and the boundary matrix $[\partial_3] : C_3 \to C_2$.
A decomposition of the 3D image into small submatrices called ``bricks'' is performed, the binary coordinate vector of each interesting chain of voxels is generated, and its boundary is computed by matrix multiplication times the boundary matrix. An embarrassing parallel approach is used to compute the boundary surface patches within each brick, that are finally joined.

The most popular algorithm used for surface extraction is probably Marching Cubes (MC). The algorithm was described by Lorentsen and Cline \cite{Lorensen1987} in 1987. Survey of Marching Cubes algorthm has been published in 2006 \cite{Newman2006}. The algorithm is based on considering the cube defining volume. Each corner vertex of the cube is related to input volumetric data. MC traverse the data and constructs the surface by using lookup table of different triangular faces depending on different patterns of the cube.  Main disadvantages of this method are time requirements, ambiquity and holes generation. Some of them were discovered shortly after the algorithm was introduced. 
Marching Cubes. In 1991 Nielson and Hamman described an Asymptotic Decider to solve the ambiguity problem on the faces of the cube.  Natarajan noted that the ambiguity problem also occurs with uniform samples \cite{Natarajan1994}. In 1995 Chernyaev extended the number of cases to 33 \cite{chernyaev1995marching}. More recently the algorithm was updated by Custodio, Pesco and Silva to enhance the quality of iso-surface triangulation \cite{Custodio2019}.

Some alternative methods have been developed, including a method for surface extraction using particle attraction; a system was described by Crossno and Angel in \cite{Crossno2002}. A graph processing that tracks the boundary cell-face adjacencies is described in \cite{Lachaud2000}. Some parallel algorithms for iso-surface extraction are discussed in \cite{Bajaj2004}.
A completely data-parallel algorithm, implementation written in OpenCL that runs entirely on the GPU is presented in~\cite{Smistad:12}. 
A Linear Algebraic Representation algebraic approach, parallelized using the OpenCL framework on Linux was discussed in~\cite{paodcvjcadanda2015}.

The present paper is organized as follows.
Section~\ref{sec:background} provides the basic topological and geometrical concepts needed to understand the \textsc{lar-surf} method, including the building of boundary matrices, the map from Cartesian indices to linear indices, and the Taubin smoothing method.
Section~\ref{sec:filter} discusses the parametric design of the unit block filtered by the parallel algorithm, including the block decomposition, the sparsity rate of the used sparse arrays, the block-level parallelism.
Section~\ref{sec:julia} is related to the algorithm implementation in Julia, and in particular to a discussion of the parallel workflow.
Section~\ref{sec:examples} presents some examples of algorithm execution on the liver and the hepatic portal system.
Section~\ref{sec:conclusion} shortly describes the next extensions of this approach, in particular to implementation with Julia support for GPU parallelism and the multi-segmentation of Medical images.

%\input{background}
\section{Background}\label{sec:background}

Some basic concepts of solid modeling, and in particular the foundational idea of representation scheme, as well as basic concepts of algebraic topology, are shortly introduced in this Section, including the computation of the matrix of a boundary operator between chain spaces.

\subsection{Representation Schemes}\label{sec:aaaa}

A \emph{representation scheme} for solid modeling is a mapping between a space of mathematical models and a space of symbolic representations, like generated by a formal grammar.
Solid pointsets (i.e., `$r$-sets') are defined~\cite{Requicha:1980:RRS:356827.356833} as compact (bounded and closed) regular and semianalytic subsets of the $d$-space. A large number of representation schemes were defined in the past forty years, including the two main classes of (a) \emph{boundary representations} (`$B$-reps'), where the solid model is represented through a representation of its boundary elements, i.e.~faces, edges and vertices, and (b) \emph{decompositive/enumerative representations}, that are a decomposition of either the object or the embedding space, respectively, into a well-defined \emph{cellular complex}. In particular, a boundary representation provides a cellular decomposition of the object's boundary into \emph{cells} of dimension zero (vertices), one (edges), and two (faces). Medical imaging can be classified as \emph{enumerative representation} of cellular decompositions of organs and tissues of interest, in particular, as subsets \emph{of 3D volume elements} (voxels) from the 3D image.

\subsection{Linear Algebraic Representation}\label{sec:aaaa}

The \emph{Linear Algebraic Representation} (\textsc{lar}), introduced in~\cite{Dicarlo:2014:TNL:2543138.2543294}, aims to represent the \emph{chain complex}~\cite{TSAS} generated by a piecewise-linear \emph{geometric complex} embedded either in 2D or in 3D. In few words, it gets a minimal characterization of geometry and topology of a cellular complex, i.e.~the embedding mapping $\mu : C_0 \to \E^d$ of 0-cells (vertices), as well a description of $(d-1)$-cells as subsets of vertices, and is able to return the whole chain complex 
%\[ 
%C_\bullet = (C_p, \partial_p) := 
%C_3 \ 
%\substack{
%\delta_2 \x
%\longleftarrow \[-1mm]
%\longrightarrow \
%\partial_3 
%}
%\ C_2 \ 
%\substack{
%\delta_1 \
%\longleftarrow \[-1mm]
%\longrightarrow \
%\partial_2 
%}
%\ C_1 \ 
%\substack{
%\delta_0 \
%\longleftarrow \[-1mm]
%\longrightarrow \
%\partial_1 
%}
%\ C_0 .
%\]

{   
   \centering\vspace{1mm}
   \includegraphics[width=0.475\textwidth]{figs/complex}

}

\noindent
and, in particular, any basis for linear chain spaces $C_p$, and any linear
boundary/coboundary map \(\partial_p\) and
\(\delta_p=\partial_{p-1}^\top\) between them. The \emph{domain} of \textsc{lar} is the set of \textbf{chain complexes} generated by cell $d$-complexes ($2\leq d\leq 3$). The computer \emph{representations} of \textsc{lar} are \textbf{sparse binary matrices} to represent both the operators and the chain bases. Note that in algebraic topology a $p$-chain is defined as a linear combination of $p$-cells with scalars from a field. When the scalar coefficients are from $\{-1, 0, +1\}$, a chain may represent \emph{any (oriented) subset of cells} from the cellular complex.

We may therefore get the $(p-1)$-boundary $\partial_p c_p$ of \emph{any} $p$-chain $c_p$, by multiplication of the coordinate representation $[\partial_p]$ of the boundary operator times the coordinate representation $[c_p]$ of the chain in terms of such scalars, i.e.~by a  matrix-vector product $ [\partial_p] [c_p] $.

It is possible to show that the \textsc{lar} representation scheme is very expressive, i.e.~that it  has a large domain,  including collections of: line segments, quads, triangles, polygons, meshes;  pixels, voxels, volume images; B-reps, enumerative and decompositive representations of solids. 
In this paper we apply \textsc{lar} methods to computation of boundary representations of solid models from segmentation (labeling) of 3D medical images.
To display a triangulation of  boundary faces  in their proper position in space, the information required is contained in the \emph{geometric chain complex} (GCC):
\[
\mu: C_0\to\E^3,\ (\delta_0, \delta_1, \delta_2)
\qquad\equiv\qquad
\texttt{(geometry, topology) = (W, (EV, FE, CF))}
\]
The GCC allows to transform the (possibly non connected) boundary 2-cycle of surfaces as a standard B-rep~\cite{shapiroSM:202}. 
The \texttt{geometry} is given by
the embedding matrix \texttt{W} of vertices (0-cells), and \texttt{topology} is
given by the three sparse matrices \texttt{CF}, \texttt{FE}, \texttt{EV} of coboundaries
{i.e.},~$\delta_3, \delta_2, \delta_1$, of  chain complex describing a   
space arrangement~\cite{paoluzzi2019finite}.
Note that ordered pairs of letters from \texttt{V,E,F,C}, correspond to \emph{\emph{\texttt{V}}ertices$\to$\emph{\texttt{E}}dges$\to$\emph{\texttt{F}}aces$\to$\emph{\texttt{C}}ells} into the 
\emph{\emph{\texttt{C}}olumn$\to$\emph{\texttt{R}}ow} order of matrix maps of operators.

\subsubsection{Construction of boundary matrix $\partial_d$}

Once fixed an ordering for all the cells (vertices, edges, pixels, and voxels), i.e.~for 0-, 1-, 2-, and 3-elements of a cell partition \texttt{V,E,F,C} of a 3D image, i.e., once fixed the $p$-bases for linear spaces $C_p$ of $p$-chains $(0\leq p\leq 3)$, we call $M_p = (m_{i,j})$ the \emph{characteristic matrix} of the $p$-basis, expressed as subsets of 0-cells, where  we have that $m_{i,j}=1$ if and only if the $j$-th $0$-cell belongs to the boundary of $i$-th $p$-cell, and $m_{i,j}=0$ otherwise.

Note that the product of binary matrices is not binary, so by computing the (sparse) matrix product $(M_{p-1} M_{p}^t) = (n_{i,j})$, with $n_{i,j} = \sum_{k} m_{i,k}m_{k,j}$, we get for each $n_{ij}$ the \emph{number of vertices} shared by $c_{p-1}^i$ and $c_{p}^j$. When this number equates the cardinality of $c_{p-1}^i$, this elementary chain is contained on the boundary of $c_{p}^j$. In a 3D image, with cubic 3-cells and squared 2-cells in between, everywhere we get $n_{i,j}=4$, we may state $c_{2}^i\subset\partial c_{3}^j$. In this case, by looking in each $j$ column of $M_{2} M_{3}^t = (n_{i,j})$, we have exactly \emph{six rows} where  $n_{i,j} = 4$.

Finally, consider the linear graded boundary operator $\partial_p : C_p \to C_{p-1}$. As such, it contains by columns the representation of domain basis elements, expressed as linear combination of the basis elements of the range space. Therefore, the operator matrix $[\partial_d]$ is readily obtained by setting $[\partial_d](i,j)=1$ iff $n_{i,j}=4$ and $[\partial_d](i,j)=0$ otherwise.  Of course, it will contain six non-zero elements for column.  It may be worth to remember that every 3-cell (voxel) of 3D image has exactly six 2-faces.

It is possible to show that all the interesting relations of incidence/adjacency between cells of different dimensions can be both computed and efficiently queried by pairwise computing some matrix products, with one of terms possibly transposed, using only the two boundary and coboundary operator matrices $[\partial_p]$ and $[\delta_p]$, and where $[\delta_p] = [\partial_p^\top]$. We~may also show that such matrices are \emph{very sparse}, with their sparseness growing rapidly with the dimensions (see Section~\ref{sec:block}). The pattern of non-zeros in matrix $[\partial_3]$ corresponding to a brick of shape $(4,4,4)$ is given in Figure~\ref{fig:boundary_matrix_4x4x4}.

\begin{figure}[tbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=0.75\linewidth]{figs/boundary_matrix_4x4x4.pdf} 
   \caption{A binary image of the coboundary operator  $\delta_2 = \partial_3^\top : C_2 \to C_3$, built for a small 3D image with shape $(4,4,4)$. Note that the number of rows equates the cardinality $4\times 4\times 4 = 64$ of the voxel set; the number of columns is $d\,n\,(1+n)^{d-1} = 3\times 4\times 25 = 300$. Of course, the number of non-zeros per row (cardinality of single voxel facets) is six, whereas the number of non-zeros per column is two, but on boundary facets.}
   \label{fig:boundary_matrix_4x4x4}
\end{figure}

\subsection{Multiindices from Cartesian indices}\label{sec:aaaa}

In order to utilize the topological algebra shortly recalled in this paper, we need to explicitly sort the cells of the various dimensions into linearly ordered sequences, possibly according to the linear order their information is linearly accommodated in computer storage.

\subsection{Taubin Smoothing}\label{sec:aaaa}

Every boundary chain extracted from an image block $\B(i,j,n)$ is a \emph{2-cycle}, i.e., a closed 2-chain---in other words, a 2-chain with empty boundary. Such 2-cycles are joined together by removing the double 2-cells (at the boundaries of bricks) after having suitably shifted their indices to an unique linear representation of the whole image. The resulting raster surface is made by mutually orthogonal raster facets, that must be smoothed in order to get a fair surface. A linear time and space algorithm for this purpose is the Laplacian smoothing, which iteratively  moves each vertex (0-cell) to the centroid of its neighbors. A well known weakness of this simple algorithm is the asymptotic convergence of the whole mesh to a single point, resulting in unfair size reduction even after few iterations.  The Taubin smoothing algorithm~\cite{Taubin:1995:SPA:218380.218473,egst.20001029} alternates two Laplacian smoothing steps with shrink and inflate effects respectively, with the result of letting pretty invariant sizes and volume of the smoothed mesh.
The best results are obtained on meshes which have small variations of edge length and face angles, like for surfaces extracted from 3D raster images, as in our case.

%\input{design}

\section{Block-parametric design}\label{sec:filter}

\subsection{Block decomposition}\label{sec:bbbb}

We assume that medical devices produce 3D images with lateral dimensions that are integer multiples of some powers of two, like 128, 256, 512, etc.
Any cuboidal portion of image is completely determined by the Cartesian indices of its voxels of lowest and highest indices, and extracted by multidimensional array \emph{slicing} as $Image([\ell_x$:$h_x, \ell_y$:$h_y, \ell_z$:$h_z])$.

For the sake of simplicity, we assume a common size on the three image axes, and the corresponding image portion $\B$, called \emph{block}, as a function of its element of the  lowest  block  coordinates $i,j,k\in [1:n]$ and of block dimension $n\in\N$:
\[
\B(i,j,k,n) := Image([in:in+n, jn:jn+n, kn:kn+n]) 
\]

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=0.5\linewidth]{figs/blocks} 
   \caption{A possible block partitioning of a radiologic image. The evidenced 2D block, of size $n^d=64^2$, is sliced by $\B([2,1,64]) = Image([128:172],[64:128)]$}
   \label{fig:blocks}
\end{figure}

Figure~\ref{fig:blocks} shows the block decomposition in a 2D image, with positive integers $(u,v)$ giving the lateral sizes of image. Note that block sides do not necessarily correspond to image edges.

\subsection{Block operator }\label{sec:block}

\paragraph{\emph{Chain coordinates }\label{sec:bbbb}}
We are going to treat each image block independently from each other. Hence we map each image block $\B(i,j,n)$ to the linear \emph{chain} space $C_2$ of dimension $n\times n$, using coordinate vectors $c_{h,k}\in \B^{n\times n} := \{0,1\}^{n\times n}$, where the basis element $c = c_{h,k} \in C_2$ is mapped via Cartesian-to-linear map to the binary vector 
\[
Image(h,k) \mapsto c_{h,k} := [0 \cdots 0\ 1\ 0 \cdots 0] \in \B^{n\times n}
\]
for each $0\leq h,k \leq n$, and where the (single) unit element is in position $nk + h \leq n\times n$.

Therefore, each pixel (or voxel) in a block image will be seen as a basis binary vector in $C_2$, and each subset of image elements, as the corresponding binary vector in $C_2$, with many ones as the cardinality of the subset.

\paragraph{\emph{Boundary operator }\label{sec:bbbb}}For a fixed block size $n$, the boundary operator $\partial_d : C_d\to C_{d-1}$, with $d\in\{2,3\}$, will be constructed once and for all using the algorithm given in~\cite{}, and inlined in the generated boundary extraction code.

It is easy to see that the operator's matrix $[\partial_d]$ is \emph{very sparse}, since it contains $2\times d$ non-zero elements (ones) for each column (of length $n^d$), i.e.~4 ones and 6 ones for the 2D and 3D case, respectively. In fact the matrix of a linear operator between linear spaces contains by columns the basis element of the domain space, represented in the target space. In our case, the former is an image element (2-cube or 3-cube), represented as the chain of its boundary---i.e. either a 1-cycle of 4 edges, or  a 2-cycle of 6 faces, respectively.

The number of rows of $[\partial_d]$ equates the dimension of the linear space $C_{d-1}$, i.e.~the number of $(d-1)$-cells---elementary $(d-1)$-chains---in the cellular partition of the image. To compute their number, we act in two steps. (a) First we map one-to-one the $n^d$ $d$-cells with $d$ adjacent $(d-1)$-cells, so getting $d\,n^d$ distinct basis elements of $C_{d-1}$. (b) Then we complete this bases by adjoining $n^{d-1}$ boundary elements for each of the $d$ dimensions of the image, so providing further $d\,n^{d-1}$ basis elements for $C_{d-1}$. The dimension of $C_{d-1}$, and therefore the number of rows of $[\partial_d]$ matrix is $d\,(n^{d-1}+n^{d}) = d\,n\,(1+n)^{d-1}$. The number of column equates the number of basis elements of $C_d$, i.e.~the number $n^d$ of block elements.

\paragraph{\emph{Sparsity and size of boundary matrix }\label{sec:bbbb}}

As we have seen, we have $2d$ non-zero elements for each column of $[\partial_d]$, so that their total number is $2d\,n^d$. The number of matrix element is $d\,n\,(1+n)^{d-1} \times n^d$, giving a ratio of 
\[
\frac{\mbox{non-zero\ elements}}{\mbox{total\ elements}} = 
\frac{2d\times n^d}{d\,n\,(1+n)^{d-1} \times n^d} =
\frac{2}{n+n^d}
\]
Using sparse matrices in CSC (Compressed Sparse Column) format we get a storage size:
\[
mem([\partial_d]_{n^d}) = 2\times \#\mbox{nzero} + \#\mbox{columns} = 2\times 2d\,n^d + n^d = (4d+1)n^d.
\]
In conclusion, for block size $n=64$, the matrix $[\partial_d]$ requires for 2D images $9\times 64^2=36,864$ memory elements, and for 3D images $13\times 64^3=3,407,872$ memory elements. Counting the bytes for the standard implementation of a sparse binary matrix (1 byte for values and 8 bytes for indices) we get $(18d+8)n^d$ bytes, giving $176$\,KB for 2D and $15.872$\,MB for 3D.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=0.4\linewidth]{figs/grid.pdf} 
   \caption{example caption}
   \label{fig:example}
\end{figure}

\subsection{Block boundary mapping}\label{sec:bbbb}

Here we refer directly to the 3D case.
Let us call \emph{segment} the bulk content $S$ of interest within the input 3D image of size $(u,v,w)$. We aim to compute the segment boundary $\partial_3 S$. 
First we set the size $n$ of the block, in order to decompose the input $Image(u,v,w)$ into a fair number 
\[
M = \ceil{u/n} \times \ceil{v/n} \times \ceil{w/n} \simeq \frac{uvw}{n^3}
\] of blocks. 
Then we consider each image portion $c_{i,j,k} = S\cap \B(i,j,k,n)$ and compute its (binary) coordinate representation  $[c]_{i,j,k}\in C_3(n,n,n)$. This one is a sparse binary vector of length $n^3$. Then assemble the $M$ representations $c$ of segment portions into a sparse binary matrix $\T{S}$, of dimension $n^d \times M$. Finally compute a matrix $\T{\B}$ of boundary portions of $S$, represented by columns as chain coordinate vectors in $C_2$:
\[
\T{\B} = [\partial_3(n)]\, \T{S}.
\]
where the boundary matrix has dimension $d\,n\,(1+n)^{d-1} \times n^d$.
Of course, the $\T{\B}$ sparse matrix has the same column number $M$ of $\T{S}$, because each column contains the boundary representation of the corresponding $S\cap Box_{i,j,k}{i,j,k}$, and the number of rows of the operator, equal to the dimension of the linear space $C_2$.

\paragraph{Embedding}
A final computational step is needed, in order to embed the 2-chains in Euclidean space $\E^3$ and to assemble the whole resulting surface. In particular, we need to compute the \emph{embedding function} $\mu : C_0 \to \E^3$, where $C_0$ is the space of 0-chains, one-to-one corresponding to the vertices of the extracted surface. The simplest solution is to associate  four 0-cells to each 2-cell of the extracted surface, i.e.~to each non-zero entry in every column of $\T{\B}$.  The $\mu$ function  can be computed by identifying, via  element position in the column, a triple of integer values $0\leq x\leq u$, $0\leq y\leq v$, and $0\leq z\leq w$ for each vertex of the 2-cell.  The mapping can be implemented using a dictionary, that will store the inverse coordinate transformation used at the beginning, i.e.~the one from linear to Cartesian coords, in order of not duplicating the output vertices.

\paragraph{Surface assembling}

All boundary surface subsets $\B_{i,j,k}(S) = \partial_3 S \cap \mbox{Box}_{i,j,k}$, provided by  columns of $\T{S}$, are embedded in the same coordinate space. In formal terms: 
\[
\texttt{Lar}(S) := (\texttt{Geom}(S), \texttt{Top}(S)) = (\texttt{V}, \texttt{CV}),
\]
where, with respect to the \emph{chain complex} $C_3\to C_2\to C_1\to C_0$ induced by the input image $Im$ and segment $S_{i,j,k}$, we get
\begin{align}
\texttt{Geom} &:= \mu(C_0(_{i,j,k})) = \texttt{V},
\\
\texttt{Top} &:= C_3(S) = \T{S} \mapsto \texttt{CV}.
\end{align}
and
\begin{align}
\texttt{Lar}(B_{i,j,k}) &:= (\texttt{Geom}, \texttt{Top}) = (\texttt{W}, \texttt{FW}),
\\
\texttt{Geom} &:= \mu(C_0(B_{i,j,k})) = \texttt{W} \subset \texttt{V},
\\
\texttt{Top} &:= C_2(B_{i,j,k}) = \T{\B}_{i,j,k} \mapsto \texttt{FW} \subset \texttt{FV}.
\end{align}

A translation transformation applied to each vertex subset $\texttt{W}_{i,j,k}$ with translation  vector $\v{t} = [i,j,k]$ will therefore move it in the final space position, so finally giving
\[
\texttt{Lar}(\B) = \oplus_{i,j,k}\texttt{Lar}(\partial_3 S_{i,j,k})) = \oplus_{i,j,k}(\texttt{W}, \texttt{FW}) .
\]

\subsection{Block-level parallelism}\label{sec:bbbb}

In the computational pipeline introduced in this paper, several steps can be efficiently performed in parallel at image-block level, depending on the embarassingly data parallel nature of the problem. In particular, little effort is needed to separate the problem into a number of parallel tasks $S_{i,j,k}$, using multiarray slicing. The granularity of parallelism, depending on the block size $n$, is further enforced by the computation of a single boundary matrix $[\partial_d(n)]$, in turn depending on $n$, so that the initial communication cost of broadcasting the matrix to nodes can be carefully controlled, and finely tuned depending on the system architecture. The whole approach is appropriate  for SIMD hybrid architectures of CPUs and GPUs, since only the initial block setup of boundary matrix and image slices, as well the final collection of computed surface portions, require inter-process communication.

%\input{implementation}

\section{Julia implementation}\label{sec:julia}

The computer code is implemented in Julia language~\cite{BEKS14} according to the workflow described below, whose stages are parallelized and/or optimized in various ways.

\subsection{Parallel workflow}\label{sec:implementation}

\paragraph{\emph{Workflow setup}}The functions in this preliminary step include:
\begin{enumerate}

\item input of 3D medical image $\mathcal{I}$ dimensions $\ell_1, \ell_2, \ell_3$, such that: $\mathcal{I} = [\ell_1]\times[\ell_2]\times[\ell_3]$, where $[\ell_k] = (1,2,\ldots,\ell_k)$;

\item analysis of resources available in the computational environment, including operating system, type and number of compute nodes (processors, cores, GPUs), number of cores per node, RAM and caches amounts;

\item depending on the above, best decision for $\emph{size}$ of 3D image blocks $\mathcal{B}$ (bricks); defaults to: $\emph{size}=64$; hence the number of bricks will be $n=\ceil{\ell_1/size}\times\ceil{\ell_2/size}\times\ceil{\ell_3/size}$. 
Hence default value: $n=256$, for images $\mathcal{I} = 512\times 512\times 256$;

\item computation of a Julia's sparse boundary matrix $[\partial_B]$, returning a value of type \texttt{SparseMatrixCSC\{Int8\}\{Int64\}}, where \texttt{Int8} and \texttt{Int64} are the types for values and indices, respectively, stored by Compressed Sparse Column (CSC) format; the average $[\partial_B]$ (for $\emph{size}$ = 64) is about 45 MB;

\item creation of either a local or distributed channel to implement a producer/consumer model of parallel/distributed computation, depending on available resources;

\item distribution of matrix $[\partial_B]$, of default size $45$ MB, to all available nodes/cores (workers) using the macro \texttt{@eveywhere}.

\end{enumerate}

\begin{figure}[tbp]
\includegraphics[width=\textwidth]{figs/schema_horizontal.pdf} 
%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
\caption{Workflow of LAR-SURF algorithm}
\label{fig:schema}
\end{figure}

With $size=64$, the number of non-zeros within the sparse matrix $[\partial_{64^3}]$ is $\mathtt{nnz} = 4 792 266$, for a memory size of $9\times \mathtt{nnz}+8\times 262144 \simeq 45$ MB. The memory size of the sparse matrix is computed by considering $8+1$ bytes for non-zero element (which are exactly 6 per row), plus 8 bytes per each index of column start.

\paragraph{\emph{Job enqueuing}}Communication and data synchronization may be managed through \emph{Channels}, which are the FIFO conduits that may provide producer/consumer communication. Overall execution time can be improved if other tasks can be run while a task is being executed, or while waiting for an external service/function to complete. The single work items of this stage follow:
\begin{enumerate}

\item extraction, from image arrays of the block views, depending on 3 Cartesian indices;

\item transform  each block \emph{from global} $[\ell_1]\times[\ell_2]\times[\ell_3]$ to \emph{local coordinates} $[n]\times[n]\times[n]$;

\item further transform of each \emph{foreground voxel} $\nu\in\mathcal{S}\subseteq\mathcal{I}$ from Cartesian to linear coordinates, using the suitable Julia's library functions.

\item enqueing the job (as a sequence of integer positions for the non-zeros image elements aligned in a memory buffer of proper \texttt{Channel} type).
\end{enumerate}

\paragraph{\emph{3-Chain encoding}}
The interesting part of the \emph{Image} $\mathcal{I}$ is called \emph{Segment} $\mathcal{S}$. The goal of the whole \emph{workflow} is to extract a \emph{boundary model} of $\mathcal{S}$ from $\mathcal{I}$. The portion of $\mathcal{S}$ inside $\mathcal{B}$, will be denoted as $\mathcal{S}(B)$.
Each block $\mathcal{B}$ of the 3D image must by converted into the \emph{coordinate representation} of a vector $\nu\in C_3$ in the linear space  of 3-chains.

In coordinates local to $\mathcal{B}$, once fixed an ordering from Cartesian to linear coords, this vector is represented by a \emph{binary array} of length $size^3$. With $size=64$, we have $64^3=262144$,  with a non-zero value (i.e.~$1$) for each foreground voxel in $\mathcal{S}(B)$. Therefore, the coded segment portion $\mathcal{S}(B)$, results with a space occupancy of about $262$ KB if encoded as a full array (i.e.~including the zero values). Whether encoded as a sparse vector, its space occupancy will correspondingly decrease.

\begin{enumerate}

\item each encoding task produces either a full or sparse binary vector. With full or sparse arrays depending by one index, we get either 262 KB or less per job, correspondingly;

\item special format for sparse CSC (Compressed Sparse Column) vectors can be used, since the \emph{value} data for non-zeros does not need storage. Hence only a single 1-array of \texttt{Int64} row positions (with total length equal to the number of non-zeros in the block, with $8\times\mathtt{nnz}$ kB storage) is needed;

\item prepare subsequences of such data vectors (non-zero linear row indices), in order to feed efficiently the available processor threads.
In case of presence of one/more GPUs, a smaller size of the block---and hence of the boundary matrix and the encoded 3-chain vectors---and then much higher vector numbers, are preferable for speed.

\end{enumerate}

\paragraph{\emph{SpMM Multiplication}}
According to the current literature~\cite{} it is more convenient to execute SpMV (sparse matrix-vector) multiplications than SpMSpV (sparse matrix-sparse vector) multiplications. Since we have 256 such jobs (one multiplication per block) to perform in the standard setting of the algorithm\footnote{
Size of block $64^3$; size of image $512^2\times 256$.
}, or more in case of either smaller blocks or image greater than the standard one, this stage must be evidently parallelized and carefully tuned, possibly by using the GPU, if available.
\begin{enumerate}

\item Various multiplication algorithms are being experimented, using several packages for sparse linear algebra and/or custom implementations;

\item the total speed of this stage will strongly depend on the hardware available, on the granularity of blocks, and on the choice between dense/sparse storage of encoded 3-chains;

\item anyway, the compute elements or threads will be feed without solution of continuity in a \emph{dataflow} process. This parallel operation is, according to our preliminary experiments, the critical one of the whole workflow, in the sense that any $\Delta T$ (either positive or negative) in this stage will contribute to the total time $T$.

\end{enumerate}

\paragraph{\emph{2-Chain decoding}}
Each multiplication of $[\partial_B] : C_3 \to C_2$, times a 3-chain $\nu\in C_3$, produces a 2-chain  $\sigma\in C_2$, i.e.~the \emph{coordinate representation} of the \emph{boundary vector} $\sigma\in C_2$.  The inverse of the coding algorithm is executed in the present stage.  This process can also be partially superimposed in time with the previous ones, depending on the size of the memory buffers used to feed the CPU cores or the GPUs and get their results. Some elementary steps follow:

\begin{enumerate}

\item conversion from position of ones (or non-zeros) in the 2-chain to linear indices of rows;

\item conversion from linear indices to Cartesian indices in coordinates local to the $\mathcal{B}$ block, using the appropriate library functions;

\item conversion from each Cartesian index value to a suitably oriented (i.e.~with proper attitude) geometry quadrilateral (or pair of triangles) in local coordinates.

\end{enumerate}

A Julia's vectorized pipeline dataflow seems the more appropriate implementation model for the job of each worker.

\paragraph{\emph{Assembling and artifact filtering}}
The results of the previous stages can be described as a \emph{collection of sets} of \emph{geometric quadrilaterals (quads)}, each one encoded as an array of quadruples of integer indices, pointing to the linear array of grid vertices associated to the image block $\mathcal{B}$.  In other words, \emph{all quads} of \textbf{each job} are now given in the \textbf{same} \emph{local coordinates}.  Besides to put each partial surface $\mathcal{S}(B) = (\texttt{V}_B, \texttt{FV}_\sigma)$ in the global coordinate system of the image, the present stage must eliminate the redundant boundary features possibly generated at the edges of the partial surface $\mathcal{S}(B)$ within each block $\mathcal{B}$ such that $B \cap \mathcal{I} \not= \emptyset$:

\begin{enumerate}

\item translate each array $\mathtt{FV}_\sigma$, of type \texttt{Lar.Cells}, by summing to each vertex index the linearized offset of the Cartesian coordinates $(n,m,p)(B)$ of the $\mathcal{B}$'s \emph{reference vertex}, i.e.~the one with (all) lowest \emph{Cartesian coordinates} within the $\mathcal{B}$ block.

\item remove both instances of \emph{double quads} generated by \texttt{Lar} software at the block boundaries (see Figure~\ref{fig:blockboundary}). They are artifacts generated by the decomposition of the whole image into a number of blocks of tractable size.

\item 
a smart strategy of removal of such artifacts may be used, which does not require any sorting nor searching on the assembled array of quads. It will consist in arranging each block with all three dimensions decreased-increased by one, so that each 2-adjacent pair of blocks will be covering each other for a full side extent of blocks of depth one. The details of this \emph{artifact filtering} are elucidated in Section~\ref{sec:covering}.

\end{enumerate}

\paragraph{\emph{Smoothing}}
The final smoothing of the generated surfaces cannot be performed block-wise, since this would introduce smoothing artifacts at the block boundaries. Anyway, the Taubin smoothing~\cite{} can be performed in parallel, since for each vertex in the final surface (except eventually the ones on the image $\mathcal{I}$ boundaries) it essentially consist in computing a new position as a proper average of its neighborhood vertices, i.e.~by applying a discrete Laplacian operator.  Some appropriate set workers may so be assigne the task of generating iteratively a new position for the vertices they take cure of. In particular, we have:
\begin{enumerate}

\item Job enqueuing, by writing sets of integers (global linear indices of vertices) in array buffers of appropriate type \texttt{Channel};

\item iterated vectorized computation of proper averages of closed vertices;

\item job dequeuing, by recovering finished tasks from a channel and assembling the results into the embedding function $\mathtt{V}: C_0 \to \E^3$ providing an array of type \texttt{Lar.Points} of \texttt{Float64 $\times$ 3}, with vertex coordinates by column.
\end{enumerate}

\subsection{Performance analysis}\label{sec:analysis}

The size of boundary matrix is critical parameter of LAR-SURF method. To determine optimal size of boundary matrix the experiment on artifical data was performed (Fig.  \ref{fig:bm_size_tesla}). Size of experimental data is set to $512\times512\times512$ and it is derived from typical size of Computed Tomography medical images. Computation is done on Tesla DGX-1 machine.

\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{figs/bm_size_tesla.pdf} 
%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
\caption{Time requirements of LAR-SURF filter used on artificial with different size of boundary matrix}
\label{fig:bm_size_tesla}
\end{figure}

To compare time requirements of LAR-SURF with Marching Cubes implemented in Python we performed experiment on Ircadb dataset \cite{ircad}. Dataset contain 20 Computed Tomography images (see table \ref{tab:ircad1}) with xy-resolution from 0.56 mm to 0.87 mm and z-resolution from 1.0 mm to 4.0 mm. The number of slices is each series varies from 74 to 260 and size of each slice is $512\times512$. Dataset contain manually segmented liver, portal vein and other structures. We performed surface extraction of liver with Marching Cubes and LAR-SURF. Time required for computation can be seen on figure \ref{fig:ircad_comparison}.

Based on t-test with $\alpha=0.99$, $p=\input{input/ircad_pvalue}$ and 
$s=\input{input/ircad_statistics}$ it can be shown that the mean of time consumed by LAR-SURF is significantly lower from time consumed by Marching Cubes.

\begin{figure}
\centering
\includegraphics[width=0.99\textwidth]{figs/ircad_comparison.pdf} 
%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
\caption{Time requirements of LAR-SURF filter and Marching Cubes on Ircad dataset. Error bars shows the 95\% confidence interval}
\label{fig:ircad_comparison}
\end{figure}

%\input{examples}
\section{Examples}\label{sec:examples}

\begin{figure}
\centering
\includegraphics[width=0.4\textwidth]{figs/porta_smoothing_1.png} 
\vspace{0.05\textwidth}
\includegraphics[width=0.4\textwidth]{figs/porta_smoothing_2.png} 
%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
\caption{Triangulated isosurface of portal vein calculated with LAR-SURF}
\label{fig:example_porta}
\end{figure}

%\input{conclusion}
\section{Conclusion}\label{sec:conclusion}

In this paper we have introduced and discussed a Julia implementation of an algebraic filter to extract the boundary sourface of a specific image segment, described as a 3-chain of voxels, from medical 3D images. We have shown a good advantage over standard marchin-cubes algorithms. Translations from cartesian indices of cells to linearized indices, and the sparse matrix-vector multiplication are the main computational kernels of this approach. The current implementation employs Julia's channels for multiprocessing, and can be extended to gain a much greater speed-up using hybrid architectures mixing  CPUs and GPUs of last generation.

\appendix
\section{Appendix}
\subsection{Symbol list}

\begin{description}
\item[$\mathcal{I}$]  three-dimensional medical image
\item[$\ell_1, \ell_2, \ell_3$]  dimensions of image
\item[$\mathcal{S}$]  segment: a subset of voxels from image segmentation
\item[$\mathcal{B}$] 3D image block
\item[$size$] lateral dimension of cubic block $\mathcal{B}$
\item[$n$]	number of blocks $\mathcal{B}$ (jobs) in $\mathcal{I}$
\item[[$\partial_\mathcal{B}$] ]  boundary matrix for block $\mathcal{B}$
\item[$C_p$] linear (vector) space of $p$-chains
\item[$\nu\in C_p$] $p$-chain
\item[[$\nu$]] coordinate representation (binary vector) of  $\nu$
\item[$\E^3$] Euclidean 3-space
\end{description}

\subsection{Definitions}
\begin{description}

\item[Boundary model] Closed manifold surface of the boundary of a solid model
\item[CSC]  Compressed Sparse Column format for sparse matrices
\item[Global coordinates]  Integer linear coordinates of $\mathcal{I}$
\item[Local coordinates] Integer linear coordinates of $\mathcal{B}$
\item[Cartesian coordinates] Integer triples $(i,j,k)$ one-to-one with voxels
\item[Voxels] Individual elements in 3D image (3-cells)
\item[$p$-chain] Formal linear combination of $p$-cells with coefficients in $\{0,1\}$
\item[Coord. repr.]  Binary vector (for $p$-chains) or binary matrix (for chain operators)
\item[Quad]	Geometric quadrilateral; convex polygon with four vertices
\item[Foreground voxel] Individual element of a segment $\mathcal{S}$
\item[Segment]	Subset of voxels resulting from image segmentation
\end{description}

\subsection{3D-Ircadb Dataset}

\begin{table}
\input{input/dataset_ircad}
\label{tab:ircad1}
\end{table}

\begin{table}
\input{input/dataset_ircad_desc}
\label{tab:ircad2}
\end{table}

\clearpage{}
\bibliographystyle{alpha}
\bibliography{library,TSAS-2019}

\end{document}

### 2019-11-10 src/paper.tex
paper revised

@@
 
-\paragraph{\emph{Chain coordinates }\label{sec:bbbb}}
+\subsubsection{Chain coordinates }\label{sec:bbbb}
 We are going to treat each image block independently from each other.
@@
 
-\paragraph{\emph{Boundary operator }\label{sec:bbbb}}For a fixed block size $n$, the boundary operator $\partial_d : C_d\to C_{d-1}$, with $d\in\{2,3\}$, will be constructed once and for all using the algorithm given in~\cite{}, and inlined in the generated boundary extraction code.
+\subsubsection{Boundary operator }\label{sec:bbbb}
+For a fixed block size $n$, the boundary operator $\partial_d : C_d\to C_{d-1}$, with $d\in\{2,3\}$, will be constructed once and for all using the algorithm given in~\cite{}, and inlined in the generated boundary extraction code.
 
@@
 
-\paragraph{\emph{Sparsity and size of boundary matrix }\label{sec:bbbb}}
+\subsubsection{Sparsity and size of boundary matrix }\label{sec:bbbb}
 
@@
 
-\paragraph{Embedding}
+\subsubsection{Embedding}
 A final computational step is needed, in order to embed the 2-chains in Euclidean space $\E^3$ and to assemble the whole resulting surface.
@@
 
-\paragraph{Surface assembling}
+\subsubsection{Surface assembling}
 
@@
 
-\paragraph{\emph{Workflow setup}}The functions in this preliminary step include:
+\subsubsection{Workflow setup}\label{}The functions in this preliminary step include:
 \begin{enumerate}
@@
 
-\paragraph{\emph{Job enqueuing}}Communication and data synchronization may be managed through \emph{Channels}, which are the FIFO conduits that may provide producer/consumer communication.
+\subsubsection{Job enqueuing}\label{}Communication and data synchronization may be managed through \emph{Channels}, which are the FIFO conduits that may provide producer/consumer communication.
 Overall execution time can be improved if other tasks can be run while a task is being executed, or while waiting for an external service/function to complete.
@@
 
-\paragraph{\emph{3-Chain encoding}}
+\subsubsection{3-Chain encoding}\label{}
 The interesting part of the \emph{Image} $\mathcal{I}$ is called \emph{Segment} $\mathcal{S}$.
@@
 
-\paragraph{\emph{SpMM Multiplication}}
+\subsubsection{SpMM Multiplication}\label{}
 According to the current literature~\cite{} it is more convenient to execute SpMV (sparse matrix-vector) multiplications than SpMSpV (sparse matrix-sparse vector) multiplications.
@@
 
-\paragraph{\emph{2-Chain decoding}}
+\subsubsection{2-Chain decoding}\label{}
 Each multiplication of $[\partial_B] : C_3 \to C_2$, times a 3-chain $\nu\in C_3$, produces a 2-chain  $\sigma\in C_2$, i.e.~the \emph{coordinate representation} of the \emph{boundary vector} $\sigma\in C_2$.
@@
 
-\paragraph{\emph{Assembling and artifact filtering}}
+\subsubsection{Assembling and artifact filtering}\label{}
 The results of the previous stages can be described as a \emph{collection of sets} of \emph{geometric quadrilaterals (quads)}, each one encoded as an array of quadruples of integer indices, pointing to the linear array of grid vertices associated to the image block $\mathcal{B}$.
@@
 
-\paragraph{\emph{Smoothing}}
+\subsubsection{Smoothing}\label{}
 The final smoothing of the generated surfaces cannot be performed block-wise, since this would introduce smoothing artifacts at the block boundaries.

### 2019-11-11 src/paper.tex
paper.tex partially reviewed

--- 2019-11-10 src/paper.tex
@@
 \date{}							% Activate to display a given date or no date
+
+\usepackage{draftwatermark}
+\SetWatermarkText{Draft}
+\SetWatermarkScale{2.0}
+\SetWatermarkColor[rgb]{0.9,0.9,0.9}
 
@@
 
-Some basic concepts of solid modeling, and in particular the foundational idea of representation scheme, as well as basic concepts of algebraic topology, are shortly introduced in this Section, including the computation of the matrix of a boundary operator between chain spaces.
+Some basic concepts of solid modeling, and in particular the foundational idea of representation scheme, as well as few basic concepts of algebraic topology, are shortly introduced in this Section, including the computation of the matrix of a boundary operator between chain spaces.
 
@@
 \qquad\equiv\qquad
-\texttt{(geometry, topology) = (W, (EV, FE, CF))}
+\texttt{(geom, top) = (W, (EV, FE, CF))}
 \]
 The GCC allows to transform the (possibly non connected) boundary 2-cycle of surfaces as a standard B-rep~\cite{shapiroSM:202}. 
-The \texttt{geometry} is given by
-the embedding matrix \texttt{W} of vertices (0-cells), and \texttt{topology} is
-given by the three sparse matrices \texttt{CF}, \texttt{FE}, \texttt{EV} of coboundaries
-{i.e.},~$\delta_3, \delta_2, \delta_1$, of  chain complex describing a   
+The geometry \texttt{geom} is given by
+the embedding matrix \texttt{W} of vertices (0-cells), and  topology \texttt{top} by the three sparse matrices (\texttt{EV}, \texttt{FE}, \texttt{CF}) of coboundaries ($\delta_1, \delta_2, \delta_3$) of the chain complex describing a
 space arrangement~\cite{paoluzzi2019finite}.
@@
 
-\subsubsection{Construction of boundary matrix $\partial_d$}
+\subsubsection*{Construction of boundary matrix $\partial_d$}
 
@@
 Note that the number of rows equates the cardinality $4\times 4\times 4 = 64$ of the voxel set; the number of columns is $d\,n\,(1+n)^{d-1} = 3\times 4\times 25 = 300$.
-Of course, the number of non-zeros per row (cardinality of single voxel facets) is six, whereas the number of non-zeros per column is two, but on boundary facets.}
+Of course, the number of non-zeros per row (cardinality of the facet set of a single voxel) is six, whereas the number of non-zeros per column is two, but on boundary facets.}
    \label{fig:boundary_matrix_4x4x4}
@@
 
-Every boundary chain extracted from an image block $\B(i,j,n)$ is a \emph{2-cycle}, i.e., a closed 2-chain---in other words, a 2-chain with empty boundary.
-Such 2-cycles are joined together by removing the double 2-cells (at the boundaries of bricks) after having suitably shifted their indices to an unique linear representation of the whole image.
+Every boundary chain extracted from an image block $\B(i,j,k,n)$ is a \emph{2-cycle}, i.e., a closed 2-chain---in other words, a 2-chain with empty boundary.
+Such 2-cycles are joined together by removing the double 2-cells (at the boundaries of adjacent bricks) after having suitably shifted their indices to an unique linear representation of the whole image.
 The resulting raster surface is made by mutually orthogonal raster facets, that must be smoothed in order to get a fair surface.
@@
 A well known weakness of this simple algorithm is the asymptotic convergence of the whole mesh to a single point, resulting in unfair size reduction even after few iterations.
-The Taubin smoothing algorithm~\cite{Taubin:1995:SPA:218380.218473,egst.20001029} alternates two Laplacian smoothing steps with shrink and inflate effects respectively, with the result of letting pretty invariant sizes and volume of the smoothed mesh.
+Conversely, the Taubin smoothing algorithm~\cite{Taubin:1995:SPA:218380.218473,egst.20001029} alternates two Laplacian smoothing steps with \emph{shrink} and \emph{inflate} effects respectively, with the result of delivering pretty invariant sizes and volume of the smoothed mesh.
 The best results are obtained on meshes which have small variations of edge length and face angles, like for surfaces extracted from 3D raster images, as in our case.
@@
 We assume that medical devices produce 3D images with lateral dimensions that are integer multiples of some powers of two, like 128, 256, 512, etc.
-Any cuboidal portion of image is completely determined by the Cartesian indices of its voxels of lowest and highest indices, and extracted by multidimensional array \emph{slicing} as $Image([\ell_x$:$h_x, \ell_y$:$h_y, \ell_z$:$h_z])$.
+Any cuboidal portion of image is completely determined by the Cartesian indices of its voxels of lowest and highest indices, and extracted by multidimensional array \emph{slicing} as $Image([\ell_x : h_x, \ell_y : h_y, \ell_z : h_z])$.
 
@@
 
-\paragraph{\emph{Chain coordinates }\label{sec:bbbb}}
+\subsubsection*{Chain coordinates }\label{sec:bbbb}
 We are going to treat each image block independently from each other.
-Hence we map each image block $\B(i,j,n)$ to the linear \emph{chain} space $C_2$ of dimension $n\times n$, using coordinate vectors $c_{h,k}\in \B^{n\times n} := \{0,1\}^{n\times n}$, where the basis element $c = c_{h,k} \in C_2$ is mapped via Cartesian-to-linear map to the binary vector
+Hence we map each image block $\B(i,j,k,n)$ to the linear \emph{chain} space $C_2$ of dimension $n\times n\times n$, using coordinate vectors $c\in \mathbf{2}^{n^d} := \{0,1\}^{n^d}$, where the basis element $c \in C_2$ is mapped via Cartesian-to-linear map to the binary vector
 \[
@@
 
-\paragraph{\emph{Boundary operator }\label{sec:bbbb}}For a fixed block size $n$, the boundary operator $\partial_d : C_d\to C_{d-1}$, with $d\in\{2,3\}$, will be constructed once and for all using the algorithm given in~\cite{}, and inlined in the generated boundary extraction code.
+\subsubsection*{Boundary operator }\label{sec:bbbb}
+For a fixed block size $n$, the boundary operator $\partial_d : C_d\to C_{d-1}$, with $d\in\{2,3\}$, will be constructed once and for all using the algorithm given in~\cite{}, and inlined in the generated boundary extraction code.
 
@@
 
-\paragraph{\emph{Sparsity and size of boundary matrix }\label{sec:bbbb}}
+\subsubsection*{Sparsity and size of boundary matrix }\label{sec:bbbb}
 
@@
 Here we refer directly to the 3D case.
-Let us call \emph{segment} the bulk content $S$ of interest within the input 3D image of size $(u,v,w)$. We aim to compute the segment boundary $\partial_3 S$. 
-First we set the size $n$ of the block, in order to decompose the input $Image(u,v,w)$ into a fair number 
+Let us call \emph{segment} the bulk content $S$ of interest within the input 3D image of size $(u,v,w)$. Our aim is to compute the segment boundary $\partial_3 S$. 
+First we set the size $n$ of the block, in order to decompose the input $Image(u,v,w)$ into a fair number of blocks
 \[
-M = \ceil{u/n} \times \ceil{v/n} \times \ceil{w/n} \simeq \frac{uvw}{n^3}
-\] of blocks. 
-Then we consider each image portion $c_{i,j,k} = S\cap \B(i,j,k,n)$ and compute its (binary) coordinate representation  $[c]_{i,j,k}\in C_3(n,n,n)$.
+M = \ceil{u/n} \times \ceil{v/n} \times \ceil{w/n} \simeq \frac{uvw}{n^3}.
+\] 
+Then, we consider each segment portion $c_{i,j,k} = S\cap \B(i,j,k,n)$ and compute its local coordinate representation  $[c]_{i,j,k}\in C_3(n,n,n)$.
 This one is a sparse binary vector of length $n^3$.
-Then assemble the $M$ representations $c$ of segment portions into a sparse binary matrix $\T{S}$, of dimension $n^d \times M$.
-Finally compute a matrix $\T{\B}$ of boundary portions of $S$, represented by columns as chain coordinate vectors in $C_2$:
+Then, assemble the $M$ representations $c$ of segment portions into a sparse binary matrix $\T{S}$, of dimension $n^d \times M$, with $d=3$.
+Finally, compute a matrix $\T{\textbf{B}}$ of boundary portions of $S$, represented by columns as chain coordinate vectors in $C_2$:
 \[
-\T{\B} = [\partial_3(n)]\, \T{S}.
+\T{\textbf{B}} = [\partial_3(n)]\, \T{S}.
 \]
-where the boundary matrix has dimension $d\,n\,(1+n)^{d-1} \times n^d$.
-Of course, the $\T{\B}$ sparse matrix has the same column number $M$ of $\T{S}$, because each column contains the boundary representation of the corresponding $S\cap Box_{i,j,k}{i,j,k}$, and the number of rows of the operator, equal to the dimension of the linear space $C_2$.
+where the boundary matrix has dimension $n^d \times dn(n+1)^{d-1}$.
+Of course, the $\T{\textbf{B}}$ sparse matrix has the same column number $M$ of $\T{S}$, because each column contains the boundary representation of the corresponding $S \cap \B(i,j,k,n)$, and the number of matrix rows is the dimension $n^d$ of the linear space $C_2$ of 2-chains.
 
-\paragraph{Embedding}
-A final computational step is needed, in order to embed the 2-chains in Euclidean space $\E^3$ and to assemble the whole resulting surface.
-In particular, we need to compute the \emph{embedding function} $\mu : C_0 \to \E^3$, where $C_0$ is the space of 0-chains, one-to-one corresponding to the vertices of the extracted surface.
-The simplest solution is to associate  four 0-cells to each 2-cell of the extracted surface, i.e.~to each non-zero entry in every column of $\T{\B}$.
+\subsubsection*{Embedding}
+A final computational step is needed, in order to embed the 2-chains in $\E^3$ space and to assemble the whole resulting surface.
+In particular, we need to compute the \emph{embedding function} $\mu : C_0 \to \E^3$, where $C_0$ is the space of 0-chains, one-to-one with the vertices of the extracted surface.
+The simplest solution is to associate  four 0-cells to each 2-cell of the extracted surface, i.e.~to each non-zero entry in every column of $\T{\textbf{B}}$.
 The $\mu$ function  can be computed by identifying, via  element position in the column, a triple of integer values $0\leq x\leq u$, $0\leq y\leq v$, and $0\leq z\leq w$ for each vertex of the 2-cell.
@@
 
-\paragraph{Surface assembling}
+\subsubsection*{Surface assembling}
 
-All boundary surface subsets $\B_{i,j,k}(S) = \partial_3 S \cap \mbox{Box}_{i,j,k}$, provided by  columns of $\T{S}$, are embedded in the same coordinate space. In formal terms: 
+All boundary surface subsets $B(i,j,k) = \partial_3 S \cap \B(i,j,k)$, provided by  columns of $\T{B}$, are embedded in the same coordinate space.
+In formal terms, using the standard terminology of LAR scheme:
 \[
@@
 \]
-where, with respect to the \emph{chain complex} $C_3\to C_2\to C_1\to C_0$ induced by the input image $Im$ and segment $S_{i,j,k}$, we get
+where, with respect to the \emph{chain complex} $C_3\to C_2\to C_1\to C_0$ induced by the input image $Im$ and segment portion $S_{i,j,k}$, we get
 \begin{align}
-\texttt{Geom} &:= \mu(C_0(_{i,j,k})) = \texttt{V},
+\texttt{Geom} &:= \mu(C_0) = \texttt{V},
 \\
@@
 \begin{align}
-\texttt{Lar}(B_{i,j,k}) &:= (\texttt{Geom}, \texttt{Top}) = (\texttt{W}, \texttt{FW}),
+\texttt{Lar}(B_{i,j,k}) &:= (\texttt{Geom}(B_{i,j,k}), \texttt{Top}(B_{i,j,k})) = (\texttt{W}, \texttt{FW}),
 \\
@@
 \\
-\texttt{Top} &:= C_2(B_{i,j,k}) = \T{\B}_{i,j,k} \mapsto \texttt{FW} \subset \texttt{FV}.
+\texttt{Top} &:= C_2(B_{i,j,k}) = \T{\textbf{B}}_{i,j,k} \mapsto \texttt{FW} \subset \texttt{FV}.
 \end{align}
@@
 \[
-\texttt{Lar}(\B) = \oplus_{i,j,k}\texttt{Lar}(\partial_3 S_{i,j,k})) = \oplus_{i,j,k}(\texttt{W}, \texttt{FW}) .
+\texttt{Lar}(\textbf{B}) = \oplus_{i,j,k}\texttt{Lar}(\partial_3 S_{i,j,k})) = \oplus_{i,j,k}(\texttt{W}, \texttt{FW}) .
 \]
@@
 In particular, little effort is needed to separate the problem into a number of parallel tasks $S_{i,j,k}$, using multiarray slicing.
-The granularity of parallelism, depending on the block size $n$, is further enforced by the computation of a single boundary matrix $[\partial_d(n)]$, in turn depending on $n$, so that the initial communication cost of broadcasting the matrix to nodes can be carefully controlled, and finely tuned depending on the system architecture.
+The granularity of parallelism, depending on the block size $n$, is further enforced by the computation of a single boundary matrix $[\partial_d(n)]$ depending on $n$, so that the initial communication cost of broadcasting the matrix to nodes can be carefully controlled, and finely tuned depending on the system architecture.
 The whole approach is appropriate  for SIMD hybrid architectures of CPUs and GPUs, since only the initial block setup of boundary matrix and image slices, as well the final collection of computed surface portions, require inter-process communication.
@@
 
-\paragraph{\emph{Workflow setup}}The functions in this preliminary step include:
+\subsubsection*{Workflow setup}\label{}The functions in this preliminary step include:
 \begin{enumerate}
@@
 
-\item depending on the above, best decision for $\emph{size}$ of 3D image blocks $\mathcal{B}$ (bricks); defaults to: $\emph{size}=64$; hence the number of bricks will be $n=\ceil{\ell_1/size}\times\ceil{\ell_2/size}\times\ceil{\ell_3/size}$.
-Hence default value: $n=256$, for images $\mathcal{I} = 512\times 512\times 256$;
+\item depending on the above, best decision for $\emph{size}$ of 3D image blocks $\mathcal{B}$ (bricks).
+With defaults $\emph{size}=64$,  the number of bricks will be $n=\ceil{\ell_1/size}\times\ceil{\ell_2/size}\times\ceil{\ell_3/size}$.
+Hence the default number of bricks is $n=256$, for standard images $512\times 512\times 256$;
 
@@
 
-\item distribution of matrix $[\partial_B]$, of default size $45$ MB, to all available nodes/cores (workers) using the macro \texttt{@eveywhere}.
+\item distribution of matrix $[\partial_B]$, of default size $45$ MB, to all available nodes/cores (workers) using the Julia macro \texttt{@eveywhere}.
 
@@
 %\includegraphics[scale=1]{input/ircad_comparison.pdf} 
-\caption{Workflow of LAR-SURF algorithm}
+\caption{Workflow of \textsc{lar-surf} algorithm}
 \label{fig:schema}
@@
 
-With $size=64$, the number of non-zeros within the sparse matrix $[\partial_{64^3}]$ is $\mathtt{nnz} = 4 792 266$, for a memory size of $9\times \mathtt{nnz}+8\times 262144 \simeq 45$ MB.
+With $size=64$, the number of non-zeros within the sparse matrix $[\partial(64^3)]$ is $\mathtt{nnz} = 4 792 266$, for a memory size of $9\times \mathtt{nnz}+8\times 262144 \simeq 45$ MB.
 The memory size of the sparse matrix is computed by considering $8+1$ bytes for non-zero element (which are exactly 6 per row), plus 8 bytes per each index of column start.
 
-\paragraph{\emph{Job enqueuing}}Communication and data synchronization may be managed through \emph{Channels}, which are the FIFO conduits that may provide producer/consumer communication.
+\subsubsection*{Job enqueuing}\label{}Communication and data synchronization may be managed through \emph{Channels}, which are the FIFO conduits that may provide producer/consumer communication.
 Overall execution time can be improved if other tasks can be run while a task is being executed, or while waiting for an external service/function to complete.
@@
 
-\paragraph{\emph{3-Chain encoding}}
+\subsubsection*{3-Chain encoding}\label{}
 The interesting part of the \emph{Image} $\mathcal{I}$ is called \emph{Segment} $\mathcal{S}$.
@@
 
-\paragraph{\emph{SpMM Multiplication}}
+\subsubsection*{SpMM Multiplication}\label{}
 According to the current literature~\cite{} it is more convenient to execute SpMV (sparse matrix-vector) multiplications than SpMSpV (sparse matrix-sparse vector) multiplications.
@@
 
-\paragraph{\emph{2-Chain decoding}}
+\subsubsection*{2-Chain decoding}\label{}
 Each multiplication of $[\partial_B] : C_3 \to C_2$, times a 3-chain $\nu\in C_3$, produces a 2-chain  $\sigma\in C_2$, i.e.~the \emph{coordinate representation} of the \emph{boundary vector} $\sigma\in C_2$.
@@
 
-\paragraph{\emph{Assembling and artifact filtering}}
+\subsubsection*{Assembling and artifact filtering}\label{}
 The results of the previous stages can be described as a \emph{collection of sets} of \emph{geometric quadrilaterals (quads)}, each one encoded as an array of quadruples of integer indices, pointing to the linear array of grid vertices associated to the image block $\mathcal{B}$.
@@
 
-\paragraph{\emph{Smoothing}}
+\subsubsection*{Smoothing}\label{}
 The final smoothing of the generated surfaces cannot be performed block-wise, since this would introduce smoothing artifacts at the block boundaries.
 Anyway, the Taubin smoothing~\cite{} can be performed in parallel, since for each vertex in the final surface (except eventually the ones on the image $\mathcal{I}$ boundaries) it essentially consist in computing a new position as a proper average of its neighborhood vertices, i.e.~by applying a discrete Laplacian operator.
-Some appropriate set workers may so be assigne the task of generating iteratively a new position for the vertices they take cure of.
+Some appropriate set of workers may so be assigned the task of generating iteratively a new position for the vertices they take cure of.
 In particular, we have:
@@
 
-\item Job enqueuing, by writing sets of integers (global linear indices of vertices) in array buffers of appropriate type \texttt{Channel};
+\item Job enqueuing, by writing sets of integers (global linear indices of vertices) in array buffers of type \texttt{Channel};
 
-\item iterated vectorized computation of proper averages of closed vertices;
+\item vectorized computation of proper averages of near vertices;
 
-\item job dequeuing, by recovering finished tasks from a channel and assembling the results into the embedding function $\mathtt{V}: C_0 \to \E^3$ providing an array of type \texttt{Lar.Points} of \texttt{Float64 $\times$ 3}, with vertex coordinates by column.
+\item job dequeuing, by recovering finished tasks from a channel and assembling the results into the embedding function $\mathtt{V}: C_0 \to \E^3$, providing an array of type \texttt{Lar.Points} of \texttt{Float64 $\times$ 3}, with vertex coordinates by column.
 \end{enumerate}

### 2019-12-02 src/paper.tex

@@
 The geometry \texttt{geom} is given by
-the embedding matrix \texttt{W} of vertices (0-cells), and  topology \texttt{top} by the three sparse matrices (\texttt{EV}, \texttt{FE}, \texttt{CF}) of coboundaries ($\delta_1, \delta_2, \delta_3$) of the chain complex describing a
+the embedding matrix \texttt{W} of vertices (0-cells), and  topology \texttt{top} by the three sparse matrices (\texttt{EV}, \texttt{FE}, \texttt{CF}) of coboundaries ($\delta_0, \delta_1, \delta_2$) of the chain complex describing a
 space arrangement~\cite{paoluzzi2019finite}.
@@
 
-\item input of 3D medical image $\mathcal{I}$ dimensions $\ell_1, \ell_2, \ell_3$, such that: $\mathcal{I} = [\ell_1]\times[\ell_2]\times[\ell_3]$, where $[\ell_k] = (1,2,\ldots,\ell_k)$;
+\item input of 3D medical image $\mathcal{I}$ with \emph{shape} $(\ell_1, \ell_2, \ell_3)$, such that: $\mathcal{I} = [\ell_1]\times[\ell_2]\times[\ell_3]$, where $[\ell_k] = [1,2,\ldots,\ell_k]$;
 
@@
 
-\item depending on the above, best decision for $\emph{size}$ of 3D image blocks $\mathcal{B}$ (bricks).
-With defaults $\emph{size}=64$,  the number of bricks will be $n=\ceil{\ell_1/size}\times\ceil{\ell_2/size}\times\ceil{\ell_3/size}$.
+\item depending on the above, best decision for \emph{size} of 3D image block (or brick) $\mathcal{B}$.
+With defaults $\emph{size}=64$,  the \emph{number of bricks} will be $n=\ceil{\ell_1/size}\times\ceil{\ell_2/size}\times\ceil{\ell_3/size}$.
 Hence the default number of bricks is $n=256$, for standard images $512\times 512\times 256$;
 
-\item computation of a Julia's sparse boundary matrix $[\partial_B]$, returning a value of type \texttt{SparseMatrixCSC\{Int8\}\{Int64\}}, where \texttt{Int8} and \texttt{Int64} are the types for values and indices, respectively, stored by Compressed Sparse Column (CSC) format; the average $[\partial_B]$ (for $\emph{size}$ = 64) is about 45 MB;
+\item computation of the Julia's sparse boundary matrix $[\partial_B]$, returning a value of type \texttt{SparseMatrixCSC\{Int8\}\{Int64\}}, where \texttt{Int8} and \texttt{Int64} are the types for values and indices, respectively, stored by Compressed Sparse Column (CSC) format; the storage of $[\partial_B]$ (for $\emph{size}$ = 64) requires about 45 MB;
 
-\item creation of either a local or distributed channel to implement a producer/consumer model of parallel/distributed computation, depending on available resources;
+\item creation of either a local or distributed \texttt{channel} to implement a producer/consumer model of parallel/distributed computation, depending on available resources;
 
-\item distribution of matrix $[\partial_B]$, of default size $45$ MB, to all available nodes/cores (workers) using the Julia macro \texttt{@eveywhere}.
+\item distribution of matrix $[\partial_B]$, of default size $45$ MB, to all available nodes/cores (Julia workers) using the Julia macro \texttt{@eveywhere}.
 
@@
 
-\subsubsection*{Job enqueuing}\label{}Communication and data synchronization may be managed through \emph{Channels}, which are the FIFO conduits that may provide producer/consumer communication.
+\subsubsection*{Job enqueuing}\label{}
+Communication and data synchronization may be managed through \emph{Channels}, which are the FIFO conduits that may provide producer/consumer communication.
 Overall execution time can be improved if other tasks can be run while a task is being executed, or while waiting for an external service/function to complete.
@@
 
-\item Various multiplication algorithms are being experimented, using several packages for sparse linear algebra and/or custom implementations;
+\item Various multiplication algorithms are being tested, using several packages for sparse linear algebra and/or custom implementations;
 
@@
 
-\item anyway, the compute elements or threads will be feed without solution of continuity in a \emph{dataflow} process.
+\item anyway, the compute elements or threads will be fed without solution of continuity in a \emph{dataflow} process.
 This parallel operation is, according to our preliminary experiments, the critical one of the whole workflow, in the sense that any $\Delta T$ (either positive or negative) in this stage will contribute to the total time $T$.
@@
 
-In this paper we have introduced and discussed a Julia implementation of an algebraic filter to extract the boundary sourface of a specific image segment, described as a 3-chain of voxels, from medical 3D images.
+In this paper we have introduced and discussed a Julia implementation of an algebraic filter to extract from medical 3D images the boundary sourface of a specific image segment, described as a 3-chain of voxels.
 We have shown a good advantage over standard marchin-cubes algorithms.

### 2019-12-20 src/implementation.tex

--- 2019-06-30 src/implementation.tex
@@
 With $size=64$, the number of non-zeros within the sparse matrix $[\partial_{64^3}]$ is $\mathtt{nnz} = 4 792 266$, for a storage size of $9\times \mathtt{nnz}+8\times 262144 \simeq 45$ MB.
+
+\begin{figure}[htbp]
+\includegraphics[width=0.99\textwidth]{figs/schema_horizontal.pdf} 
+%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
+\caption{Workflow of LAR-SURF algorithm}
+\label{fig:schema}
+\end{figure}
+
+With $size=64$, the number of non-zeros within the sparse matrix $[\partial_{64^3}]$ is $\mathtt{nnz} = 4 792 266$, for a memory size of $9\times \mathtt{nnz}+8\times 262144 \simeq 45$ MB.
+The memory size of the sparse matrix is computed by considering $8+1$ bytes for non-zero element (which are exactly 6 per row), plus 8 bytes per each index of column start.
 
@@
 \subsection{Performance analysis}\label{sec:analysis}
+
+The size of boundary matrix is critical parameter of LAR-SURF method.
+To determine optimal size of boundary matrix the experiment on artifical data was performed (Fig.
+\ref{fig:bm_size_tesla}).
+Size of experimental data is set to $512\times512\times512$ and it is derived from typical size of Computed Tomography medical images.
+Computation is done on Tesla DGX-1 machine.
+
+\begin{figure}
+\centering
+\includegraphics[width=0.6\textwidth]{figs/bm_size_tesla.pdf} 
+%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
+\caption{Time requirements of LAR-SURF filter used on artificial with different size of boundary matrix}
+\label{fig:bm_size_tesla}
+\end{figure}
+
+To compare time requirements of LAR-SURF with Marching Cubes implemented in Python we performed experiment on Ircadb dataset \cite{ircad}.
+Dataset contain 20 Computed Tomography images (see table \ref{tab:ircad1}) with xy-resolution from 0.56 mm to 0.87 mm and z-resolution from 1.0 mm to 4.0 mm.
+The number of slices is each series varies from 74 to 260 and size of each slice is $512\times512$.
+Dataset contain manually segmented liver, portal vein and other structures.
+We performed surface extraction of liver with Marching Cubes and LAR-SURF.
+Time required for computation can be seen on figure \ref{fig:ircad_comparison}.
+
+Based on t-test with $\alpha=0.99$, $p=\input{input/ircad_pvalue}$ and 
+$s=\input{input/ircad_statistics}$ it can be shown that the mean of time consumed by LAR-SURF is significantly lower from time consumed by Marching Cubes.
+
+\begin{figure}
+\centering
+\includegraphics[width=0.99\textwidth]{figs/ircad_comparison.pdf} 
+%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
+\caption{Time requirements of LAR-SURF filter and Marching Cubes on Ircad dataset. Error bars shows the 95\% confidence interval}
+\label{fig:ircad_comparison}
+\end{figure}

### 2019-12-30 src/paper.tex
new examples added

@@
 
+% M. Jirik added - code highlithing
+% \usepackage{minted}
+\usepackage[cache=false]{minted}
+% \usepackage{minted}
 %SetFonts
@@
 
-\subsection{Representation Schemes}\label{sec:aaaa}
+\subsection{Representation Schemes}\label{sec:schemes}
 
@@
 
-\subsection{Linear Algebraic Representation}\label{sec:aaaa}
+\subsection{Linear Algebraic Representation}\label{sec:lar}
 
@@
 
-\subsection{Multiindices from Cartesian indices}\label{sec:aaaa}
+\subsection{Multiindices from Cartesian indices}\label{sec:inds-from-cart}
 
@@
 
-\subsection{Taubin Smoothing}\label{sec:aaaa}
+\subsection{Taubin Smoothing}\label{sec:taubin}
 
@@
 
-\subsection{Block decomposition}\label{sec:bbbb}
+\subsection{Block decomposition}\label{sec:block-decomposition}
 
@@
 
-\subsubsection*{Chain coordinates }\label{sec:bbbb}
+\subsubsection*{Chain coordinates }\label{sec:chain-coords}
 We are going to treat each image block independently from each other.
@@
 
-\subsubsection*{Boundary operator }\label{sec:bbbb}
+\subsubsection*{Boundary operator }\label{sec:boundary-operator}
 For a fixed block size $n$, the boundary operator $\partial_d : C_d\to C_{d-1}$, with $d\in\{2,3\}$, will be constructed once and for all using the algorithm given in~\cite{}, and inlined in the generated boundary extraction code.
@@
 
-\subsubsection*{Sparsity and size of boundary matrix }\label{sec:bbbb}
+\subsubsection*{Sparsity and size of boundary matrix }\label{sec:bm-size}
 
@@
 
+%TODO fix the name
 \begin{figure}[htbp] %  figure placement: here, top, bottom, or page
@@
 
-\subsection{Block boundary mapping}\label{sec:bbbb}
+\subsection{Block boundary mapping}\label{sec:block-mapping}
 
@@
 
-\subsection{Block-level parallelism}\label{sec:bbbb}
+\subsection{Block-level parallelism}\label{sec:block-parallelism}
 
@@
 
-\subsubsection*{Workflow setup}\label{}The functions in this preliminary step include:
+\subsubsection*{Workflow setup}\label{sec:workflow}
+The functions in this preliminary step include:
 \begin{enumerate}
@@
 
-\subsubsection*{Job enqueuing}\label{}
+\subsubsection*{Job enqueuing}\label{sec:job-enq}
 Communication and data synchronization may be managed through \emph{Channels}, which are the FIFO conduits that may provide producer/consumer communication.
@@
 
-\subsubsection*{3-Chain encoding}\label{}
+\subsubsection*{3-Chain encoding}\label{chain-coding}
 The interesting part of the \emph{Image} $\mathcal{I}$ is called \emph{Segment} $\mathcal{S}$.
@@
 
-\subsubsection*{SpMM Multiplication}\label{}
+\subsubsection*{SpMM Multiplication}\label{SpMM-multiplication}
 According to the current literature~\cite{} it is more convenient to execute SpMV (sparse matrix-vector) multiplications than SpMSpV (sparse matrix-sparse vector) multiplications.
@@
 
-\subsubsection*{2-Chain decoding}\label{}
+\subsubsection*{2-Chain decoding}\label{sec:two-chain-decoding}
 Each multiplication of $[\partial_B] : C_3 \to C_2$, times a 3-chain $\nu\in C_3$, produces a 2-chain  $\sigma\in C_2$, i.e.~the \emph{coordinate representation} of the \emph{boundary vector} $\sigma\in C_2$.
@@
 
-\subsubsection*{Assembling and artifact filtering}\label{}
+\subsubsection*{Assembling and artifact filtering}\label{sec:artifact-filtering}
 The results of the previous stages can be described as a \emph{collection of sets} of \emph{geometric quadrilaterals (quads)}, each one encoded as an array of quadruples of integer indices, pointing to the linear array of grid vertices associated to the image block $\mathcal{B}$.
@@
 
-\subsubsection*{Smoothing}\label{}
+\subsubsection*{Smoothing}\label{sec:smoothing}
 The final smoothing of the generated surfaces cannot be performed block-wise, since this would introduce smoothing artifacts at the block boundaries.
@@
 
+\begin{verbatim}
+using Distributed
+addprocs(3)
+using LarSurf
+
+LarSurf.lsp_setup([20, 20, 20]) # set the block size
+
+segmentation = LarSurf.generate_truncated_sphere(10)
+V, FV = LarSurf.lsp_get_surface(segmentation)
+FVtri = LarSurf.triangulate_quads(FV)
+Vs = LarSurf.Smoothing.smoothing_FV_taubin(V, FV, 0.4, -0.3, 50)
+
+objlines = LarSurf.Lar.lar2obj(Vs, FVtri, "truncated_sphere_tri_taubin.obj")
+\end{verbatim}
+
+We include more examples of extracted surfaces based on Micro-CT data of corrosion casts of pig liver.
+\cite{eberlova2017use}.
+% The resolution
+Lar-surf
+
+\begin{minted}{python}
+
+using Pkg
+\end{minted}
+% \begin{minted}[breaklines,escapeinside=||,mathescape=true, linenos, numbersep=3pt, gobble=2, frame=lines, fontsize=\small, framesep=2mm]{julia}
+% Your awesome julia code here\end{minted}
+
 \begin{figure}
@@
 \label{fig:example_porta}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=0.3\textwidth]{figs/liver_01_red_3.png} 
+\vspace{0.05\textwidth}
+\includegraphics[width=0.3\textwidth]{figs/portalvein_01_yellow_3.png} 
+\vspace{0.05\textwidth}
+% \includegraphics[height=0.3\textwidth]{figs/nrn10_100_green.png} 
+\includegraphics[height=0.3\textwidth]{figs/nrn10_100_low_res.png} 
+%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
+\caption{Triangulated isosurface of portal vein calculated with LAR-SURF}
+\label{fig:example_ircad}
 \end{figure}
@@
 
+For perfomance analysis the public dataset from Research Institute against Digestive Cancer (IRCAD) \cite{ircadb} was used. The table \ref{tab:ircad1} and \ref{tab:ircad2} describe the dataset.
+
 \begin{table}
@@
 \bibliographystyle{alpha}
-\bibliography{library,TSAS-2019}
+% \bibliography{library,TSAS-2019,references}
+\bibliography{TSAS-2019,references}
 

### 2019-12-31 src/paper.tex
code highlititing

@@
 
-% M. Jirik added - code highlithing
+%%%%%%% M. Jirik added - code highlithing
 % \usepackage{minted}
-\usepackage[cache=false]{minted}
+% \usepackage[cache=false]{minted}
+\usepackage{listings}
 % \usepackage{minted}
 %SetFonts
+\lstdefinelanguage{Julia}%
+  {morekeywords={abstract,break,case,catch,const,continue,do,else,elseif,%
+      end,export,false,for,function,immutable,import,importall,if,in,%
+      macro,module,otherwise,quote,return,switch,true,try,type,typealias,%
+      using,while},%
+   sensitive=true,%
+   alsoother={$},%
+   morecomment=[l]\#,%
+   morecomment=[n]{\#=}{=\#},%
+   morestring=[s]{"}{"},%
+   morestring=[m]{'}{'},%
+}[keywords,comments,strings]%
+
+\lstset{%
+    language         = Julia,
+    basicstyle       = \ttfamily,
+    keywordstyle     = \bfseries\color{blue},
+    stringstyle      = \color{magenta},
+    % commentstyle     = \color{ForestGreen},
+    commentstyle     = \color{green},
+    showstringspaces = false,
+}
+%%%%%%
 
@@
 
-\begin{verbatim}
+% \begin{verbatim}
+% \begin{lstlisting}
+% using Distributed
+% addprocs(3)
+% using LarSurf
+% LarSurf.lsp_setup([20, 20, 20])
+
+% segmentation = LarSurf.generate_truncated_sphere(10)
+% V, FV = LarSurf.lsp_get_surface(segmentation)
+% FVtri = LarSurf.triangulate_quads(FV)
+% Vs = LarSurf.Smoothing.smoothing_FV_taubin(V,FV,0.4,-0.3,50)
+
+% objlines = LarSurf.Lar.lar2obj(Vs, FVtri, "tri_taubin.obj")
+% \end{lstlisting}
+% \end{verbatim}
+We include more examples of extracted surfaces based on Micro-CT data of corrosion casts of pig liver.
+\cite{eberlova2017use}.
+The size of the visualized data is $100\times100\times100$ voxels. The number of of triangles is 
+544784 and number of vertices is 272826.
+
+% The resolution
+% 
+% pth = Pio3d.datasets_join_path("medical/orig/3Dircadb1.1/MASKS_DICOM/liver")
+\begin{lstlisting}[caption=Read from DICOM files]
 using Distributed
-addprocs(3)
+using Pio3d
+addprocs(3)  # set number of processors
 using LarSurf
 
-LarSurf.lsp_setup([20, 20, 20]) # set the block size
+LarSurf.lsp_setup([64, 64, 64])  # set block size
 
-segmentation = LarSurf.generate_truncated_sphere(10)
-V, FV = LarSurf.lsp_get_surface(segmentation)
+# read data
+datap = Pio3d.read3d("3Dircadb1.1/MASKS_DICOM/liver")
+segmentation = datap["data3d"]
+voxelsize_mm = datap["voxelsize_mm"]
+
+# get surface
+V, FV = LarSurf.lsp_get_surface(segmentation, voxelsize_mm)
 FVtri = LarSurf.triangulate_quads(FV)
-Vs = LarSurf.Smoothing.smoothing_FV_taubin(V, FV, 0.4, -0.3, 50)
 
-objlines = LarSurf.Lar.lar2obj(Vs, FVtri, "truncated_sphere_tri_taubin.obj")
-\end{verbatim}
+# do smoothing and save data
+Vs = LarSurf.Smoothing.smoothing_FV_taubin(V,FV,0.5,-0.2,40)
+objlines = LarSurf.Lar.lar2obj(Vs, FVtri, "liver.obj")
+\end{lstlisting}
 
-We include more examples of extracted surfaces based on Micro-CT data of corrosion casts of pig liver.
-\cite{eberlova2017use}.
-% The resolution
-Lar-surf
+% \begin{minted}{python}
 
-\begin{minted}{python}
-
-using Pkg
-\end{minted}
+% using Pkg
+% \end{minted}
 % \begin{minted}[breaklines,escapeinside=||,mathescape=true, linenos, numbersep=3pt, gobble=2, frame=lines, fontsize=\small, framesep=2mm]{julia}

### 2020-01-01 src/paper.tex
images and data description

@@
 \usepackage{listings}
+\usepackage[dvipsnames]{xcolor}
 % \usepackage{minted}
@@
     stringstyle      = \color{magenta},
-    % commentstyle     = \color{ForestGreen},
-    commentstyle     = \color{green},
+    commentstyle     = \color{ForestGreen},
+    % commentstyle     = \color{green},
     showstringspaces = false,
@@
 
-% \begin{verbatim}
-% \begin{lstlisting}
-% using Distributed
-% addprocs(3)
-% using LarSurf
-% LarSurf.lsp_setup([20, 20, 20])
+\begin{figure}
+\centering
+\includegraphics[width=0.3\textwidth]{figs/liver_01_red_3.png} 
+\vspace{0.05\textwidth}
+\includegraphics[width=0.3\textwidth]{figs/portalvein_01_yellow_3.png} 
+\vspace{0.05\textwidth}
+% \includegraphics[height=0.3\textwidth]{figs/nrn10_100_green.png} 
+\includegraphics[height=0.3\textwidth]{figs/nrn10_100_low_res.png} 
+%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
+\caption{
+Triangulated isosurface of liver parts extracted with LAR-SURF algorithm. 
+On the left image the surface of the liver parenchyma of human. 
+The middle image shows the Portal Vein from the same Ircad dataset \cite{ircad} 
+with resolution $1.6\times0.57\times0.57$ [mm].
+On the right image is the pig liver microvasculature based on corrosion cast prepared by Eberlova
+\cite{eberlova2017use}. The resolution of the Micro-CT data is 4.682 $\mu{}m$.
+}
+\label{fig:example_liver_macro_micro}
+\end{figure}
 
-% segmentation = LarSurf.generate_truncated_sphere(10)
-% V, FV = LarSurf.lsp_get_surface(segmentation)
-% FVtri = LarSurf.triangulate_quads(FV)
-% Vs = LarSurf.Smoothing.smoothing_FV_taubin(V,FV,0.4,-0.3,50)
+The implementation of our algorithm is available in LarSurf package. 
+Use of this package can be seen on listings \ref{lst:example1} where the liver segmentation with 2865131 voxels from
+Ircad dataset is used as an input for our surface extraction algorithm. The size of 3D volumetric 
+image is $129 \times 512 \times 512$
+and the voxel resolution is $1.6\times0.57\times0.57$ [mm]. 
+The output surface model created by  182124 triangles and the number of vertices is 90822. The visualization can 
+be seen on left the image of figure \ref{fig:example_liver_macro_micro}. 
+The portal vein surface extraction can be performed with a small change of input path in code.
+It can be seen in the same image in the middle. The The 3D image resolution is the same. 
+The number of input voxels is 103533. 
+The output surface is created by 90822 vertices and 182124 triangles.
 
-% objlines = LarSurf.Lar.lar2obj(Vs, FVtri, "tri_taubin.obj")
-% \end{lstlisting}
-% \end{verbatim}
-We include more examples of extracted surfaces based on Micro-CT data of corrosion casts of pig liver.
-\cite{eberlova2017use}.
-The size of the visualized data is $100\times100\times100$ voxels. The number of of triangles is 
-544784 and number of vertices is 272826.
-
-% The resolution
-% 
-% pth = Pio3d.datasets_join_path("medical/orig/3Dircadb1.1/MASKS_DICOM/liver")
-\begin{lstlisting}[caption=Read from DICOM files]
+\begin{lstlisting}[caption={Get surface from DICOM volumetric data}, label={lst:example1}]
 using Distributed
-using Pio3d
+using Pio3d  # Read 3D data from DICOM files
 addprocs(3)  # set number of processors
@@
 
-# read data
+# read data from DICOM files
 datap = Pio3d.read3d("3Dircadb1.1/MASKS_DICOM/liver")
@@
 \end{lstlisting}
+
+The right image of figure \ref{fig:example_liver_macro_micro} is surface of a microvasculature of pig liver. 
+Volumetric image is based on Micro-CT data of corrosion casts of pig liver.
+\cite{eberlova2017use}.
+The size of the visualized data is $100\times100\times100$ voxels and size of voxel is 4.682 $\mu{}m$.
+The number of of triangles is 
+544784 and number of vertices is 272826.
 
@@
 \label{fig:example_porta}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=0.3\textwidth]{figs/liver_01_red_3.png} 
-\vspace{0.05\textwidth}
-\includegraphics[width=0.3\textwidth]{figs/portalvein_01_yellow_3.png} 
-\vspace{0.05\textwidth}
-% \includegraphics[height=0.3\textwidth]{figs/nrn10_100_green.png} 
-\includegraphics[height=0.3\textwidth]{figs/nrn10_100_low_res.png} 
-%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
-\caption{Triangulated isosurface of portal vein calculated with LAR-SURF}
-\label{fig:example_ircad}
 \end{figure}

### 2020-01-01 src/paper.tex
Updates from Overleaf

@@
 
+% \includegraphics[height=0.3\textwidth]{figs/nrn10_100_green.png} 
+%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
 \begin{figure}
 \centering
-\includegraphics[width=0.3\textwidth]{figs/liver_01_red_3.png} 
-\vspace{0.05\textwidth}
-\includegraphics[width=0.3\textwidth]{figs/portalvein_01_yellow_3.png} 
-\vspace{0.05\textwidth}
-% \includegraphics[height=0.3\textwidth]{figs/nrn10_100_green.png} 
-\includegraphics[height=0.3\textwidth]{figs/nrn10_100_low_res.png} 
-%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
+\includegraphics[width=0.4\textwidth]{figs/liver_01_red_3.png} 
+% \vspace{0.05\textwidth}
+\includegraphics[width=0.4\textwidth]{figs/portalvein_01_yellow_3.png} 
+% \vspace{0.01\textwidth}
+\includegraphics[height=0.4\textwidth]{figs/nrn10_100_low_res.png} 
+% \vspace{0.05\textwidth}
+\includegraphics[width=0.4\textwidth]{figs/porta_smoothing_2.png} 
 \caption{
 Triangulated isosurface of liver parts extracted with LAR-SURF algorithm. 
-On the left image the surface of the liver parenchyma of human. 
-The middle image shows the Portal Vein from the same Ircad dataset \cite{ircad} 
+On the upper left image the surface of the liver parenchyma of human can be seen. 
+Both right images show the Portal Vein from the same Ircad dataset \cite{ircad} 
 with resolution $1.6\times0.57\times0.57$ [mm].
-On the right image is the pig liver microvasculature based on corrosion cast prepared by Eberlova
+On the left bottom image is the pig liver microvasculature based on corrosion cast prepared by Eberlova
 \cite{eberlova2017use}. The resolution of the Micro-CT data is 4.682 $\mu{}m$.
-}
-\label{fig:example_liver_macro_micro}
+} \label{fig:example_liver_macro_micro}
 \end{figure}
+
+% \begin{figure}
+% \centering
+% \includegraphics[width=0.4\textwidth]{figs/porta_smoothing_1.png} 
+% \vspace{0.05\textwidth}
+% \includegraphics[width=0.4\textwidth]{figs/porta_smoothing_2.png} 
+% %\includegraphics[scale=1]{input/ircad_comparison.pdf} 
+% \caption{Triangulated isosurface of portal vein calculated with LAR-SURF}
+% \label{fig:example_porta}
+% \end{figure}
 
@@
 The output surface model created by  182124 triangles and the number of vertices is 90822. The visualization can 
-be seen on left the image of figure \ref{fig:example_liver_macro_micro}. 
+be seen on figure \ref{fig:example_liver_macro_micro}. 
 The portal vein surface extraction can be performed with a small change of input path in code.
-It can be seen in the same image in the middle. The The 3D image resolution is the same. 
+The The 3D image resolution is the same. 
+It can be seen in the same figure. 
 The number of input voxels is 103533. 
@@
 % Your awesome julia code here\end{minted}
-
-\begin{figure}
-\centering
-\includegraphics[width=0.4\textwidth]{figs/porta_smoothing_1.png} 
-\vspace{0.05\textwidth}
-\includegraphics[width=0.4\textwidth]{figs/porta_smoothing_2.png} 
-%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
-\caption{Triangulated isosurface of portal vein calculated with LAR-SURF}
-\label{fig:example_porta}
-\end{figure}
 

### 2020-04-17 src/paper.tex
new square magenta image
Updates from Overleaf

@@
 
-%TODO fix the name
+% TODO fix the name
 \begin{figure}[htbp] %  figure placement: here, top, bottom, or page
@@
 
-\item remove both instances of \emph{double quads} generated by \texttt{Lar} software at the block boundaries (see Figure~\ref{fig:blockboundary}).
+\item remove both instances of \emph{double quads} generated by \texttt{Lar} software at the block boundaries 
+(see Figure~\ref{fig:blockboundary}). 
+% TODO fix reference probably uncomment next line to link it to Figure 3
+% (see Figure~\ref{fig:example}). 
 They are artifacts generated by the decomposition of the whole image into a number of blocks of tractable size.
@@
 The details of this \emph{artifact filtering} are elucidated in Section~\ref{sec:covering}.
+% TODO fix the reference
 
@@
 The final smoothing of the generated surfaces cannot be performed block-wise, since this would introduce smoothing artifacts at the block boundaries.
-Anyway, the Taubin smoothing~\cite{} can be performed in parallel, since for each vertex in the final surface (except eventually the ones on the image $\mathcal{I}$ boundaries) it essentially consist in computing a new position as a proper average of its neighborhood vertices, i.e.~by applying a discrete Laplacian operator.
+Anyway, the Taubin smoothing~\cite{Taubin1995} can be performed in parallel, since for each vertex in the final surface (except eventually the ones on the image $\mathcal{I}$ boundaries) it essentially consist in computing a new position as a proper average of its neighborhood vertices, i.e.~by applying a discrete Laplacian operator.
 Some appropriate set of workers may so be assigned the task of generating iteratively a new position for the vertices they take cure of.
@@
 % \vspace{0.01\textwidth}
-\includegraphics[height=0.4\textwidth]{figs/nrn10_100_low_res.png} 
+\includegraphics[height=0.4\textwidth]{src/figs/nrn10_100_magenta_high_res.png} 
+% \includegraphics[height=0.4\textwidth]{figs/nrn10_100_low_res.png} 
 % \vspace{0.05\textwidth}

### 2020-04-25 src/background.tex
Updates from Overleaf

@@
 
-\subsection{Representation Schemes}\label{sec:aaaa}
+Some basic concepts of solid modeling, and in particular the foundational idea of representation scheme, as well as few basic concepts of algebraic topology, are shortly introduced in this section, including the computation of the matrix of a boundary operator between chain spaces.
 
-A \emph{representation scheme} for solid modeling is a mapping between a space of mathematical models and a space of symbolic representations, like generated by a formal grammar.
+\subsection{Representation Schemes}\label{sec:schemes}
+
+A \emph{representation scheme} for solid modeling is a mapping between a space of mathematical models and  space of symbolic representations like generated by a formal grammar.
 Solid pointsets (i.e., `$r$-sets') are defined~\cite{Requicha:1980:RRS:356827.356833} as compact (bounded and closed) regular and semianalytic subsets of the $d$-space.
@@
 In particular, a boundary representation provides a cellular decomposition of the object's boundary into \emph{cells} of dimension zero (vertices), one (edges), and two (faces).
-Medical imaging can be classified as \emph{enumerative representation} of cellular decompositions of organs and tissues of interest, in particular, as subsets \emph{of 3D volume elements} (voxels) from the 3D image.
+Medical imaging can be classified as the \emph{enumerative representation} of cellular decompositions of organs and tissues of interest, in particular, as subsets \emph{of 3D volume elements} (voxels) from the 3D image.
 
-\subsection{Linear Algebraic Representation}\label{sec:aaaa}
+\subsection{Linear Algebraic Representation}\label{sec:lar}
 
-The \emph{Linear Algebraic Representation} (\textsc{lar})~\cite{Dicarlo:2014:TNL:2543138.2543294} aims to represent the \emph{chain complex}~\cite{TSAS} generated by a piecewise-linear \emph{geometric complex} embedded either in 2D or in 3D.
-In few words, it gets a minimal characterization of geometry and topology of a cellular complex, i.e.~the embedding mapping $\mu : C_0 \to \E^d$ of 0-cells (vertices), as well a description of $(d-1)$-cells as subsets of vertices, and is able to return the whole chain complex
-\[ 
+The \emph{Linear Algebraic Representation} (\textsc{lar}), introduced in~\cite{Dicarlo:2014:TNL:2543138.2543294}, aims to represent the \emph{chain complex}~\cite{TSAS} generated by a piecewise-linear \emph{geometric complex} embedded either in 2D or in 3D.
+In a few words, it gets a minimal characterization of geometry and topology of a cellular complex, i.e.~the embedding mapping $\mu : C_0 \to \E^d$ of 0-cells (vertices), as well a description of $(d-1)$-cells as subsets of vertices, and is able to return the whole chain complex
+\begin{equation}
 C_\bullet = (C_p, \partial_p) := 
-C_3 \ 
+C_3 
 \substack{
-\delta_2 \\x
-\longleftarrow \\[-1mm]
+\delta_2 \\
+\longleftarrow \\
 \longrightarrow \\
@@
 }
-\ C_2 \ 
+C_2 
 \substack{
 \delta_1 \\
-\longleftarrow \\[-1mm]
+\longleftarrow \\
 \longrightarrow \\
@@
 }
-\ C_1 \ 
+C_1  
 \substack{
 \delta_0 \\
-\longleftarrow \\[-1mm]
+\longleftarrow  \\
 \longrightarrow \\
@@
 }
-\ C_0 .
-\] 
+C_0 .
+\end{equation}
+
+%\[ 
+%C_\bullet = (C_p, \partial_p) := 
+%C_3 \ 
+%\substack{
+%\delta_2 \x
+%\longleftarrow \[-1mm]
+%\longrightarrow \
+%\partial_3 
+%}
+%\ C_2 \ 
+%\substack{
+%\delta_1 \
+%\longleftarrow \[-1mm]
+%\longrightarrow \
+%\partial_2 
+%}
+%\ C_1 \ 
+%\substack{
+%\delta_0 \
+%\longleftarrow \[-1mm]
+%\longrightarrow \
+%\partial_1 
+%}
+%\ C_0 .
+%\]
+
+% {   
+%   \centering\vspace{1mm}
+%   \includegraphics[width=0.475\textwidth]{figs/complex}
+
+% }
+
+\noindent
 and, in particular, any basis for linear chain spaces $C_p$, and any linear
@@
 When the scalar coefficients are from $\{-1, 0, +1\}$, a chain may represent \emph{any (oriented) subset of cells} from the cellular complex.
+Scalars from $\{0, 1\}$ are used for non-oriented complexes.
 
-We may therefore get the $(p-1)$-boundary $\partial_p c_p$ of \emph{any} $p$-chain $c_p$, by multiplication of the coordinate representation $[\partial_p]$ of the boundary operator times the coordinate representation $[c_p]$ of the chain in terms of such scalars, i.e.~by a  matrix-vector product $ [\partial_p] [c_p] $.
+We may, therefore, get the $(p-1)$-boundary $\partial_p c_p$ of \emph{any} $p$-chain $c_p$, by multiplication of the coordinate representation $[\partial_p]$ of the boundary operator times the coordinate representation $[c_p]$ of the chain in terms of such scalars, i.e.~by a  matrix-vector product $ [\partial_p] [c_p] $.
 
@@
 \qquad\equiv\qquad
-\texttt{(geometry, topology) = (W, (EV, FE, CF))}
+\texttt{(geom, top) = (V, (EV, FE, CF))}
 \]
 The GCC allows to transform the (possibly non connected) boundary 2-cycle of surfaces as a standard B-rep~\cite{shapiroSM:202}. 
-The \texttt{geometry} is given by
-the embedding matrix \texttt{W} of vertices (0-cells), and \texttt{topology} is
-given by the three sparse matrices \texttt{CF}, \texttt{FE}, \texttt{EV} of coboundaries
-{i.e.},~$\delta_3, \delta_2, \delta_1$, of  chain complex describing a   
+The geometry \texttt{geom} is given by
+the embedding matrix \texttt{V} of vertices (0-cells), and  topology \texttt{top} by the three sparse matrices (\texttt{EV}, \texttt{FE}, \texttt{CF}) of coboundaries ($\delta_0, \delta_1, \delta_2$) of the chain complex describing a
 space arrangement~\cite{paoluzzi2019finite}.
@@
 
-\subsubsection{Construction of boundary matrix $\partial_d$}
+\subsubsection*{Construction of boundary matrix $\partial_d$}
 
-Once fixed an ordering for all the cells (vertices, edges, pixels, and voxels), i.e.~for 0-, 1-, 2-, and 3-elements of a cell partition \texttt{V,E,F,C} of a 3D image, i.e., once fixed the $p$-bases for linear spaces $C_p$ of $p$-chains $(0\leq p\leq 3)$, we call $M_p = (m_{i,j})$ the \emph{characteristic matrix} of the $p$-basis, expressed as subsets of 0-cells, where  we have that $m_{i,j}=1$ iff the $j$-th $0$-cell belongs to the boundary of $i$-th $p$-cell, and $m_{i,j}=0$ otherwise.
+First, let us fix an ordering for all the cells of a partition of input data (with vertices $V$, edges $E$, pixels $F$, and voxels $C$), i.e.~for each 0-, 1-, 2-, and 3-elements of a cell partition \texttt{V,E,F,C} of a 3D image, i.e., once fixed the $p$-bases for linear spaces $C_p$ of $p$-chains $(0\leq p\leq 3)$, we call $M_p = (m_{i,j})$ the \emph{characteristic matrix} of the $p$-basis, expressed as subsets so that
+$m_{i,j}=1$ if and only if the $j$-th  $0$-cell $c^j_0$ belongs 
+% , where  we have that $m_{i,j}=1$ if and only if the $j$-th $0$-cell belongs 
+to the boundary of $i$-th $p$-cell $m_{i,j}$, and $m_{i,j}=0$ otherwise.
 
-Note that, by computing the (sparse) matrix product $(M_{p-1} M_{p}^t) = (n_{i,j})$, with $n_{i,j} = \sum_{k} m_{i,k}m_{k,j}$, we get for each $n_{ij}$ the \emph{number of vertices} shared by $c_{p-1}^i$ and $c_{p}^j$.
+Let us note that the product of binary matrices is not binary, so by computing the (sparse) matrix product $(M_{p-1} M_{p}^t) = (n_{i,j})$, with $n_{i,j} = \sum_{k} m_{i,k}m_{k,j}$, we get for each $n_{ij}$ the \emph{number of vertices} shared by $c_{p-1}^i$ and $c_{p}^j$.
 When this number equates the cardinality of $c_{p-1}^i$, this elementary chain is contained on the boundary of $c_{p}^j$.
-In a 3D image, with cubic 3-cells and squared 2-cells in between, everywhere we get $n_{i,j}=4$, we may state $c_{2}^i\subset\partial c_{3}^j$.
-In this case, by looking in each $j$ column of $M_{2} M_{3}^t = (n_{i,j})$, we have exactly \emph{six rows} where  $n_{i,j} = 4$.
+In a volumetric data, with cubic 3-cells and squared 2-cells in-between, everywhere we get $n_{i,j}=4$, we may state $c_{2}^i\subset\partial c_{3}^j$.
+Therefore, in each $j$ column of $M_{2} M_{3}^t = (n_{i,j})$, we have exactly \emph{six rows} where  $n_{i,j} = 4$, since a cube (3-chain) has six boundary faces (2-chains).
+The unit incidence coefficients in $\left[\partial_3\right]$ are accordingly located by filtering the coefficients with value 4.
 
 Finally, consider the linear graded boundary operator $\partial_p : C_p \to C_{p-1}$.
-As such, it contains by columns the representation of domain basis elements, expressed as linear combination of the basis elements of the range space.
-Therefore, the operator matrix $[\partial_d]$ is readily obtained by setting $[\partial_d](i,j)=1$ iff $n_{i,j}=4$ and $[\partial_d](i,j)=0$ otherwise.
-Of course, it will contain six non-zero elements for column.
-It may be worth to remember that every 3-cell (voxel) of 3D image has exactly six 2-faces.
+As such, it contains by columns the representation of domain basis elements, expressed as a linear combination of the basis elements of the range space.
+Therefore, the operator matrix $[\partial_d]$ is readily obtained by setting $[\partial_d](i,j)=1$ if $n_{i,j}=4$ and $[\partial_d](i,j)=0$ otherwise.
+Of course, it will contain six non-zero elements for the column.
+It may be worth remembering every 3-cell (voxel) of the volumetric data has exactly six 2-faces.
 
@@
 We~may also show that such matrices are \emph{very sparse}, with their sparseness growing rapidly with the dimensions (see Section~\ref{sec:block}).
-The pattern of non-zeros in matrix $[\partial_3]$ corresponding to a brick of shape $(4,4,4)$ is given in Figure~\ref{fig:boundary_matrix_4x4x4}.
+The pattern of non-zeros in matrix $[\partial_3]$ corresponding to a brick of shape $(4,4,4)$ is given in Fig.~\ref{fig:boundary_matrix_4x4x4}.
 
-\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+\begin{figure}[tbp] %  figure placement: here, top, bottom, or page
    \centering
    \includegraphics[width=0.75\linewidth]{figs/boundary_matrix_4x4x4.pdf} 
-   \caption{A binary image of the coboundary operator  $\delta_2 = \partial_3^\top : C_2 \to C_3$, built for a small 3D image with shape $(4,4,4)$.
-Note that the number of rows equates the cardinality $4\times 4\times 4 = 64$ of the voxel set; the number of columns is $d\,n\,(1+n)^{d-1} = 3\times 4\times 25 = 300$.
-Of course, the number of non-zeros per row (cardinality of single voxel facets) is six, whereas the number of non-zeros per column is two, but on boundary facets.}
+   \caption{
+   The binary image of \emph{sparse} coboundary matrix  $\left[\delta_2\right] = \left[\partial_3\right]^t : C_2 \to C_3$,
+%   The binary image of the coboundary operator  $\delta_2 = \partial_3^\top : C_2 \to C_3$, 
+   built for a small volumetric data (or a brick) with shape $(4,4,4)$.
+Note that the number of rows equates the size $4\times 4\times 4 = 64$ of the voxel set; the number of columns is $d\,n\,(1+n)^{d-1} = 3\times 4\times 25 = 300$.
+Of course, the number of non-zeros per row (cardinality of the facet set of a single voxel) is six, whereas the number of non-zeros per column is two, but on boundary facets.}
    \label{fig:boundary_matrix_4x4x4}
@@
 
-\subsection{Taubin Smoothing}\label{sec:aaaa}
+\subsection{Taubin Smoothing}\label{sec:taubin}
+
+Every boundary chain extracted from an image block $\B(i,j,k,n)$ is a \emph{2-cycle}, i.e., a closed 2-chain---in other words, a 2-chain with empty boundary.
+Such 2-cycles are joined together by removing the double 2-cells (at the boundaries of adjacent bricks) after having suitably shifted their indices to an unique linear representation of the whole image.
+The resulting raster surface is made by mutually orthogonal raster facets, that must be smoothed in order to get a fair surface.
+A linear time and space algorithm for this purpose is the Laplacian smoothing, which iteratively  moves each vertex (0-cell) to the centroid of its neighbors.
+A well known weakness of this simple algorithm is the asymptotic convergence of the whole mesh to a single point, resulting in unfair size reduction even after few iterations.
+Conversely, the Taubin smoothing algorithm~\cite{Taubin:1995:SPA:218380.218473,egst.20001029} alternates two Laplacian smoothing steps with \emph{shrink} and \emph{inflate} effects respectively, with the result of delivering pretty invariant sizes and volume of the smoothed mesh.
+The best results are obtained on meshes which have small variations of edge length and face angles, like for surfaces extracted from 3D raster images, as in our case.

### 2020-04-25 src/intro.tex
Updates from Overleaf

\section{Introduction}\label{sec:intro}

Isosurface extraction to produce geometric models of surfaces from volumetric data is important in many applications. It is often used for indirect visualization of the medical data or for flow modeling \cite{Rohan2018a}. 
% In this paper we suggest an approach based on

% Input volumetric data are represented by a 3D voxel field and can be generated by segmentation of sections from computed tomography. 
The most popular algorithm used for surface extraction is probably Marching Cubes (MC). The algorithm was described by Lorentsen and Cline \cite{Lorensen1987} in 1987. The survey of Marching Cubes algorithms has been published in 2006 \cite{Newman2006}. The algorithm is based on considering the cube defining volume. Each corner vertex of the cube is related to input volumetric data. MC traverse the data and constructs the surface by using a lookup table of different triangular faces depending on different patterns of the cube.  The main disadvantages of this method are time requirements, ambiguity, and holes generation. Some of them were discovered shortly after the algorithm was introduced. 
Marching Cubes. In 1991 Nielson and Hamman described an Asymptotic Decider to solve the ambiguity problem on the faces of the cube.  Natarajan noted that the ambiguity problem also occurs with uniform samples \cite{Natarajan1994}. In 1995 Chernyaev extended the number of cases to 33 \cite{chernyaev1995marching}. More recently the algorithm was updated by Custodio, Pesco, and Silva to enhance the quality of iso-surface triangulation \cite{Custodio2019}.

Some alternative methods have been developed, including a method for surface extraction using particle attraction; a system was described by Crossno and Angel in \cite{Crossno1997}. A graph processing that tracks the boundary cell-face adjacencies is described in \cite{Lachaud2000}. Some parallel algorithms for iso-surface extraction are discussed in \cite{Bajaj2004}.
A completely data-parallel algorithm, implemented in OpenCL that runs entirely on the GPU is presented in~\cite{Smistad12}. 
A Linear Algebraic Representation approach, parallelized using the OpenCL framework on Linux, was discussed in~\cite{paodcvjcadanda2015}.

% In this paper
% TODO slightly change the formulation
Here we discuss an alternative approach for surface extraction. Our \textsc{lar-surf} (Linear Algebraic Representation Surface extraction) filter is based on basic algebraic topology and linear algebra, using linear spaces $C_p$ of chains (of cells) of dimension $0 \leq p \leq 3$ and the boundary matrix $[\partial_3] : C_3 \to C_2$.
A decomposition of the input volumetric data into small submatrices called \emph{bricks} is performed, then the binary coordinate vector of each interesting chain of voxels is generated, and its boundary is computed by matrix multiplication times the boundary matrix producing the binary representation of the boundary surface (the surface which defines the boundary). 
Embarrassing parallel data decomposition is used to compute the boundary surface patches within each of the bricks, that are finally joined and smoothed via the Taubin algorithm \cite{Taubin1995}.

The present paper is organized as follows.
Section~\ref{sec:background} provides the basic topological and geometrical concepts needed to understand the \textsc{lar-surf} method, including the building of boundary matrices, the map from Cartesian indices to linear indices, and the Taubin smoothing method.
Section~\ref{sec:filter} discusses the parametric design of the unit block filtered by the parallel algorithm, including the block decomposition, the sparsity rate of the used sparse arrays, and the block-level parallelism.
Section~\ref{sec:julia} is related to the algorithm implementation in Julia, and in particular to a discussion of the parallel workflow.
Section~\ref{sec:examples} presents some examples of algorithm execution on the liver and the hepatic portal system.
Section~\ref{sec:conclusion} shortly describes the next extensions of this approach, in particular the implementation with Julia's support for GPU parallelism and the multi-segmentation of Medical images.

### 2020-04-25 src/paper.tex
Updates from Overleaf

--- 2020-01-01 src/paper.tex
@@
-\documentclass[11pt, oneside]{amsart}   	% use "amsart" instead of "article" for AMSLaTeX format
-\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
-\geometry{letterpaper}                   		% ... or a4paper or a5paper or ...
+\documentclass[9pt,academicons]{article}
+
+\usepackage{CADA}
+
+%%%%%%%%%%%%%%%% original head %%%%%%%%%%%%%%%%%%
+
+% \documentclass[11pt, oneside]{amsart}   	% use "amsart" instead of "article" for AMSLaTeX format
+% \usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
+% \geometry{letterpaper}                   		% ... or a4paper or a5paper or ...
 %\geometry{landscape}                		% Activate for rotated page geometry
@@
 %SetFonts
-\usepackage[utf8x]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage{lmodern}
+% \usepackage[utf8x]{inputenc}
+% \usepackage[T1]{fontenc}
+% \usepackage{lmodern}
 
@@
 \usepackage{siunitx}
-\usepackage{booktabs}
+\usepackage{booktabs} % tables
 
@@
    sensitive=true,%
-   alsoother={$},%
+%   alsoother={$},%
    morecomment=[l]\#,%
@@
 %%%%%%
+% TODO remove
+\usepackage{todonotes}
 
-\title{Algebraic filtering of surfaces from medical images with Julia}
-\author{Miroslav Jirik and Alberto Paoluzzi}
+\title{Algebraic Filtering of Surfaces from 3D Medical Images with Julia}
+% \author{Miroslav Jirik and Alberto Paoluzzi}
 \date{}							% Activate to display a given date or no date
 
-\usepackage{draftwatermark}
-\SetWatermarkText{Draft}
-\SetWatermarkScale{2.0}
-\SetWatermarkColor[rgb]{0.9,0.9,0.9}
+% \usepackage{draftwatermark}
+% \SetWatermarkText{Draft}
+% \SetWatermarkScale{2.0}
+% \SetWatermarkColor[rgb]{0.9,0.9,0.9}
 
@@
 
-\begin{abstract}
-In this paper we introduce a novel algebraic \textsc{lar-surf} filter, well founded on algebraic topology methods, to extract and smooth the boundary surface of any subset of voxels arising from  segmentation of a 3D medical image.
-The input is defined as a \emph{chain}, i.e.~as a vector from a linear space of 3-chains, represented in coordinates as a sparse binary vector.
-The output is produced by a linear mapping between spaces of 3- and 2-chains through the boundary operator $\partial_3:C_3 \to C_2$.
-In particular, when the input set of voxels is either not (4-)connected, or contains one or more empty regions inside, \textsc{lar-surf} generates a non connected set of closed surfaces, i.e.~a set of 2-cycles---using the language of algebraic topology.
-The only data structures used by this approach are sparse arrays with one or two indices, i.e.~sparse vectors and matrices.
-This work is based on LAR (Linear Algebraic Representation) methods, and is implemented in Julia language, natively supporting parallel computing on hybrid architectures.
-\end{abstract}
+% ############################################
+% Here is the fun part. Please enter your:            
+%      Name (First, Middle initial, Last)                  
+%      Your ORCID number replace the "000-0000-1234-5678" part.    
+% ############################################
 
-\tableofcontents
-
-%\section{}
-%\subsection{}
-
-%\input{intro}
-\section{Introduction}\label{sec:intro}
-
-Isosurface extraction to produce geometric models of surfaces from volumetric data is important in many aplications.
-It is often used for indirect visualization of the medical data or for flow modeling \cite{Rohan2018a}.
-In this paper we discuss an approach based on basic algebraic topology and linear algebra, using linear spaces $C_p$ of chains (of cells) of dimension $0 \leq p \leq 3$ and the boundary matrix $[\partial_3] : C_3 \to C_2$.
-A decomposition of the 3D image into small submatrices called ``bricks'' is performed, the binary coordinate vector of each interesting chain of voxels is generated, and its boundary is computed by matrix multiplication times the boundary matrix.
-An embarrassing parallel approach is used to compute the boundary surface patches within each brick, that are finally joined.
-
-The most popular algorithm used for surface extraction is probably Marching Cubes (MC).
-The algorithm was described by Lorentsen and Cline \cite{Lorensen1987} in 1987.
-Survey of Marching Cubes algorthm has been published in 2006 \cite{Newman2006}.
-The algorithm is based on considering the cube defining volume.
-Each corner vertex of the cube is related to input volumetric data.
-MC traverse the data and constructs the surface by using lookup table of different triangular faces depending on different patterns of the cube.
-Main disadvantages of this method are time requirements, ambiquity and holes generation.
-Some of them were discovered shortly after the algorithm was introduced.
-Marching Cubes.
-In 1991 Nielson and Hamman described an Asymptotic Decider to solve the ambiguity problem on the faces of the cube.
-Natarajan noted that the ambiguity problem also occurs with uniform samples \cite{Natarajan1994}.
-In 1995 Chernyaev extended the number of cases to 33 \cite{chernyaev1995marching}.
-More recently the algorithm was updated by Custodio, Pesco and Silva to enhance the quality of iso-surface triangulation \cite{Custodio2019}.
-
-Some alternative methods have been developed, including a method for surface extraction using particle attraction; a system was described by Crossno and Angel in \cite{Crossno2002}.
-A graph processing that tracks the boundary cell-face adjacencies is described in \cite{Lachaud2000}.
-Some parallel algorithms for iso-surface extraction are discussed in \cite{Bajaj2004}.
-A completely data-parallel algorithm, implementation written in OpenCL that runs entirely on the GPU is presented in~\cite{Smistad:12}. 
-A Linear Algebraic Representation algebraic approach, parallelized using the OpenCL framework on Linux was discussed in~\cite{paodcvjcadanda2015}.
-
-The present paper is organized as follows.
-Section~\ref{sec:background} provides the basic topological and geometrical concepts needed to understand the \textsc{lar-surf} method, including the building of boundary matrices, the map from Cartesian indices to linear indices, and the Taubin smoothing method.
-Section~\ref{sec:filter} discusses the parametric design of the unit block filtered by the parallel algorithm, including the block decomposition, the sparsity rate of the used sparse arrays, the block-level parallelism.
-Section~\ref{sec:julia} is related to the algorithm implementation in Julia, and in particular to a discussion of the parallel workflow.
-Section~\ref{sec:examples} presents some examples of algorithm execution on the liver and the hepatic portal system.
-Section~\ref{sec:conclusion} shortly describes the next extensions of this approach, in particular to implementation with Julia support for GPU parallelism and the multi-segmentation of Medical images.
-
-%\input{background}
-\section{Background}\label{sec:background}
-
-Some basic concepts of solid modeling, and in particular the foundational idea of representation scheme, as well as few basic concepts of algebraic topology, are shortly introduced in this Section, including the computation of the matrix of a boundary operator between chain spaces.
-
-\subsection{Representation Schemes}\label{sec:schemes}
-
-A \emph{representation scheme} for solid modeling is a mapping between a space of mathematical models and a space of symbolic representations, like generated by a formal grammar.
-Solid pointsets (i.e., `$r$-sets') are defined~\cite{Requicha:1980:RRS:356827.356833} as compact (bounded and closed) regular and semianalytic subsets of the $d$-space.
-A large number of representation schemes were defined in the past forty years, including the two main classes of (a) \emph{boundary representations} (`$B$-reps'), where the solid model is represented through a representation of its boundary elements, i.e.~faces, edges and vertices, and (b) \emph{decompositive/enumerative representations}, that are a decomposition of either the object or the embedding space, respectively, into a well-defined \emph{cellular complex}.
-In particular, a boundary representation provides a cellular decomposition of the object's boundary into \emph{cells} of dimension zero (vertices), one (edges), and two (faces).
-Medical imaging can be classified as \emph{enumerative representation} of cellular decompositions of organs and tissues of interest, in particular, as subsets \emph{of 3D volume elements} (voxels) from the 3D image.
-
-\subsection{Linear Algebraic Representation}\label{sec:lar}
-
-The \emph{Linear Algebraic Representation} (\textsc{lar}), introduced in~\cite{Dicarlo:2014:TNL:2543138.2543294}, aims to represent the \emph{chain complex}~\cite{TSAS} generated by a piecewise-linear \emph{geometric complex} embedded either in 2D or in 3D.
-In few words, it gets a minimal characterization of geometry and topology of a cellular complex, i.e.~the embedding mapping $\mu : C_0 \to \E^d$ of 0-cells (vertices), as well a description of $(d-1)$-cells as subsets of vertices, and is able to return the whole chain complex
-%\[ 
-%C_\bullet = (C_p, \partial_p) := 
-%C_3 \ 
-%\substack{
-%\delta_2 \x
-%\longleftarrow \[-1mm]
-%\longrightarrow \
-%\partial_3 
-%}
-%\ C_2 \ 
-%\substack{
-%\delta_1 \
-%\longleftarrow \[-1mm]
-%\longrightarrow \
-%\partial_2 
-%}
-%\ C_1 \ 
-%\substack{
-%\delta_0 \
-%\longleftarrow \[-1mm]
-%\longrightarrow \
-%\partial_1 
-%}
-%\ C_0 .
-%\]
-
-{   
-   \centering\vspace{1mm}
-   \includegraphics[width=0.475\textwidth]{figs/complex}
-
+\authorSection{
+	\anAuthor{Miroslav Jirik}{0000-0002-8002-2079}{1,3},
+	\anAuthor{Alberto Paoluzzi}{0000-0002-3958-8089}{2}, 
+% 	\anAuthor{First M. Last}{000-0000-1234-5678}{3} 
 }
 
-\noindent
-and, in particular, any basis for linear chain spaces $C_p$, and any linear
-boundary/coboundary map \(\partial_p\) and
-\(\delta_p=\partial_{p-1}^\top\) between them.
-The \emph{domain} of \textsc{lar} is the set of \textbf{chain complexes} generated by cell $d$-complexes ($2\leq d\leq 3$).
-The computer \emph{representations} of \textsc{lar} are \textbf{sparse binary matrices} to represent both the operators and the chain bases.
-Note that in algebraic topology a $p$-chain is defined as a linear combination of $p$-cells with scalars from a field.
-When the scalar coefficients are from $\{-1, 0, +1\}$, a chain may represent \emph{any (oriented) subset of cells} from the cellular complex.
+% ######################################
+% Here we need your affiliation and contact e-mail. Please edit   
+% affiliation as well as the e-mail fields.                                  
+% ######################################
+\affiliationSection{
+	\anAffiliation{1}{
+	Charles University,
+    Faculty of Medicine in Pilsen, 
+    Biomedical Center}{jirik@lfp.cuni.cz}
+	\anAffiliation{2}{Roma Tre University, Department of Mathematics and Physics}{paoluzzi@dia.uniroma3.it}
+	\anAffiliation{3}{NTIS, Faculty of Applied Sciences, University of West Bohemia}{}
+}
 
-We may therefore get the $(p-1)$-boundary $\partial_p c_p$ of \emph{any} $p$-chain $c_p$, by multiplication of the coordinate representation $[\partial_p]$ of the boundary operator times the coordinate representation $[c_p]$ of the chain in terms of such scalars, i.e.~by a  matrix-vector product $ [\partial_p] [c_p] $.
+% ######################################
+% Please decide on who the corresponding author is going to be 
+% and complete the section below                                            
+% ######################################
+\correspondingAuthor{Miroslav Jirik}{jirik@lfp.cuni.cz}
 
-It is possible to show that the \textsc{lar} representation scheme is very expressive, i.e.~that it  has a large domain,  including collections of: line segments, quads, triangles, polygons, meshes;  pixels, voxels, volume images; B-reps, enumerative and decompositive representations of solids.
-In this paper we apply \textsc{lar} methods to computation of boundary representations of solid models from segmentation (labeling) of 3D medical images.
-To display a triangulation of  boundary faces  in their proper position in space, the information required is contained in the \emph{geometric chain complex} (GCC):
-\[
-\mu: C_0\to\E^3,\ (\delta_0, \delta_1, \delta_2)
-\qquad\equiv\qquad
-\texttt{(geom, top) = (W, (EV, FE, CF))}
-\]
-The GCC allows to transform the (possibly non connected) boundary 2-cycle of surfaces as a standard B-rep~\cite{shapiroSM:202}. 
-The geometry \texttt{geom} is given by
-the embedding matrix \texttt{W} of vertices (0-cells), and  topology \texttt{top} by the three sparse matrices (\texttt{EV}, \texttt{FE}, \texttt{CF}) of coboundaries ($\delta_0, \delta_1, \delta_2$) of the chain complex describing a
-space arrangement~\cite{paoluzzi2019finite}.
-Note that ordered pairs of letters from \texttt{V,E,F,C}, correspond to \emph{\emph{\texttt{V}}ertices$\to$\emph{\texttt{E}}dges$\to$\emph{\texttt{F}}aces$\to$\emph{\texttt{C}}ells} into the 
-\emph{\emph{\texttt{C}}olumn$\to$\emph{\texttt{R}}ow} order of matrix maps of operators.
+% ########################################
+% Please type in your abstract below after the "\abstract{" part.
+% ########################################
 
-\subsubsection*{Construction of boundary matrix $\partial_d$}
+\abstract{
+In this paper we introduce a novel algebraic 
+% \textsc{lar-surf} 
+filter, well-founded on algebraic topology methods, to extract and smooth the boundary surface of any subset of voxels arising from  the segmentation of a 3D medical image. 
+% TODO 3D medical image
+The input of our Linear Algebraic Representation Surface extraction filter (\textsc{lar-surf}) is defined as a \emph{chain}, i.e.~as a vector from a linear space of 3-chains, represented in coordinates as a sparse binary vector.
+The output is produced by a linear mapping between spaces of 3- and 2-chains through the boundary operator $\partial_3:C_3 \to C_2$.
+% In particular, when the input set of voxels is either not (4-)connected, or contains one or more empty regions inside, \textsc{lar-surf} generates a non-connected set of closed surfaces, i.e.
+quadrilateral faces or~a set of 2-cycles---using the language of algebraic topology.
+The only data structures used by this approach are sparse arrays with one or two indices, i.e.~sparse vectors and matrices.
+This work is based on Linear Algebraic Representation methods and is implemented in Julia language, natively supporting parallel  computing on hybrid hardware architectures.
+}
 
-Once fixed an ordering for all the cells (vertices, edges, pixels, and voxels), i.e.~for 0-, 1-, 2-, and 3-elements of a cell partition \texttt{V,E,F,C} of a 3D image, i.e., once fixed the $p$-bases for linear spaces $C_p$ of $p$-chains $(0\leq p\leq 3)$, we call $M_p = (m_{i,j})$ the \emph{characteristic matrix} of the $p$-basis, expressed as subsets of 0-cells, where  we have that $m_{i,j}=1$ if and only if the $j$-th $0$-cell belongs to the boundary of $i$-th $p$-cell, and $m_{i,j}=0$ otherwise.
+\keywords{
+Medical 3D, Computational Topology, Linear Algebraic Representation, LAR, Julia, Surface Extraction
+}
 
-Note that the product of binary matrices is not binary, so by computing the (sparse) matrix product $(M_{p-1} M_{p}^t) = (n_{i,j})$, with $n_{i,j} = \sum_{k} m_{i,k}m_{k,j}$, we get for each $n_{ij}$ the \emph{number of vertices} shared by $c_{p-1}^i$ and $c_{p}^j$.
-When this number equates the cardinality of $c_{p-1}^i$, this elementary chain is contained on the boundary of $c_{p}^j$.
-In a 3D image, with cubic 3-cells and squared 2-cells in between, everywhere we get $n_{i,j}=4$, we may state $c_{2}^i\subset\partial c_{3}^j$.
-In this case, by looking in each $j$ column of $M_{2} M_{3}^t = (n_{i,j})$, we have exactly \emph{six rows} where  $n_{i,j} = 4$.
+% DOI: 10.14733/cadconfP.2020.xxx-yyy
+\doi{10.3722/cadaps.2020.xxx-yyy}
 
-Finally, consider the linear graded boundary operator $\partial_p : C_p \to C_{p-1}$.
-As such, it contains by columns the representation of domain basis elements, expressed as linear combination of the basis elements of the range space.
-Therefore, the operator matrix $[\partial_d]$ is readily obtained by setting $[\partial_d](i,j)=1$ iff $n_{i,j}=4$ and $[\partial_d](i,j)=0$ otherwise.
-Of course, it will contain six non-zero elements for column.
-It may be worth to remember that every 3-cell (voxel) of 3D image has exactly six 2-faces.
+% \tableofcontents
 
-It is possible to show that all the interesting relations of incidence/adjacency between cells of different dimensions can be both computed and efficiently queried by pairwise computing some matrix products, with one of terms possibly transposed, using only the two boundary and coboundary operator matrices $[\partial_p]$ and $[\delta_p]$, and where $[\delta_p] = [\partial_p^\top]$.
-We~may also show that such matrices are \emph{very sparse}, with their sparseness growing rapidly with the dimensions (see Section~\ref{sec:block}).
-The pattern of non-zeros in matrix $[\partial_3]$ corresponding to a brick of shape $(4,4,4)$ is given in Figure~\ref{fig:boundary_matrix_4x4x4}.
+\input{intro}
+\input{background}
+\input{design}
+\input{implementation}
+\input{examples}
+\input{conclusion}
 
-\begin{figure}[tbp] %  figure placement: here, top, bottom, or page
-   \centering
-   \includegraphics[width=0.75\linewidth]{figs/boundary_matrix_4x4x4.pdf} 
-   \caption{A binary image of the coboundary operator  $\delta_2 = \partial_3^\top : C_2 \to C_3$, built for a small 3D image with shape $(4,4,4)$.
-Note that the number of rows equates the cardinality $4\times 4\times 4 = 64$ of the voxel set; the number of columns is $d\,n\,(1+n)^{d-1} = 3\times 4\times 25 = 300$.
-Of course, the number of non-zeros per row (cardinality of the facet set of a single voxel) is six, whereas the number of non-zeros per column is two, but on boundary facets.}
-   \label{fig:boundary_matrix_4x4x4}
-\end{figure}
-
-\subsection{Multiindices from Cartesian indices}\label{sec:inds-from-cart}
-
-In order to utilize the topological algebra shortly recalled in this paper, we need to explicitly sort the cells of the various dimensions into linearly ordered sequences, possibly according to the linear order their information is linearly accommodated in computer storage.
-
-\subsection{Taubin Smoothing}\label{sec:taubin}
-
-Every boundary chain extracted from an image block $\B(i,j,k,n)$ is a \emph{2-cycle}, i.e., a closed 2-chain---in other words, a 2-chain with empty boundary.
-Such 2-cycles are joined together by removing the double 2-cells (at the boundaries of adjacent bricks) after having suitably shifted their indices to an unique linear representation of the whole image.
-The resulting raster surface is made by mutually orthogonal raster facets, that must be smoothed in order to get a fair surface.
-A linear time and space algorithm for this purpose is the Laplacian smoothing, which iteratively  moves each vertex (0-cell) to the centroid of its neighbors.
-A well known weakness of this simple algorithm is the asymptotic convergence of the whole mesh to a single point, resulting in unfair size reduction even after few iterations.
-Conversely, the Taubin smoothing algorithm~\cite{Taubin:1995:SPA:218380.218473,egst.20001029} alternates two Laplacian smoothing steps with \emph{shrink} and \emph{inflate} effects respectively, with the result of delivering pretty invariant sizes and volume of the smoothed mesh.
-The best results are obtained on meshes which have small variations of edge length and face angles, like for surfaces extracted from 3D raster images, as in our case.
-
-%\input{design}
-
-\section{Block-parametric design}\label{sec:filter}
-
-\subsection{Block decomposition}\label{sec:block-decomposition}
-
-We assume that medical devices produce 3D images with lateral dimensions that are integer multiples of some powers of two, like 128, 256, 512, etc.
-Any cuboidal portion of image is completely determined by the Cartesian indices of its voxels of lowest and highest indices, and extracted by multidimensional array \emph{slicing} as $Image([\ell_x : h_x, \ell_y : h_y, \ell_z : h_z])$.
-
-For the sake of simplicity, we assume a common size on the three image axes, and the corresponding image portion $\B$, called \emph{block}, as a function of its element of the  lowest  block  coordinates $i,j,k\in [1:n]$ and of block dimension $n\in\N$:
-\[
-\B(i,j,k,n) := Image([in:in+n, jn:jn+n, kn:kn+n]) 
-\]
-
-\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
-   \centering
-   \includegraphics[width=0.5\linewidth]{figs/blocks} 
-   \caption{A possible block partitioning of a radiologic image. The evidenced 2D block, of size $n^d=64^2$, is sliced by $\B([2,1,64]) = Image([128:172],[64:128)]$}
-   \label{fig:blocks}
-\end{figure}
-
-Figure~\ref{fig:blocks} shows the block decomposition in a 2D image, with positive integers $(u,v)$ giving the lateral sizes of image.
-Note that block sides do not necessarily correspond to image edges.
-
-\subsection{Block operator }\label{sec:block}
-
-\subsubsection*{Chain coordinates }\label{sec:chain-coords}
-We are going to treat each image block independently from each other.
-Hence we map each image block $\B(i,j,k,n)$ to the linear \emph{chain} space $C_2$ of dimension $n\times n\times n$, using coordinate vectors $c\in \mathbf{2}^{n^d} := \{0,1\}^{n^d}$, where the basis element $c \in C_2$ is mapped via Cartesian-to-linear map to the binary vector
-\[
-Image(h,k) \mapsto c_{h,k} := [0 \cdots 0\ 1\ 0 \cdots 0] \in \B^{n\times n}
-\]
-for each $0\leq h,k \leq n$, and where the (single) unit element is in position $nk + h \leq n\times n$.
-
-Therefore, each pixel (or voxel) in a block image will be seen as a basis binary vector in $C_2$, and each subset of image elements, as the corresponding binary vector in $C_2$, with many ones as the cardinality of the subset.
-
-\subsubsection*{Boundary operator }\label{sec:boundary-operator}
-For a fixed block size $n$, the boundary operator $\partial_d : C_d\to C_{d-1}$, with $d\in\{2,3\}$, will be constructed once and for all using the algorithm given in~\cite{}, and inlined in the generated boundary extraction code.
-
-It is easy to see that the operator's matrix $[\partial_d]$ is \emph{very sparse}, since it contains $2\times d$ non-zero elements (ones) for each column (of length $n^d$), i.e.~4 ones and 6 ones for the 2D and 3D case, respectively.
-In fact the matrix of a linear operator between linear spaces contains by columns the basis element of the domain space, represented in the target space.
-In our case, the former is an image element (2-cube or 3-cube), represented as the chain of its boundary---i.e.
-either a 1-cycle of 4 edges, or  a 2-cycle of 6 faces, respectively.
-
-The number of rows of $[\partial_d]$ equates the dimension of the linear space $C_{d-1}$, i.e.~the number of $(d-1)$-cells---elementary $(d-1)$-chains---in the cellular partition of the image.
-To compute their number, we act in two steps.
-(a) First we map one-to-one the $n^d$ $d$-cells with $d$ adjacent $(d-1)$-cells, so getting $d\,n^d$ distinct basis elements of $C_{d-1}$.
-(b) Then we complete this bases by adjoining $n^{d-1}$ boundary elements for each of the $d$ dimensions of the image, so providing further $d\,n^{d-1}$ basis elements for $C_{d-1}$.
-The dimension of $C_{d-1}$, and therefore the number of rows of $[\partial_d]$ matrix is $d\,(n^{d-1}+n^{d}) = d\,n\,(1+n)^{d-1}$.
-The number of column equates the number of basis elements of $C_d$, i.e.~the number $n^d$ of block elements.
-
-\subsubsection*{Sparsity and size of boundary matrix }\label{sec:bm-size}
-
-As we have seen, we have $2d$ non-zero elements for each column of $[\partial_d]$, so that their total number is $2d\,n^d$.
-The number of matrix element is $d\,n\,(1+n)^{d-1} \times n^d$, giving a ratio of
-\[
-\frac{\mbox{non-zero\ elements}}{\mbox{total\ elements}} = 
-\frac{2d\times n^d}{d\,n\,(1+n)^{d-1} \times n^d} =
-\frac{2}{n+n^d}
-\]
-Using sparse matrices in CSC (Compressed Sparse Column) format we get a storage size:
-\[
-mem([\partial_d]_{n^d}) = 2\times \#\mbox{nzero} + \#\mbox{columns} = 2\times 2d\,n^d + n^d = (4d+1)n^d.
-\]
-In conclusion, for block size $n=64$, the matrix $[\partial_d]$ requires for 2D images $9\times 64^2=36,864$ memory elements, and for 3D images $13\times 64^3=3,407,872$ memory elements.
-Counting the bytes for the standard implementation of a sparse binary matrix (1 byte for values and 8 bytes for indices) we get $(18d+8)n^d$ bytes, giving $176$\,KB for 2D and $15.872$\,MB for 3D.
-
-%TODO fix the name
-\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
-   \centering
-   \includegraphics[width=0.4\linewidth]{figs/grid.pdf} 
-   \caption{example caption}
-   \label{fig:example}
-\end{figure}
-
-\subsection{Block boundary mapping}\label{sec:block-mapping}
-
-Here we refer directly to the 3D case.
-Let us call \emph{segment} the bulk content $S$ of interest within the input 3D image of size $(u,v,w)$. Our aim is to compute the segment boundary $\partial_3 S$. 
-First we set the size $n$ of the block, in order to decompose the input $Image(u,v,w)$ into a fair number of blocks
-\[
-M = \ceil{u/n} \times \ceil{v/n} \times \ceil{w/n} \simeq \frac{uvw}{n^3}.
-\] 
-Then, we consider each segment portion $c_{i,j,k} = S\cap \B(i,j,k,n)$ and compute its local coordinate representation  $[c]_{i,j,k}\in C_3(n,n,n)$.
-This one is a sparse binary vector of length $n^3$.
-Then, assemble the $M$ representations $c$ of segment portions into a sparse binary matrix $\T{S}$, of dimension $n^d \times M$, with $d=3$.
-Finally, compute a matrix $\T{\textbf{B}}$ of boundary portions of $S$, represented by columns as chain coordinate vectors in $C_2$:
-\[
-\T{\textbf{B}} = [\partial_3(n)]\, \T{S}.
-\]
-where the boundary matrix has dimension $n^d \times dn(n+1)^{d-1}$.
-Of course, the $\T{\textbf{B}}$ sparse matrix has the same column number $M$ of $\T{S}$, because each column contains the boundary representation of the corresponding $S \cap \B(i,j,k,n)$, and the number of matrix rows is the dimension $n^d$ of the linear space $C_2$ of 2-chains.
-
-\subsubsection*{Embedding}
-A final computational step is needed, in order to embed the 2-chains in $\E^3$ space and to assemble the whole resulting surface.
-In particular, we need to compute the \emph{embedding function} $\mu : C_0 \to \E^3$, where $C_0$ is the space of 0-chains, one-to-one with the vertices of the extracted surface.
-The simplest solution is to associate  four 0-cells to each 2-cell of the extracted surface, i.e.~to each non-zero entry in every column of $\T{\textbf{B}}$.
-The $\mu$ function  can be computed by identifying, via  element position in the column, a triple of integer values $0\leq x\leq u$, $0\leq y\leq v$, and $0\leq z\leq w$ for each vertex of the 2-cell.
-The mapping can be implemented using a dictionary, that will store the inverse coordinate transformation used at the beginning, i.e.~the one from linear to Cartesian coords, in order of not duplicating the output vertices.
-
-\subsubsection*{Surface assembling}
-
-All boundary surface subsets $B(i,j,k) = \partial_3 S \cap \B(i,j,k)$, provided by  columns of $\T{B}$, are embedded in the same coordinate space.
-In formal terms, using the standard terminology of LAR scheme:
-\[
-\texttt{Lar}(S) := (\texttt{Geom}(S), \texttt{Top}(S)) = (\texttt{V}, \texttt{CV}),
-\]
-where, with respect to the \emph{chain complex} $C_3\to C_2\to C_1\to C_0$ induced by the input image $Im$ and segment portion $S_{i,j,k}$, we get
-\begin{align}
-\texttt{Geom} &:= \mu(C_0) = \texttt{V},
-\\
-\texttt{Top} &:= C_3(S) = \T{S} \mapsto \texttt{CV}.
-\end{align}
-and
-\begin{align}
-\texttt{Lar}(B_{i,j,k}) &:= (\texttt{Geom}(B_{i,j,k}), \texttt{Top}(B_{i,j,k})) = (\texttt{W}, \texttt{FW}),
-\\
-\texttt{Geom} &:= \mu(C_0(B_{i,j,k})) = \texttt{W} \subset \texttt{V},
-\\
-\texttt{Top} &:= C_2(B_{i,j,k}) = \T{\textbf{B}}_{i,j,k} \mapsto \texttt{FW} \subset \texttt{FV}.
-\end{align}
-
-A translation transformation applied to each vertex subset $\texttt{W}_{i,j,k}$ with translation  vector $\v{t} = [i,j,k]$ will therefore move it in the final space position, so finally giving
-\[
-\texttt{Lar}(\textbf{B}) = \oplus_{i,j,k}\texttt{Lar}(\partial_3 S_{i,j,k})) = \oplus_{i,j,k}(\texttt{W}, \texttt{FW}) .
-\]
-
-\subsection{Block-level parallelism}\label{sec:block-parallelism}
-
-In the computational pipeline introduced in this paper, several steps can be efficiently performed in parallel at image-block level, depending on the embarassingly data parallel nature of the problem.
-In particular, little effort is needed to separate the problem into a number of parallel tasks $S_{i,j,k}$, using multiarray slicing.
-The granularity of parallelism, depending on the block size $n$, is further enforced by the computation of a single boundary matrix $[\partial_d(n)]$ depending on $n$, so that the initial communication cost of broadcasting the matrix to nodes can be carefully controlled, and finely tuned depending on the system architecture.
-The whole approach is appropriate  for SIMD hybrid architectures of CPUs and GPUs, since only the initial block setup of boundary matrix and image slices, as well the final collection of computed surface portions, require inter-process communication.
-
-%\input{implementation}
-
-\section{Julia implementation}\label{sec:julia}
-
-The computer code is implemented in Julia language~\cite{BEKS14} according to the workflow described below, whose stages are parallelized and/or optimized in various ways.
-
-\subsection{Parallel workflow}\label{sec:implementation}
-
-\subsubsection*{Workflow setup}\label{sec:workflow}
-The functions in this preliminary step include:
-\begin{enumerate}
-
-\item input of 3D medical image $\mathcal{I}$ with \emph{shape} $(\ell_1, \ell_2, \ell_3)$, such that: $\mathcal{I} = [\ell_1]\times[\ell_2]\times[\ell_3]$, where $[\ell_k] = [1,2,\ldots,\ell_k]$;
-
-\item analysis of resources available in the computational environment, including operating system, type and number of compute nodes (processors, cores, GPUs), number of cores per node, RAM and caches amounts;
-
-\item depending on the above, best decision for \emph{size} of 3D image block (or brick) $\mathcal{B}$.
-With defaults $\emph{size}=64$,  the \emph{number of bricks} will be $n=\ceil{\ell_1/size}\times\ceil{\ell_2/size}\times\ceil{\ell_3/size}$.
-Hence the default number of bricks is $n=256$, for standard images $512\times 512\times 256$;
-
-\item computation of the Julia's sparse boundary matrix $[\partial_B]$, returning a value of type \texttt{SparseMatrixCSC\{Int8\}\{Int64\}}, where \texttt{Int8} and \texttt{Int64} are the types for values and indices, respectively, stored by Compressed Sparse Column (CSC) format; the storage of $[\partial_B]$ (for $\emph{size}$ = 64) requires about 45 MB;
-
-\item creation of either a local or distributed \texttt{channel} to implement a producer/consumer model of parallel/distributed computation, depending on available resources;
-
-\item distribution of matrix $[\partial_B]$, of default size $45$ MB, to all available nodes/cores (Julia workers) using the Julia macro \texttt{@eveywhere}.
-
-\end{enumerate}
-
-\begin{figure}[tbp]
-\includegraphics[width=\textwidth]{figs/schema_horizontal.pdf} 
-%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
-\caption{Workflow of \textsc{lar-surf} algorithm}
-\label{fig:schema}
-\end{figure}
-
-With $size=64$, the number of non-zeros within the sparse matrix $[\partial(64^3)]$ is $\mathtt{nnz} = 4 792 266$, for a memory size of $9\times \mathtt{nnz}+8\times 262144 \simeq 45$ MB.
-The memory size of the sparse matrix is computed by considering $8+1$ bytes for non-zero element (which are exactly 6 per row), plus 8 bytes per each index of column start.
-
-\subsubsection*{Job enqueuing}\label{sec:job-enq}
-Communication and data synchronization may be managed through \emph{Channels}, which are the FIFO conduits that may provide producer/consumer communication.
-Overall execution time can be improved if other tasks can be run while a task is being executed, or while waiting for an external service/function to complete.
-The single work items of this stage follow:
-\begin{enumerate}
-
-\item extraction, from image arrays of the block views, depending on 3 Cartesian indices;
-
-\item transform  each block \emph{from global} $[\ell_1]\times[\ell_2]\times[\ell_3]$ to \emph{local coordinates} $[n]\times[n]\times[n]$;
-
-\item further transform of each \emph{foreground voxel} $\nu\in\mathcal{S}\subseteq\mathcal{I}$ from Cartesian to linear coordinates, using the suitable Julia's library functions.
-
-\item enqueing the job (as a sequence of integer positions for the non-zeros image elements aligned in a memory buffer of proper \texttt{Channel} type).
-\end{enumerate}
-
-\subsubsection*{3-Chain encoding}\label{chain-coding}
-The interesting part of the \emph{Image} $\mathcal{I}$ is called \emph{Segment} $\mathcal{S}$.
-The goal of the whole \emph{workflow} is to extract a \emph{boundary model} of $\mathcal{S}$ from $\mathcal{I}$.
-The portion of $\mathcal{S}$ inside $\mathcal{B}$, will be denoted as $\mathcal{S}(B)$.
-Each block $\mathcal{B}$ of the 3D image must by converted into the \emph{coordinate representation} of a vector $\nu\in C_3$ in the linear space  of 3-chains.
-
-In coordinates local to $\mathcal{B}$, once fixed an ordering from Cartesian to linear coords, this vector is represented by a \emph{binary array} of length $size^3$.
-With $size=64$, we have $64^3=262144$,  with a non-zero value (i.e.~$1$) for each foreground voxel in $\mathcal{S}(B)$.
-Therefore, the coded segment portion $\mathcal{S}(B)$, results with a space occupancy of about $262$ KB if encoded as a full array (i.e.~including the zero values).
-Whether encoded as a sparse vector, its space occupancy will correspondingly decrease.
-
-\begin{enumerate}
-
-\item each encoding task produces either a full or sparse binary vector. With full or sparse arrays depending by one index, we get either 262 KB or less per job, correspondingly;
-
-\item special format for sparse CSC (Compressed Sparse Column) vectors can be used, since the \emph{value} data for non-zeros does not need storage.
-Hence only a single 1-array of \texttt{Int64} row positions (with total length equal to the number of non-zeros in the block, with $8\times\mathtt{nnz}$ kB storage) is needed;
-
-\item prepare subsequences of such data vectors (non-zero linear row indices), in order to feed efficiently the available processor threads.
-In case of presence of one/more GPUs, a smaller size of the block---and hence of the boundary matrix and the encoded 3-chain vectors---and then much higher vector numbers, are preferable for speed.
-
-\end{enumerate}
-
-\subsubsection*{SpMM Multiplication}\label{SpMM-multiplication}
-According to the current literature~\cite{} it is more convenient to execute SpMV (sparse matrix-vector) multiplications than SpMSpV (sparse matrix-sparse vector) multiplications.
-Since we have 256 such jobs (one multiplication per block) to perform in the standard setting of the algorithm\footnote{
-Size of block $64^3$; size of image $512^2\times 256$.
-}, or more in case of either smaller blocks or image greater than the standard one, this stage must be evidently parallelized and carefully tuned, possibly by using the GPU, if available.
-\begin{enumerate}
-
-\item Various multiplication algorithms are being tested, using several packages for sparse linear algebra and/or custom implementations;
-
-\item the total speed of this stage will strongly depend on the hardware available, on the granularity of blocks, and on the choice between dense/sparse storage of encoded 3-chains;
-
-\item anyway, the compute elements or threads will be fed without solution of continuity in a \emph{dataflow} process.
-This parallel operation is, according to our preliminary experiments, the critical one of the whole workflow, in the sense that any $\Delta T$ (either positive or negative) in this stage will contribute to the total time $T$.
-
-\end{enumerate}
-
-\subsubsection*{2-Chain decoding}\label{sec:two-chain-decoding}
-Each multiplication of $[\partial_B] : C_3 \to C_2$, times a 3-chain $\nu\in C_3$, produces a 2-chain  $\sigma\in C_2$, i.e.~the \emph{coordinate representation} of the \emph{boundary vector} $\sigma\in C_2$.
-The inverse of the coding algorithm is executed in the present stage.
-This process can also be partially superimposed in time with the previous ones, depending on the size of the memory buffers used to feed the CPU cores or the GPUs and get their results.
-Some elementary steps follow:
-
-\begin{enumerate}
-
-\item conversion from position of ones (or non-zeros) in the 2-chain to linear indices of rows;
-
-\item conversion from linear indices to Cartesian indices in coordinates local to the $\mathcal{B}$ block, using the appropriate library functions;
-
-\item conversion from each Cartesian index value to a suitably oriented (i.e.~with proper attitude) geometry quadrilateral (or pair of triangles) in local coordinates.
-
-\end{enumerate}
-
-A Julia's vectorized pipeline dataflow seems the more appropriate implementation model for the job of each worker.
-
-\subsubsection*{Assembling and artifact filtering}\label{sec:artifact-filtering}
-The results of the previous stages can be described as a \emph{collection of sets} of \emph{geometric quadrilaterals (quads)}, each one encoded as an array of quadruples of integer indices, pointing to the linear array of grid vertices associated to the image block $\mathcal{B}$.
-In other words, \emph{all quads} of \textbf{each job} are now given in the \textbf{same} \emph{local coordinates}.
-Besides to put each partial surface $\mathcal{S}(B) = (\texttt{V}_B, \texttt{FV}_\sigma)$ in the global coordinate system of the image, the present stage must eliminate the redundant boundary features possibly generated at the edges of the partial surface $\mathcal{S}(B)$ within each block $\mathcal{B}$ such that $B \cap \mathcal{I} \not= \emptyset$:
-
-\begin{enumerate}
-
-\item translate each array $\mathtt{FV}_\sigma$, of type \texttt{Lar.Cells}, by summing to each vertex index the linearized offset of the Cartesian coordinates $(n,m,p)(B)$ of the $\mathcal{B}$'s \emph{reference vertex}, i.e.~the one with (all) lowest \emph{Cartesian coordinates} within the $\mathcal{B}$ block.
-
-\item remove both instances of \emph{double quads} generated by \texttt{Lar} software at the block boundaries (see Figure~\ref{fig:blockboundary}).
-They are artifacts generated by the decomposition of the whole image into a number of blocks of tractable size.
-
-\item 
-a smart strategy of removal of such artifacts may be used, which does not require any sorting nor searching on the assembled array of quads.
-It will consist in arranging each block with all three dimensions decreased-increased by one, so that each 2-adjacent pair of blocks will be covering each other for a full side extent of blocks of depth one.
-The details of this \emph{artifact filtering} are elucidated in Section~\ref{sec:covering}.
-
-\end{enumerate}
-
-\subsubsection*{Smoothing}\label{sec:smoothing}
-The final smoothing of the generated surfaces cannot be performed block-wise, since this would introduce smoothing artifacts at the block boundaries.
-Anyway, the Taubin smoothing~\cite{} can be performed in parallel, since for each vertex in the final surface (except eventually the ones on the image $\mathcal{I}$ boundaries) it essentially consist in computing a new position as a proper average of its neighborhood vertices, i.e.~by applying a discrete Laplacian operator.
-Some appropriate set of workers may so be assigned the task of generating iteratively a new position for the vertices they take cure of.
-In particular, we have:
-\begin{enumerate}
-
-\item Job enqueuing, by writing sets of integers (global linear indices of vertices) in array buffers of type \texttt{Channel};
-
-\item vectorized computation of proper averages of near vertices;
-
-\item job dequeuing, by recovering finished tasks from a channel and assembling the results into the embedding function $\mathtt{V}: C_0 \to \E^3$, providing an array of type \texttt{Lar.Points} of \texttt{Float64 $\times$ 3}, with vertex coordinates by column.
-\end{enumerate}
-
-\subsection{Performance analysis}\label{sec:analysis}
-
-The size of boundary matrix is critical parameter of LAR-SURF method.
-To determine optimal size of boundary matrix the experiment on artifical data was performed (Fig.
-\ref{fig:bm_size_tesla}).
-Size of experimental data is set to $512\times512\times512$ and it is derived from typical size of Computed Tomography medical images.
-Computation is done on Tesla DGX-1 machine.
-
-\begin{figure}
-\centering
-\includegraphics[width=0.6\textwidth]{figs/bm_size_tesla.pdf} 
-%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
-\caption{Time requirements of LAR-SURF filter used on artificial with different size of boundary matrix}
-\label{fig:bm_size_tesla}
-\end{figure}
-
-To compare time requirements of LAR-SURF with Marching Cubes implemented in Python we performed experiment on Ircadb dataset \cite{ircad}.
-Dataset contain 20 Computed Tomography images (see table \ref{tab:ircad1}) with xy-resolution from 0.56 mm to 0.87 mm and z-resolution from 1.0 mm to 4.0 mm.
-The number of slices is each series varies from 74 to 260 and size of each slice is $512\times512$.
-Dataset contain manually segmented liver, portal vein and other structures.
-We performed surface extraction of liver with Marching Cubes and LAR-SURF.
-Time required for computation can be seen on figure \ref{fig:ircad_comparison}.
-
-Based on t-test with $\alpha=0.99$, $p=\input{input/ircad_pvalue}$ and 
-$s=\input{input/ircad_statistics}$ it can be shown that the mean of time consumed by LAR-SURF is significantly lower from time consumed by Marching Cubes.
-
-\begin{figure}
-\centering
-\includegraphics[width=0.99\textwidth]{figs/ircad_comparison.pdf} 
-%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
-\caption{Time requirements of LAR-SURF filter and Marching Cubes on Ircad dataset. Error bars shows the 95\% confidence interval}
-\label{fig:ircad_comparison}
-\end{figure}
-
-%\input{examples}
-\section{Examples}\label{sec:examples}
-
-\begin{figure}
-\centering
-\includegraphics[width=0.3\textwidth]{figs/liver_01_red_3.png} 
-\vspace{0.05\textwidth}
-\includegraphics[width=0.3\textwidth]{figs/portalvein_01_yellow_3.png} 
-\vspace{0.05\textwidth}
-% \includegraphics[height=0.3\textwidth]{figs/nrn10_100_green.png} 
-\includegraphics[height=0.3\textwidth]{figs/nrn10_100_low_res.png} 
-%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
-\caption{
-Triangulated isosurface of liver parts extracted with LAR-SURF algorithm. 
-On the left image the surface of the liver parenchyma of human. 
-The middle image shows the Portal Vein from the same Ircad dataset \cite{ircad} 
-with resolution $1.6\times0.57\times0.57$ [mm].
-On the right image is the pig liver microvasculature based on corrosion cast prepared by Eberlova
-\cite{eberlova2017use}. The resolution of the Micro-CT data is 4.682 $\mu{}m$.
+\section*{ACKNOWLEDGEMENTS}
+\sloppy{
+This work was supported by Charles University Research Centre program UNCE/MED/006 ``University Center of Clinical and Experimental Liver Surgery'' 
+and Ministry of Education project ITI CZ.02.1.01/0.0/0.0/17\_048/0007280: Application of modern technologies in medicine and industry.
 } 
-\label{fig:example_liver_macro_micro}
-\end{figure}
-
-The implementation of our algorithm is available in LarSurf package. 
-Use of this package can be seen on listings \ref{lst:example1} where the liver segmentation with 2865131 voxels from
-Ircad dataset is used as an input for our surface extraction algorithm. The size of 3D volumetric 
-image is $129 \times 512 \times 512$
-and the voxel resolution is $1.6\times0.57\times0.57$ [mm]. 
-The output surface model created by  182124 triangles and the number of vertices is 90822. The visualization can 
-be seen on left the image of figure \ref{fig:example_liver_macro_micro}. 
-The portal vein surface extraction can be performed with a small change of input path in code.
-It can be seen in the same image in the middle. The The 3D image resolution is the same. 
-The number of input voxels is 103533. 
-The output surface is created by 90822 vertices and 182124 triangles.
-
-\begin{lstlisting}[caption={Get surface from DICOM volumetric data}, label={lst:example1}]
-using Distributed
-using Pio3d  # Read 3D data from DICOM files
-addprocs(3)  # set number of processors
-using LarSurf
-
-LarSurf.lsp_setup([64, 64, 64])  # set block size
-
-# read data from DICOM files
-datap = Pio3d.read3d("3Dircadb1.1/MASKS_DICOM/liver")
-segmentation = datap["data3d"]
-voxelsize_mm = datap["voxelsize_mm"]
-
-# get surface
-V, FV = LarSurf.lsp_get_surface(segmentation, voxelsize_mm)
-FVtri = LarSurf.triangulate_quads(FV)
-
-# do smoothing and save data
-Vs = LarSurf.Smoothing.smoothing_FV_taubin(V,FV,0.5,-0.2,40)
-objlines = LarSurf.Lar.lar2obj(Vs, FVtri, "liver.obj")
-\end{lstlisting}
-
-The right image of figure \ref{fig:example_liver_macro_micro} is surface of a microvasculature of pig liver. 
-Volumetric image is based on Micro-CT data of corrosion casts of pig liver.
-\cite{eberlova2017use}.
-The size of the visualized data is $100\times100\times100$ voxels and size of voxel is 4.682 $\mu{}m$.
-The number of of triangles is 
-544784 and number of vertices is 272826.
-
-% \begin{minted}{python}
-
-% using Pkg
-% \end{minted}
-% \begin{minted}[breaklines,escapeinside=||,mathescape=true, linenos, numbersep=3pt, gobble=2, frame=lines, fontsize=\small, framesep=2mm]{julia}
-% Your awesome julia code here\end{minted}
-
-\begin{figure}
-\centering
-\includegraphics[width=0.4\textwidth]{figs/porta_smoothing_1.png} 
-\vspace{0.05\textwidth}
-\includegraphics[width=0.4\textwidth]{figs/porta_smoothing_2.png} 
-%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
-\caption{Triangulated isosurface of portal vein calculated with LAR-SURF}
-\label{fig:example_porta}
-\end{figure}
-
-%\input{conclusion}
-\section{Conclusion}\label{sec:conclusion}
-
-In this paper we have introduced and discussed a Julia implementation of an algebraic filter to extract from medical 3D images the boundary sourface of a specific image segment, described as a 3-chain of voxels.
-We have shown a good advantage over standard marchin-cubes algorithms.
-Translations from cartesian indices of cells to linearized indices, and the sparse matrix-vector multiplication are the main computational kernels of this approach.
-The current implementation employs Julia's channels for multiprocessing, and can be extended to gain a much greater speed-up using hybrid architectures mixing  CPUs and GPUs of last generation.
+The research was also supported by the project LO 1506 of the Czech Ministry of Education, Youth and Sports.
+% The authors appreciate the access to computing and storage facilities owned by parties and projects contributing to the National Grid Infrastructure MetaCentrum  provided under the program ``Projects of Large Research, Development, and Innovations Infrastructures'' (CESNET LM2015042).
 
 \appendix
-\section{Appendix}
-\subsection{Symbol list}
+\input{appendix}
 
-\begin{description}
-\item[$\mathcal{I}$]  three-dimensional medical image
-\item[$\ell_1, \ell_2, \ell_3$]  dimensions of image
-\item[$\mathcal{S}$]  segment: a subset of voxels from image segmentation
-\item[$\mathcal{B}$] 3D image block
-\item[$size$] lateral dimension of cubic block $\mathcal{B}$
-\item[$n$]	number of blocks $\mathcal{B}$ (jobs) in $\mathcal{I}$
-\item[[$\partial_\mathcal{B}$] ]  boundary matrix for block $\mathcal{B}$
-\item[$C_p$] linear (vector) space of $p$-chains
-\item[$\nu\in C_p$] $p$-chain
-\item[[$\nu$]] coordinate representation (binary vector) of  $\nu$
-\item[$\E^3$] Euclidean 3-space
-\end{description}
+\section*{ORCID}
+\orcid{Miroslav Jirik}{0000-0002-8002-2079}
+\orcid{Alberto Paoluzzi}{0000-0002-3958-8089}
 
-\subsection{Definitions}
-\begin{description}
-
-\item[Boundary model] Closed manifold surface of the boundary of a solid model
-\item[CSC]  Compressed Sparse Column format for sparse matrices
-\item[Global coordinates]  Integer linear coordinates of $\mathcal{I}$
-\item[Local coordinates] Integer linear coordinates of $\mathcal{B}$
-\item[Cartesian coordinates] Integer triples $(i,j,k)$ one-to-one with voxels
-\item[Voxels] Individual elements in 3D image (3-cells)
-\item[$p$-chain] Formal linear combination of $p$-cells with coefficients in $\{0,1\}$
-\item[Coord. repr.]  Binary vector (for $p$-chains) or binary matrix (for chain operators)
-\item[Quad]	Geometric quadrilateral; convex polygon with four vertices
-\item[Foreground voxel] Individual element of a segment $\mathcal{S}$
-\item[Segment]	Subset of voxels resulting from image segmentation
-\end{description}
-
-\subsection{3D-Ircadb Dataset}
-
-For perfomance analysis the public dataset from Research Institute against Digestive Cancer (IRCAD) \cite{ircadb} was used. The table \ref{tab:ircad1} and \ref{tab:ircad2} describe the dataset.
-
-\begin{table}
-\input{input/dataset_ircad}
-\label{tab:ircad1}
-\end{table}
-
-\begin{table}
-\input{input/dataset_ircad_desc}
-\label{tab:ircad2}
-\end{table}
-
-\clearpage{}
-\bibliographystyle{alpha}
+% \clearpage{}
+\referenceSection
+\bibliographystyle{CADA}
 % \bibliography{library,TSAS-2019,references}
@@
 
+\bigskip
 \end{document}

### 2020-04-25 src/appendix.tex

--- 2019-06-30 src/main.tex
@@
-\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
-\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
-\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
-%\geometry{landscape}                		% Activate for rotated page geometry
-%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
-\usepackage{graphicx}				% Use pdf, png, jpg, or eps§ with pdflatex; use eps in DVI mode
-								% TeX will automatically convert eps --> pdf in pdflatex		
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{hyperref}
-
-%SetFonts
-\usepackage[utf8x]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage{lmodern}
-
-\input{macros}
-
-%SetFonts
-
-\title{Algebraic filtering of surfaces from medical images with Julia}
-\author{Miroslav Jirik and Alberto Paoluzzi}
-\date{}							% Activate to display a given date or no date
-
-\begin{document}
-\maketitle
-
-\begin{abstract}
-In this paper we introduce a novel algebraic \textsc{lar-surf} filter, well founded on algebraic topology methods, to extract and smooth the boundary surface of any subset of voxels arising from  segmentation of a 3D medical image.
-The input is defined as a \emph{chain}, i.e.~as a vector from a linear space of 3-chains, represented in coordinates as a sparse binary vector.
-The output is produced by a linear mapping between spaces of 3- and 2-chains through the boundary operator $\partial_3:C_3 \to C_2$.
-In particular, when the input set of voxels is either not (4-)connected, or contains one or more empty regions inside, \textsc{lar-surf} generates a non connected set of closed surfaces, i.e.~a set of 2-cycles---using the language of algebraic topology.
-The only data structures used by this approach are sparse arrays with one or two indices, i.e.~sparse vectors and matrices.
-This work is based on LAR (Linear Algebraic Representation) methods, and is implemented in Julia language, natively supporting parallel computing on hybrid architectures.
-\end{abstract}
-
-\tableofcontents
-
-%\section{}
-%\subsection{}
-
-\include{intro}
-\include{background}
-\include{design}
-\include{implementation}
-\include{examples}
-\include{conclusion}
-
-\appendix
 \section{Appendix}
@@
 \item[$\mathcal{I}$]  three-dimensional medical image
-\\[-7mm]
 \item[$\ell_1, \ell_2, \ell_3$]  dimensions of image
-\\[-7mm]
 \item[$\mathcal{S}$]  segment: a subset of voxels from image segmentation
-\\[-7mm]
-\item[$\mathcal{B}$] 3D image block 
-\\[-7mm]
-\item[$size$] lateral dimension of cubic block $\mathcal{B}$ 
-\\[-7mm]
-\item[$n$]	number of blocks $\mathcal{B}$ (jobs) in $\mathcal{I}$
-\\[-7mm]
-\item[[$\partial_\mathcal{B}$] ]  boundary matrix for block $\mathcal{B}$
-\\[-7mm]
+\item[$\mathcal{B}$] 3D image brick
+\item[$size$] lateral dimension of cubic brick $\mathcal{B}$
+\item[$n$]	number of bricks $\mathcal{B}$ (jobs) in $\mathcal{I}$
+\item[[$\partial_\mathcal{B}$] ]  boundary matrix for brick $\mathcal{B}$
 \item[$C_p$] linear (vector) space of $p$-chains
-\\[-7mm]
 \item[$\nu\in C_p$] $p$-chain
-\\[-7mm]
 \item[[$\nu$]] coordinate representation (binary vector) of  $\nu$
-\\[-7mm]
 \item[$\E^3$] Euclidean 3-space
@@
 \item[Boundary model] Closed manifold surface of the boundary of a solid model
-\\[-7mm]
 \item[CSC]  Compressed Sparse Column format for sparse matrices
-\\[-7mm]
 \item[Global coordinates]  Integer linear coordinates of $\mathcal{I}$
-\\[-7mm]
 \item[Local coordinates] Integer linear coordinates of $\mathcal{B}$
-\\[-7mm]
 \item[Cartesian coordinates] Integer triples $(i,j,k)$ one-to-one with voxels
-\\[-7mm]
 \item[Voxels] Individual elements in 3D image (3-cells)
-\\[-7mm]
 \item[$p$-chain] Formal linear combination of $p$-cells with coefficients in $\{0,1\}$
-\\[-7mm]
 \item[Coord. repr.]  Binary vector (for $p$-chains) or binary matrix (for chain operators)
-\\[-7mm]
 \item[Quad]	Geometric quadrilateral; convex polygon with four vertices
-\\[-7mm]
 \item[Foreground voxel] Individual element of a segment $\mathcal{S}$
-\\[-7mm]
 \item[Segment]	Subset of voxels resulting from image segmentation
@@
 
-% Bibliography
-\bibliographystyle{acm}
-%\bibliography{TOMS2017.bib,proposal.bib,cadj2017.bib,bib.bib,schemes.bib}
-\bibliography{TSAS-2019.bib}
+\subsection{3D-Ircadb Dataset}
 
-\end{document}
+For perfomance analysis the public dataset from Research Institute against Digestive Cancer (IRCAD) \cite{ircadb} was used. 
+The table 
+% \ref{tab:ircad1} and 
+\ref{tab:ircad2} describe the dataset.
+
+% \begin{table}
+% \input{input/dataset_ircad}
+% \caption{Ircad dataset description, \cite{ircadb}}
+% \label{tab:ircad1}
+% \end{table}
+
+\begin{table}[h!]
+\centering
+\input{input/dataset_ircad_desc}
+\caption{Ircad dataset description \cite{ircadb}. It contains 20 Computed Tomography images of abdomen with 
+manually segmented tissues.}
+\label{tab:ircad2}
+\end{table}

### 2020-04-25 src/design.tex

@@
 
-\subsection{Block decomposition}\label{sec:bbbb}
+\subsection{Block decomposition}\label{sec:block-decomposition}
 
-We assume that medical devices produce 3D images with lateral dimensions that are integer multiples of some powers of two, like 128, 256, 512, etc.
-Any cuboidal portion of image is completely determined by the Cartesian indices of its voxels of lowest and highest indices, and extracted by multidimensional array \emph{slicing} as $Image([\ell_x$:$h_x, \ell_y$:$h_y, \ell_z$:$h_z])$.
+Let us assume that medical devices produce 3D images with lateral dimensions that are integer multiples of some powers of two, like 128, 256, 512, etc.
+Any cuboidal portion of the image is completely determined by the Cartesian indices of its voxels of lowest and highest indices and extracted by multidimensional array \emph{slicing} as $image([\ell_x : h_x, \ell_y : h_y, \ell_z : h_z])$.
 
-For the sake of simplicity, we assume a common size on the three image axes, and the corresponding image portion $\B$, called \emph{block}, as a function of its element of the  lowest  block  coordinates $i,j,k\in [1:n]$ and of block dimension $n\in\N$:
+For the sake of simplicity, we assume a common size on the three image axes, and the corresponding image portion $\B$, called \emph{brick}, as a function of its element of the  lowest  brick coordinates $i,j,k\in [1:n]$ and the brick lateral size $n\in\N$:
 \[
-\B(i,j,k,n) := Image([in:in+n, jn:jn+n, kn:kn+n]) 
+\B(i,j,k,n) := image([in:in+n, jn:jn+n, kn:kn+n]) 
 \]
@@
 
-\paragraph{Chain coordinates }\label{sec:bbbb} 
+\subsubsection*{Chain coordinates }\label{sec:chain-coords}
 We are going to treat each image block independently from each other.
-Hence we map each image subspace $\B(i,j,n)$ to the linear \emph{chain} space $C_2$ of dimension $n\times n$, using coordinate vectors $c_{h,k}\in \\B^{n\times n} := \{0,1\}^{n\times n}$, where the basis element $c = c_{h,k} \in C_2$ is mapped via Cartesian-to-linear map to the binary vector
+Hence we map each image block $\B(i,j,k,n)$ to the linear \emph{chain} space $C_2$ of dimension $n\times n\times n$, using coordinate vectors $c\in \mathbf{2}^{n^d} := \{0,1\}^{n^d}$, where the basis element $c \in C_2$ is mapped via Cartesian-to-linear map to the binary vector
 \[
@@
 
-\paragraph{Boundary operator }\label{sec:bbbb} For a fixed block size $n$, the boundary operator $\partial_d : C_d\to C_{d-1}$, with $d\in\{2,3\}$, will be constructed once and for all using the algorithm given in~\cite{}, and inlined in the generated boundary extraction code.
+\subsubsection*{Boundary operator }\label{sec:boundary-operator}
+For a fixed block size $n$, the boundary operator $\partial_d : C_d\to C_{d-1}$, with $d\in\{2,3\}$, will be constructed once and for all using the algorithm given in~\cite{}, and inlined in the generated boundary extraction code.
 
@@
 
-\paragraph{Sparsity and size of boundary matrix }\label{sec:bbbb}
+\subsubsection*{Sparsity and size of boundary matrix }\label{sec:bm-size}
 
@@
 
+% TODO fix the name
 \begin{figure}[htbp] %  figure placement: here, top, bottom, or page
@@
    \includegraphics[width=0.4\linewidth]{figs/grid.pdf} 
-   \caption{example caption}
-   \label{fig:example}
+   \caption{Faces on the brick boundary}
+   \label{fig:blockboundary}
 \end{figure}
 
-\subsection{Block boundary mapping}\label{sec:bbbb}
+\subsection{Block boundary mapping}\label{sec:block-mapping}
 
 Here we refer directly to the 3D case.
-Let us call \emph{segment} the bulk content $S$ of interest within the input 3D image of size $(u,v,w)$. We aim to compute the segment boundary $\partial_3 S$. 
-First we set the size $n$ of the block, in order to decompose the input $Image(u,v,w)$ into a fair number 
+Let us call \emph{segment} the bulk content $S$ of interest within the input 3D image of size $(u,v,w)$. Our aim is to compute the segment boundary $\partial_3 S$. 
+First we set the size $n$ of the block, in order to decompose the input $Image(u,v,w)$ into a fair number of blocks
 \[
-M = \ceil{u/n} \times \ceil{v/n} \times \ceil{w/n} \simeq \frac{uvw}{n^3}
-\] of blocks. 
-Then we consider each image portion $c_{i,j,k} = S\cap \B(i,j,k,n)$ and compute its (binary) coordinate representation  $[c]_{i,j,k}\in C_3(n,n,n)$.
+M = \ceil{u/n} \times \ceil{v/n} \times \ceil{w/n} \simeq \frac{uvw}{n^3}.
+\] 
+Then, we consider each segment portion $c_{i,j,k} = S\cap \B(i,j,k,n)$ and compute its local coordinate representation  $[c]_{i,j,k}\in C_3(n,n,n)$.
 This one is a sparse binary vector of length $n^3$.
-Then assemble the $M$ representations $c$ of segment portions into a sparse binary matrix $\T{S}$, of dimension $n^d \times M$.
-Finally compute a matrix $\T{\B}$ of boundary portions of $S$, represented by columns as chain coordinate vectors in $C_2$:
+Then, assemble the $M$ representations $c$ of segment portions into a sparse binary matrix $\T{S}$, of dimension $n^d \times M$, with $d=3$.
+Finally, compute a matrix $\T{\textbf{B}}$ of boundary portions of $S$, represented by columns as chain coordinate vectors in $C_2$:
 \[
-\T{\B} = [\partial_3(n)]\, \T{S}.
+\T{\textbf{B}} = [\partial_3(n)]\, \T{S}.
 \]
-where the boundary matrix has dimension $d\,n\,(1+n)^{d-1} \times n^d$.
-Of course, the $\T{\B}$ sparse matrix has the same column number $M$ of $\T{S}$, because each column contains the boundary representation of the corresponding $S\cap Box_{i,j,k}{i,j,k}$, and the number of rows of the operator, equal to the dimension of the linear space $C_2$.
+where the boundary matrix has dimension $n^d \times dn(n+1)^{d-1}$.
+Of course, the $\T{\textbf{B}}$ sparse matrix has the same column number $M$ of $\T{S}$, because each column contains the boundary representation of the corresponding $S \cap \B(i,j,k,n)$, and the number of matrix rows is the dimension $n^d$ of the linear space $C_2$ of 2-chains.
 
-\paragraph{Embedding}
-A final computational step is needed, in order to embed the 2-chains in Euclidean space $\E^3$ and to assemble the whole resulting surface.
-In particular, we need to compute the \emph{embedding function} $\mu : C_0 \to \E^3$, where $C_0$ is the space of 0-chains, one-to-one corresponding to the vertices of the extracted surface.
-The simplest solution is to associate  four 0-cells to each 2-cell of the extracted surface, i.e.~to each non-zero entry in every column of $\T{\B}$.
+\subsubsection*{Embedding}
+A final computational step is needed, in order to embed the 2-chains in $\E^3$ space and to assemble the whole resulting surface.
+In particular, we need to compute the \emph{embedding function} $\mu : C_0 \to \E^3$, where $C_0$ is the space of 0-chains, one-to-one with the vertices of the extracted surface.
+The simplest solution is to associate  four 0-cells to each 2-cell of the extracted surface, i.e.~to each non-zero entry in every column of $\T{\textbf{B}}$.
 The $\mu$ function  can be computed by identifying, via  element position in the column, a triple of integer values $0\leq x\leq u$, $0\leq y\leq v$, and $0\leq z\leq w$ for each vertex of the 2-cell.
@@
 \[
-\texttt{Lar}(\B) = \oplus_{i,j,k}\texttt{Lar}(\partial_3 S_{i,j,k})) = \oplus_{i,j,k}(\texttt{W}, \texttt{FW}) .
+\texttt{Lar}(\textbf{B}) = \oplus_{i,j,k}\texttt{Lar}(\partial_3 S_{i,j,k})) = \oplus_{i,j,k}(\texttt{W}, \texttt{FW}) .
 \]
 
-\subsection{Block-level parallelism}\label{sec:bbbb}
-
-In the computational pipeline introduced in this paper, several steps can be efficiently performed in parallel at image-block level, depending on the embarassingly data parallel nature of the problem.
-In particular, little effort is needed to separate the problem into a number of parallel tasks $S_{i,j,k}$, using multiarray slicing.
-The granularity of parallelism, depending on the block size $n$, is further enforced by the computation of a single boundary matrix $[\partial_d(n)]$, in turn depending on $n$, so that the initial communication cost of broadcasting the matrix to nodes can be carefully controlled, and finely tuned depending on the system architecture.
-The whole approach is appropriate  for SIMD hybrid architectures of CPUs and GPUs, since only the initial block setup of boundary matrix and image slices, as well the final collection of computed surface portions, require inter-process communication.
+\subsection{Block-level parallelism}\label{sec:block-parallelism}
+In the computational pipeline introduced in this paper, several steps can be efficiently performed in parallel at the image-block level, depending on the embarrassingly data-parallel nature of the problem.
+In particular, little effort is needed to separate the problem into a several of parallel tasks $S_{i,j,k}$, using multiarray slicing.
+The granularity of parallelism, depending on the block size $n$, is further enforced by the computation of a single boundary matrix $[\partial_d(n)]$ depending on $n$, so that the initial communication cost of broadcasting the matrix to nodes can be carefully controlled, and finely tuned depending on the system architecture.
+The whole approach is appropriate  for SIMD (Single Instruction, Multiple Data) hybrid architectures of CPUs and GPUs, since only the initial block setup of boundary matrix and image slices, as well the final collection of computed surface portions, require inter-process communication.

### 2020-04-25 src/examples.tex
new examples added and text with statistic

--- 2019-06-23 src/examples.tex
@@
 
-The 
-For an experiments the we used the dataset Ircad1b.
+% \includegraphics[height=0.3\textwidth]{figs/nrn10_100_green.png} 
+%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
+\begin{figure}
+\centering
+% \includegraphics[width=0.3\textwidth]{figs/ircad01}
+\includegraphics[width=0.4\textwidth]{figs/liver_01_red_3.png} 
+% \vspace{0.05\textwidth}
+\includegraphics[width=0.4\textwidth]{figs/portalvein_01_yellow_3.png} 
+% \vspace{0.01\textwidth}
+\includegraphics[height=0.4\textwidth]{src/figs/nrn10_100_magenta_high_res.png} 
+% \includegraphics[height=0.4\textwidth]{figs/nrn10_100_low_res.png} 
+% \vspace{0.05\textwidth}
+\includegraphics[width=0.4\textwidth]{figs/porta_smoothing_2.png} 
+\caption{
+Triangulated isosurface of liver parts extracted with LAR-SURF algorithm. 
+On the upper left image the surface of the liver parenchyma of human can be seen. 
+Both right images show the Portal Vein from the same Ircad dataset \cite{ircad} 
+with resolution $1.6\times0.57\times0.57$ [mm].
+On the left bottom image is the pig liver microvasculature based on corrosion cast prepared by Eberlova
+\cite{eberlova2017use}. The resolution of the Micro-CT data is 4.682 $\mu{}m$.
+} \label{fig:example_liver_macro_micro}
+\end{figure}
 
-\begin{table}
-\input{input/dataset_ircad}
-\end{table}
-\begin{table}
-\input{input/dataset_ircad_desc}
-\end{table}
+% \begin{figure}
+% \centering
+% \includegraphics[width=0.4\textwidth]{figs/porta_smoothing_1.png} 
+% \vspace{0.05\textwidth}
+% \includegraphics[width=0.4\textwidth]{figs/porta_smoothing_2.png} 
+% %\includegraphics[scale=1]{input/ircad_comparison.pdf} 
+% \caption{Triangulated isosurface of portal vein calculated with LAR-SURF}
+% \label{fig:example_porta}
+% \end{figure}
+
+The implementation of our algorithm is available in LarSurf package. 
+Use of this package can be seen on listings \ref{lst:example1} where the liver segmentation with 2865131 voxels from
+Ircad dataset is used as an input for our surface extraction algorithm. The size of 3D volumetric 
+image is $129 \times 512 \times 512$
+and the voxel resolution is $1.6\times0.57\times0.57$ [mm]. 
+The output surface model created by  182124 triangles and the number of vertices is 90822. The visualization can 
+be seen on figure \ref{fig:example_liver_macro_micro}. 
+The portal vein surface extraction can be performed with a small change of input path in code.
+The The 3D image resolution is the same. 
+It can be seen in the same figure. 
+The number of input voxels is 103533. 
+The output surface is created by 90822 vertices and 182124 triangles.
+
+\begin{lstlisting}[caption={Get surface from DICOM volumetric data}, label={lst:example1}]
+using Distributed
+using Pio3d  # Read 3D data from DICOM files
+addprocs(3)  # set number of processors
+using LarSurf
+
+LarSurf.lsp_setup([64, 64, 64])  # set block size
+
+# read data from DICOM files
+datap = Pio3d.read3d("3Dircadb1.1/MASKS_DICOM/liver")
+segmentation = datap["data3d"]
+voxelsize_mm = datap["voxelsize_mm"]
+
+# get surface
+V, FV = LarSurf.lsp_get_surface(segmentation, voxelsize_mm)
+FVtri = LarSurf.triangulate_quads(FV)
+
+# do smoothing and save data
+Vs = LarSurf.Smoothing.smoothing_FV_taubin(V,FV,0.5,-0.2,40)
+objlines = LarSurf.Lar.lar2obj(Vs, FVtri, "liver.obj")
+\end{lstlisting}
+
+The right image of figure \ref{fig:example_liver_macro_micro} is surface of a microvasculature of pig liver. 
+Volumetric image is based on Micro-CT data of corrosion casts of pig liver.
+\cite{eberlova2017use}.
+The size of the visualized data is $100\times100\times100$ voxels and size of voxel is 4.682 $\mu{}m$.
+The number of of triangles is 
+544784 and number of vertices is 272826.
+
+% \begin{minted}{python}
+
+% using Pkg
+% \end{minted}
+% \begin{minted}[breaklines,escapeinside=||,mathescape=true, linenos, numbersep=3pt, gobble=2, frame=lines, fontsize=\small, framesep=2mm]{julia}
+% Your awesome julia code here\end{minted}

### 2020-04-26 src/conclusion.tex

\section{Conclusion}\label{sec:conclusion}

% In this paper we have introduced and discussed a Julia implementation of an algebraic filter to extract from medical 3D images the boundary sourface of a specific image segment, described as a 3-chain of voxels. We have shown a good advantage over standard marchin-cubes algorithms. Translations from cartesian indices of cells to linearized indices, and the sparse matrix-vector multiplication are the main computational kernels of this approach. The current implementation employs Julia's channels for multiprocessing, and can be extended to gain a much greater speed-up using hybrid architectures mixing  CPUs and GPUs of last generation.

% TODO from abstract
We introduced a Julia implementation of an algebraic filter to extract from 3D medical images the
boundary surface of some specific image segment, described as a 3-chain of voxels. Translations from
Cartesian indices of cells to linearized indices, the computation of the sparse boundary matrices, and the
sparse matrix-vector multiplication are the main computational kernels of this approach. We may show
a good speed-up over marching-cubes algorithms. The existing implementation employs Julia's channels
for multiprocessing. Currently, the computational pipeline is being strongly improved to gain a greater
speed-up using native Julia implementation \texttt{CUDA.jl} of Nvidia programming platform 
% TODO cite Besard2017 [7]
, and the Julia's
\texttt{SuiteSparseGraphBLAS.jl} framework 
% TODO cite BULUK 2017 [8] 
for graph algorithms with the language of linear algebra. In
particular, we are extending its use pattern in order to work with general cellular complexes.

### 2020-04-26 src/intro_old.tex

\section{Introduction}\label{sec:intro}

Isosurface extraction produces geometric model of surface 
 from volumetric data is important in many aplications. It is often used for indirect visualization of the medical data or for flow modeling \cite{Rohan2018a}.

The most popular algorithm used for surface extraction is probably Marching Cubes (MC). The algorithm was described by Lorentsen and Cline \cite{Lorensen1987} in 1987. Survey of Marching Cubes algorthm has been published in 2006 \cite{Newman2006}. The algorithm is based on considering the cube defining volume. Each corner vertex of the cube is related to input volumetric data. MC traverse the data and constructs the surface by using lookup table of different triangular faces depending on different patterns of the cube.  Main disadvantages of this method are time requirements, ambiquity and holes generation. Some of them were discoverd shortly after the algorithm was introduced. 
Marching Cubes. In 1991 Nielson and Hamman described Asymptotic Decider to solve the ambiguity problem on the faces of the cube.  Natarajan noted that the ambiguity problem also occurs in cubes \cite{Natarajan1994}. In 1995 Chernyaev extended the number of cases to 33 \cite{chernyaev1995marching}. More recently the algorithm was updated by Custodio to   enhance the quality of triangulation \cite{Custodio2019}.

The alternative methods have been deveoloped including method for surface extraction using particle attraction system was described by Crossno and Angel in \cite{Crossno2002} and method processing on a graph that tracks cell face adjacencies is described in \cite{Lachaud2000}. The parallel algorithms for surface extraction are discussed in \cite{Bajaj2004}.

### 2020-04-26 src/examples.tex
Updates from Overleaf

@@
 \centering
-% \includegraphics[width=0.3\textwidth]{figs/ircad01}
-\includegraphics[width=0.4\textwidth]{figs/liver_01_red_3.png} 
-% \vspace{0.05\textwidth}
-\includegraphics[width=0.4\textwidth]{figs/portalvein_01_yellow_3.png} 
-% \vspace{0.01\textwidth}
-\includegraphics[height=0.4\textwidth]{src/figs/nrn10_100_magenta_high_res.png} 
-% \includegraphics[height=0.4\textwidth]{figs/nrn10_100_low_res.png} 
-% \vspace{0.05\textwidth}
-\includegraphics[width=0.4\textwidth]{figs/porta_smoothing_2.png} 
+\includegraphics[width=0.45\textwidth]{src/figs/ircad01_segmentation_65.png}
+\includegraphics[width=0.45\textwidth]{src/figs/ircad01_liver_tricolore_01.png}
+\includegraphics[width=0.45\textwidth]{src/figs/ircad01_porta_blue_01.png}
+\includegraphics[width=0.45\textwidth]{src/figs/nrn10_200_pink_02.png}
+% \includegraphics[width=0.4\textwidth]{figs/liver_01_red_3.png} 
+% % \vspace{0.05\textwidth}
+% \includegraphics[width=0.4\textwidth]{figs/portalvein_01_yellow_3.png} 
+% % \vspace{0.01\textwidth}
+% \includegraphics[height=0.4\textwidth]{src/figs/nrn10_100_magenta_high_res.png} 
+% % \includegraphics[height=0.4\textwidth]{figs/nrn10_100_low_res.png} 
+% % \vspace{0.05\textwidth}
+% \includegraphics[width=0.4\textwidth]{figs/porta_smoothing_2.png} 
 \caption{
-Triangulated isosurface of liver parts extracted with LAR-SURF algorithm. 
-On the upper left image the surface of the liver parenchyma of human can be seen. 
-Both right images show the Portal Vein from the same Ircad dataset \cite{ircad} 
-with resolution $1.6\times0.57\times0.57$ [mm].
-On the left bottom image is the pig liver microvasculature based on corrosion cast prepared by Eberlova
-\cite{eberlova2017use}. The resolution of the Micro-CT data is 4.682 $\mu{}m$.
+Liver structures. The upper left image shows one slice  with segmented organs from  the Ircad dataset \cite{ircad}.
+The other images show the triangulated isosurfaces of the macroscopic and microscopic strucures of the liver extracted with \textsc{lar-surf} algorithm. 
+On the upper right image can be seen the wire frame model of the human liver along with portal vein and hepatic artery. 
+% Both right images show the Portal Vein from the same Ircad dataset \cite{ircad} 
+The left bottom image shows the detail of the portal vein  
+with resolution $1.6\times0.57\times0.57$ $[mm]$.
+On the bottom right image is shown the microvasculature of a pig liver based on corrosion cast prepared by Eberlova
+\cite{eberlova2017use}. The size of the speciment is 0.936 $[mm]$ along each axis and the resolution of the Micro-CT data is 4.682 $\mu{}m$.
 } \label{fig:example_liver_macro_micro}

### 2020-04-26 src/design.tex

@@
 \subsubsection*{Boundary operator }\label{sec:boundary-operator}
-For a fixed block size $n$, the boundary operator $\partial_d : C_d\to C_{d-1}$, with $d\in\{2,3\}$, will be constructed once and for all using the algorithm given in~\cite{}, and inlined in the generated boundary extraction code.
+For a fixed block size $n$, the boundary operator $\partial_d : C_d\to C_{d-1}$, with $d\in\{2,3\}$, will be constructed once and for all using the algorithm given in~
+\cite{TSAS}, 
+% \cite{DiCarlo2014}, 
+and inlined in the generated boundary extraction code.
 

### 2020-04-26 src/implementation.tex

@@
-\section{Julia implementation}\label{sec:implementation}
+\section{Julia implementation}\label{sec:julia}
 
-The computer code is implemented in Julia language~\cite{} according to the workflow described below, whose stages are parallelized and/or optimized in various ways.
+The computer code is implemented in Julia language~\cite{BEKS14} according to the workflow described below, whose stages are parallelized and/or optimized in various ways.
 
@@
 
-\paragraph{Workflow setup} The functions in this preliminary step include:
+\subsubsection*{Workflow setup}\label{sec:workflow}
+The functions in this preliminary step include:
 \begin{enumerate}
 
-\item input of 3D medical image $\mathcal{I}$ dimensions $\ell_1, \ell_2, \ell_3$, such that: $\mathcal{I} = [\ell_1]\times[\ell_2]\times[\ell_3]$, where $[\ell_k] = (1,2,\ldots,\ell_k)$;
+\item input of 3D medical image $\mathcal{I}$ with \emph{shape} $(\ell_1, \ell_2, \ell_3)$, such that: $\mathcal{I} = [\ell_1]\times[\ell_2]\times[\ell_3]$, where $[\ell_k] = [1,2,\ldots,\ell_k]$;
 
@@
 
-\item depending on the above, best decision for $\emph{size}$ of 3D image blocks $\mathcal{B}$ (bricks); defaults to: $\emph{size}=64$; hence the number of bricks will be $n=\ceil{\ell_1/size}\times\ceil{\ell_2/size}\times\ceil{\ell_3/size}$.
-Hence default value: $n=256$, for images $\mathcal{I} = 512\times 512\times 256$;
+\item depending on the above, best decision for \emph{size} of 3D image block (or brick) $\mathcal{B}$.
+With defaults $\emph{size}=64$,  the \emph{number of bricks} will be $n=\ceil{\ell_1/size}\times\ceil{\ell_2/size}\times\ceil{\ell_3/size}$.
+Hence the default number of bricks is $n=256$, for standard images $512\times 512\times 256$;
 
-\item computation of a Julia's sparse boundary matrix $[\partial_B]$, returning a value of type \texttt{SparseMatrixCSC\{Int8\}\{Int64\}}, where \texttt{Int8} and \texttt{Int64} are the types for values and indices, respectively, stored by Compressed Sparse Column (CSC) format; the average $[\partial_B]$ (for $\emph{size}$ = 64) is about 45 MB;
+\sloppy{
+\item computation of the Julia's sparse boundary matrix $[\partial_B]$, returning a value of type \texttt{SparseMatrixCSC\{Int8\}\{Int64\}}, where \texttt{Int8} and \texttt{Int64} are the types for values and indices, respectively, stored by Compressed Sparse Column (CSC) format; the storage of $[\partial_B]$ (for $\emph{size}$ = 64) requires about 45 MB;
+}
 
-\item creation of either a local or distributed channel to implement a producer/consumer model of parallel/distributed computation, depending on available resources;
+\item creation of either a local or distributed \texttt{channel} to implement a producer/consumer model of parallel/distributed computation, depending on available resources;
 
-\item distribution of matrix $[\partial_B]$, of default size $45$ MB, to all available nodes/cores (workers) using the macro \texttt{@eveywhere}.
+\item distribution of matrix $[\partial_B]$, of default size $45$ MB, to all available nodes/cores (Julia workers) using the Julia macro \texttt{@eveywhere}.
 
@@
 
-The memory occupancy of the sparse matrix is computed by considering $8+1$ bytes for non-zero element (which are exactly 6 per row), plus 8 bytes per each index of column start.  
-With $size=64$, the number of non-zeros within the sparse matrix $[\partial_{64^3}]$ is $\mathtt{nnz} = 4 792 266$, for a storage size of $9\times \mathtt{nnz}+8\times 262144 \simeq 45$ MB.
-
-\begin{figure}[htbp]
-\includegraphics[width=0.99\textwidth]{figs/schema_horizontal.pdf} 
+\begin{figure}[tbp]
+\includegraphics[width=\textwidth]{figs/schema_horizontal.pdf} 
 %\includegraphics[scale=1]{input/ircad_comparison.pdf} 
-\caption{Workflow of LAR-SURF algorithm}
+\caption{Workflow of \textsc{lar-surf} algorithm}
 \label{fig:schema}
@@
 
-With $size=64$, the number of non-zeros within the sparse matrix $[\partial_{64^3}]$ is $\mathtt{nnz} = 4 792 266$, for a memory size of $9\times \mathtt{nnz}+8\times 262144 \simeq 45$ MB.
+With $size=64$, the number of non-zeros within the sparse matrix $[\partial(64^3)]$ is $\mathtt{nnz} = 4 792 266$, for a memory size of $9\times \mathtt{nnz}+8\times 262144 \simeq 45$ MB.
 The memory size of the sparse matrix is computed by considering $8+1$ bytes for non-zero element (which are exactly 6 per row), plus 8 bytes per each index of column start.
 
-\paragraph{Job enqueuing} Communication and data synchronization may be managed through \emph{Channels}, which are the FIFO conduits that may provide producer/consumer communication.
+\subsubsection*{Job enqueuing}
+% \subsubsection{Job enqueuing}
+\label{sec:job-enq}
+Communication and data synchronization may be managed through \emph{Channels}, which are the FIFO conduits that may provide producer/consumer communication.
 Overall execution time can be improved if other tasks can be run while a task is being executed, or while waiting for an external service/function to complete.
@@
 
-\paragraph{3-Chain encoding} 
+\subsubsection*{3-Chain encoding}
+\label{chain-coding}
 The interesting part of the \emph{Image} $\mathcal{I}$ is called \emph{Segment} $\mathcal{S}$.
@@
 
-\paragraph{SpMM Multiplication} 
+\subsubsection*{SpMM Multiplication}\label{SpMM-multiplication}
 According to the current literature~\cite{} it is more convenient to execute SpMV (sparse matrix-vector) multiplications than SpMSpV (sparse matrix-sparse vector) multiplications.
@@
 
-\paragraph{2-Chain decoding} 
+\subsubsection*{2-Chain decoding}\label{sec:two-chain-decoding}
 Each multiplication of $[\partial_B] : C_3 \to C_2$, times a 3-chain $\nu\in C_3$, produces a 2-chain  $\sigma\in C_2$, i.e.~the \emph{coordinate representation} of the \emph{boundary vector} $\sigma\in C_2$.
@@
 
-\paragraph{Assembling and artifact filtering} 
+\subsubsection*{Assembling and artifact filtering}\label{sec:artifact-filtering}
 The results of the previous stages can be described as a \emph{collection of sets} of \emph{geometric quadrilaterals (quads)}, each one encoded as an array of quadruples of integer indices, pointing to the linear array of grid vertices associated to the image block $\mathcal{B}$.
@@
 Besides to put each partial surface $\mathcal{S}(B) = (\texttt{V}_B, \texttt{FV}_\sigma)$ in the global coordinate system of the image, the present stage must eliminate the redundant boundary features possibly generated at the edges of the partial surface $\mathcal{S}(B)$ within each block $\mathcal{B}$ such that $B \cap \mathcal{I} \not= \emptyset$:
+
 \begin{enumerate}
@@
 
-\item remove both instances of \emph{double quads} generated by \texttt{Lar} software at the block boundaries (see Figure~\ref{fig:blockboundary}).
+\item remove both instances of \emph{double quads} generated by \texttt{Lar} software at the block boundaries 
+(see Fig.~\ref{fig:blockboundary}). 
+% TODO fix reference probably uncomment next line to link it to Figure 3
+% (see Figure~\ref{fig:example}). 
 They are artifacts generated by the decomposition of the whole image into a number of blocks of tractable size.
@@
 It will consist in arranging each block with all three dimensions decreased-increased by one, so that each 2-adjacent pair of blocks will be covering each other for a full side extent of blocks of depth one.
-The details of this \emph{artifact filtering} are elucidated in Section~\ref{sec:covering}.
+% The details of this \emph{artifact filtering} are elucidated in  section~\ref{sec:covering}.
+% TODO fix the reference
+% TODO is there such a section?
 
@@
 
-\paragraph{Smoothing} 
+\subsubsection*{Smoothing}\label{sec:smoothing}
 The final smoothing of the generated surfaces cannot be performed block-wise, since this would introduce smoothing artifacts at the block boundaries.
-Anyway, the Taubin smoothing~\cite{} can be performed in parallel, since for each vertex in the final surface (except eventually the ones on the image $\mathcal{I}$ boundaries) it essentially consist in computing a new position as a proper average of its neighborhood vertices, i.e.~by applying a discrete Laplacian operator.
-Some appropriate set workers may so be assigne the task of generating iteratively a new position for the vertices they take cure of.
+Anyway, the Taubin smoothing~\cite{Taubin1995} can be performed in parallel, since for each vertex in the final surface (except eventually the ones on the image $\mathcal{I}$ boundaries) it essentially consist in computing a new position as a proper average of its neighborhood vertices, i.e.~by applying a discrete Laplacian operator.
+Some appropriate set of workers may so be assigned the task of generating iteratively a new position for the vertices they take cure of.
 In particular, we have:
@@
 \centering
-\includegraphics[width=0.6\textwidth]{figs/bm_size_tesla.pdf} 
+\begin{subfigure}{0.495\textwidth}
+\includegraphics[width=0.99\textwidth]{figs/bm_size_tesla.pdf} 
 %\includegraphics[scale=1]{input/ircad_comparison.pdf} 
-\caption{Time requirements of LAR-SURF filter used on artificial with different size of boundary matrix}
+\subcaption{Time requirements of the \textsc{lar-surf} filter used on artificial volumetric data with various size of boundary matrix}
+% \caption{Time requirements of LAR-SURF filter used on artificial with different size of boundary matrix}
 \label{fig:bm_size_tesla}
+\end{subfigure}
+\begin{subfigure}{0.495\textwidth}
+
+\includegraphics[width=0.99\textwidth]{figs/ircad_comparison.pdf} 
+%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
+\subcaption{Time requirements of the LAR-SURF filter and Marching Cubes on Ircad dataset. Error bars shows the 95\% confidence interval}
+\end{subfigure}
+\caption{Performance analysis of the \textsc{lar-surf} filter.}
+
 \end{figure}
 
-To compare time requirements of LAR-SURF with Marching Cubes implemented in Python we performed experiment on Ircadb dataset \cite{ircad}.
-Dataset contain 20 Computed Tomography images (see table \ref{tab:ircad1}) with xy-resolution from 0.56 mm to 0.87 mm and z-resolution from 1.0 mm to 4.0 mm.
+To compare time requirements of LAR-SURF with Marching Cubes implemented in Python we performed experiment on Ircadb dataset \cite{ircadb}.
+Dataset contain 20 Computed Tomography images (see table \ref{tab:ircad2}) with xy-resolution from 0.56 mm to 0.87 mm and z-resolution from 1.0 mm to 4.0 mm.
 The number of slices is each series varies from 74 to 260 and size of each slice is $512\times512$.
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=0.99\textwidth]{figs/ircad_comparison.pdf} 
-%\includegraphics[scale=1]{input/ircad_comparison.pdf} 
-\caption{Time requirements of LAR-SURF filter and Marching Cubes on Ircad dataset. Error bars shows the 95\% confidence interval}
-\label{fig:ircad_comparison}
-\end{figure}
+% \begin{figure}
+% \centering
+% \includegraphics[width=0.99\textwidth]{figs/ircad_comparison.pdf} 
+% %\includegraphics[scale=1]{input/ircad_comparison.pdf} 
+% \caption{Time requirements of LAR-SURF filter and Marching Cubes on Ircad dataset. Error bars shows the 95\% confidence interval}
+% \label{fig:ircad_comparison}
+% \end{figure}

### 2020-04-26 src/background.tex
submission

@@
 
-\subsection{Representation Schemes}\label{sec:schemes}
+\subsection{Representation Scheme}\label{sec:schemes}
 
@@
 Solid pointsets (i.e., `$r$-sets') are defined~\cite{Requicha:1980:RRS:356827.356833} as compact (bounded and closed) regular and semianalytic subsets of the $d$-space.
-A large number of representation schemes were defined in the past forty years, including the two main classes of (a) \emph{boundary representations} (`$B$-reps'), where the solid model is represented through a representation of its boundary elements, i.e.~faces, edges and vertices, and (b) \emph{decompositive/enumerative representations}, that are a decomposition of either the object or the embedding space, respectively, into a well-defined \emph{cellular complex}.
+A large number of representation schemes were defined in the past forty years, including the two main classes of (a) \emph{boundary representations} (`$B$-reps'), where the solid model is represented through a representation of its boundary elements, i.e.~faces, edges and vertices, and (b) \emph{decompositive/enumerative representations} \cite{Requicha:1980:RRS:356827.356833}, that are a decomposition of either the object or the embedding space, respectively, into a well-defined \emph{cellular complex}.
 In particular, a boundary representation provides a cellular decomposition of the object's boundary into \emph{cells} of dimension zero (vertices), one (edges), and two (faces).
-Medical imaging can be classified as the \emph{enumerative representation} of cellular decompositions of organs and tissues of interest, in particular, as subsets \emph{of 3D volume elements} (voxels) from the 3D image.
+Medical imaging can be classified as the \emph{enumerative representation} of cellular decompositions of organs and tissues of interest \cite{Paoluzzi2016}, in particular, as subsets \emph{of 3D volume elements} (voxels) from the 3D image.
 
@@
 
-The \emph{Linear Algebraic Representation} (\textsc{lar}), introduced in~\cite{Dicarlo:2014:TNL:2543138.2543294}, aims to represent the \emph{chain complex}~\cite{TSAS} generated by a piecewise-linear \emph{geometric complex} embedded either in 2D or in 3D.
+The \emph{Linear Algebraic Representation} (\textsc{lar}), introduced in~\cite{Dicarlo:2014:TNL:2543138.2543294}, aims to represent the \emph{chain complex}~\cite{TSAS,DiCarlo2009} generated by a piecewise-linear \emph{geometric complex} embedded either in 2D or in 3D.
 In a few words, it gets a minimal characterization of geometry and topology of a cellular complex, i.e.~the embedding mapping $\mu : C_0 \to \E^d$ of 0-cells (vertices), as well a description of $(d-1)$-cells as subsets of vertices, and is able to return the whole chain complex
@@
 It is possible to show that all the interesting relations of incidence/adjacency between cells of different dimensions can be both computed and efficiently queried by pairwise computing some matrix products, with one of terms possibly transposed, using only the two boundary and coboundary operator matrices $[\partial_p]$ and $[\delta_p]$, and where $[\delta_p] = [\partial_p^\top]$.
-We~may also show that such matrices are \emph{very sparse}, with their sparseness growing rapidly with the dimensions (see Section~\ref{sec:block}).
+We~may also show that such matrices are \emph{very sparse}, with their sparseness growing rapidly with the dimensions (see Section~\ref{sec:brick}).
 The pattern of non-zeros in matrix $[\partial_3]$ corresponding to a brick of shape $(4,4,4)$ is given in Fig.~\ref{fig:boundary_matrix_4x4x4}.
@@
    \centering
-   \includegraphics[width=0.75\linewidth]{figs/boundary_matrix_4x4x4.pdf} 
+   \includegraphics[width=0.75\linewidth]{src/figs/boundary_matrix_4x4x4_desc.png} 
    \caption{

### 2020-04-26 src/conclusion.tex
submission

@@
 
-% TODO from abstract
+% % TODO from abstract
+% We introduced a Julia implementation of an algebraic filter to extract from 3D medical images the
+% boundary surface of some specific image segment, described as a 3-chain of voxels. Translations from
+% Cartesian indices of cells to linearized indices, the computation of the sparse boundary matrices, and the
+% sparse matrix-vector multiplication are the main computational kernels of this approach. We may show
+% a good speed-up over marching-cubes algorithms. The existing implementation employs Julia's channels
+% for multiprocessing. Currently, the computational pipeline is being strongly improved to gain a greater
+% speed-up using native Julia implementation \texttt{CUDA.jl} of Nvidia programming platform 
+% % TODO cite Besard2017 [7]
+% , and the Julia's
+% \texttt{SuiteSparseGraphBLAS.jl} framework 
+% % TODO cite BULUK 2017 [8] 
+% for graph algorithms with the language of linear algebra. In
+% particular, we are extending its use pattern in order to work with general cellular complexes.
+
 We introduced a Julia implementation of an algebraic filter to extract from 3D medical images the
@@
 Cartesian indices of cells to linearized indices, the computation of the sparse boundary matrices, and the
-sparse matrix-vector multiplication are the main computational kernels of this approach. We may show
-a good speed-up over marching-cubes algorithms. The existing implementation employs Julia's channels
-for multiprocessing. Currently, the computational pipeline is being strongly improved to gain a greater
+sparse matrix-vector multiplication are the main computational kernels of this approach.
+
+The implementation of the \textsc{lar-surf} filter is available in the open-source repository and it can be installed using standard Julia package manager \cite{larsurf-github}.
+
+% NEW next two sentences are new
+We showed a good speed-up over marching-cubes algorithms. 
+The existing implementation employs Julia's channels
+for multiprocessing. 
+Our performance experiment showed an optimal size of the brick size.
+Parallelization makes a large portion of spared computational cost. 
+%
+Moreover, we expect additional improvement in the future because our approach is appropriate  for SIMD (Single Instruction, Multiple Data) hybrid architectures of CPUs and GPUs, since only the initial block setup of boundary matrix and image slices, as well the final collection of computed surface portions, require inter-process communication.
+
+\sloppy{
+Currently, the computational pipeline is being strongly improved to gain a greater
 speed-up using native Julia implementation \texttt{CUDA.jl} of Nvidia programming platform 
 % TODO cite Besard2017 [7]
-, and the Julia's
+% \cite{Besard:2017}
+\cite{Besard2019}
+, and Julia's
 \texttt{SuiteSparseGraphBLAS.jl} framework 
+\cite{Buluc2017}
 % TODO cite BULUK 2017 [8] 
@@
 particular, we are extending its use pattern in order to work with general cellular complexes.
+}

### 2020-04-26 src/design.tex
submission

@@
-\section{Block-parametric design}\label{sec:filter}
+\section{Brick-parametric design}\label{sec:filter}
 
-\subsection{Block decomposition}\label{sec:block-decomposition}
+\subsection{Brick decomposition}\label{sec:brick-decomposition}
 
@@
     \centering
-   \includegraphics[width=0.5\linewidth]{figs/blocks} 
-   \caption{A possible block partitioning of a radiologic image. The evidenced 2D block, of size $n^d=64^2$, is sliced by $\B([2,1,64]) = Image([128:172],[64:128)]$}
-   \label{fig:blocks}
+    \begin{subfigure}{0.49\textwidth}
+       \centering
+       \includegraphics[width=0.70\linewidth]{src/figs/blocks.png} 
+       \subcaption{A possible brick partitioning of a radiologic image. The evidenced 2D brick, of size $n^d=64^2$, is sliced by $\B([2,1,64]) = Image([128:172],[64:128)]$}
+       \label{fig:bricks}
+    \end{subfigure} 
+    \begin{subfigure}{0.49\textwidth}
+        \centering
+        \includegraphics[width=0.99\linewidth]{figs/grid.pdf} 
+        \subcaption{Faces on the brick boundary}
+        \label{fig:brickboundary}
+    \end{subfigure} 
+    \caption{Brick decomposition}
 \end{figure}
 
-Figure~\ref{fig:blocks} shows the block decomposition in a 2D image, with positive integers $(u,v)$ giving the lateral sizes of image.
-Note that block sides do not necessarily correspond to image edges.
+Figure~\ref{fig:bricks} shows the brick decomposition in a 2D image, with positive integers $(u,v)$ giving the lateral sizes of image.
+Note that brick sides do not necessarily correspond to image edges.
 
-\subsection{Block operator }\label{sec:block}
+\subsection{Brick operator }\label{sec:brick}
 
 \subsubsection*{Chain coordinates }\label{sec:chain-coords}
-We are going to treat each image block independently from each other.
-Hence we map each image block $\B(i,j,k,n)$ to the linear \emph{chain} space $C_2$ of dimension $n\times n\times n$, using coordinate vectors $c\in \mathbf{2}^{n^d} := \{0,1\}^{n^d}$, where the basis element $c \in C_2$ is mapped via Cartesian-to-linear map to the binary vector
+We are going to treat each image brick independently from each other.
+Hence we map each image brick $\B(i,j,k,n)$ to the linear \emph{chain} space $C_2$ of dimension $n\times n\times n$, using coordinate vectors $c\in \mathbf{2}^{n^d} := \{0,1\}^{n^d}$, where the basis element $c \in C_2$ is mapped via Cartesian-to-linear map to the binary vector
 \[
@@
 
-Therefore, each pixel (or voxel) in a block image will be seen as a basis binary vector in $C_2$, and each subset of image elements, as the corresponding binary vector in $C_2$, with many ones as the cardinality of the subset.
+Therefore, each pixel (or voxel) in a brick image will be seen as a basis binary vector in $C_2$, and each subset of image elements, as the corresponding binary vector in $C_2$, with many ones as the cardinality of the subset.
 
 \subsubsection*{Boundary operator }\label{sec:boundary-operator}
-For a fixed block size $n$, the boundary operator $\partial_d : C_d\to C_{d-1}$, with $d\in\{2,3\}$, will be constructed once and for all using the algorithm given in~
+For a fixed brick size $n$, the boundary operator $\partial_d : C_d\to C_{d-1}$, with $d\in\{2,3\}$, will be constructed once and for all using the algorithm given in~
 \cite{TSAS}, 
@@
 The dimension of $C_{d-1}$, and therefore the number of rows of $[\partial_d]$ matrix is $d\,(n^{d-1}+n^{d}) = d\,n\,(1+n)^{d-1}$.
-The number of column equates the number of basis elements of $C_d$, i.e.~the number $n^d$ of block elements.
+The number of column equates the number of basis elements of $C_d$, i.e.~the number $n^d$ of brick elements.
 
@@
 \]
-In conclusion, for block size $n=64$, the matrix $[\partial_d]$ requires for 2D images $9\times 64^2=36,864$ memory elements, and for 3D images $13\times 64^3=3,407,872$ memory elements.
+In conclusion, for brick size $n=64$, the matrix $[\partial_d]$ requires for 2D images $9\times 64^2=36,864$ memory elements, and for 3D images $13\times 64^3=3,407,872$ memory elements.
 Counting the bytes for the standard implementation of a sparse binary matrix (1 byte for values and 8 bytes for indices) we get $(18d+8)n^d$ bytes, giving $176$\,KB for 2D and $15.872$\,MB for 3D.
 
-% TODO fix the name
-\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
-   \centering
-   \includegraphics[width=0.4\linewidth]{figs/grid.pdf} 
-   \caption{Faces on the brick boundary}
-   \label{fig:blockboundary}
-\end{figure}
+% \begin{figure}[htbp] %  figure placement: here, top, bottom, or page
+%   \centering
+%   \includegraphics[width=0.4\linewidth]{figs/grid.pdf} 
+%   \caption{Faces on the brick boundary}
+%   \label{fig:brickboundary}
+% \end{figure}
 
-\subsection{Block boundary mapping}\label{sec:block-mapping}
+\subsection{Brick boundary mapping}\label{sec:brick-mapping}
 
@@
 Let us call \emph{segment} the bulk content $S$ of interest within the input 3D image of size $(u,v,w)$. Our aim is to compute the segment boundary $\partial_3 S$. 
-First we set the size $n$ of the block, in order to decompose the input $Image(u,v,w)$ into a fair number of blocks
+First we set the size $n$ of the brick, in order to decompose the input $Image(u,v,w)$ into a fair number of bricks
 \[
@@
 
-\subsection{Block-level parallelism}\label{sec:block-parallelism}
-In the computational pipeline introduced in this paper, several steps can be efficiently performed in parallel at the image-block level, depending on the embarrassingly data-parallel nature of the problem.
+\subsection{Brick-level parallelism}\label{sec:brick-parallelism}
+In the computational pipeline introduced in this paper, several steps can be efficiently performed in parallel at the image-brick level, depending on the embarrassingly data-parallel nature of the problem.
 In particular, little effort is needed to separate the problem into a several of parallel tasks $S_{i,j,k}$, using multiarray slicing.
-The granularity of parallelism, depending on the block size $n$, is further enforced by the computation of a single boundary matrix $[\partial_d(n)]$ depending on $n$, so that the initial communication cost of broadcasting the matrix to nodes can be carefully controlled, and finely tuned depending on the system architecture.
-The whole approach is appropriate  for SIMD (Single Instruction, Multiple Data) hybrid architectures of CPUs and GPUs, since only the initial block setup of boundary matrix and image slices, as well the final collection of computed surface portions, require inter-process communication.
+The granularity of parallelism, depending on the brick size $n$, is further enforced by the computation of a single boundary matrix $[\partial_d(n)]$ depending on $n$, so that the initial communication cost of broadcasting the matrix to nodes can be carefully controlled, and finely tuned depending on the system architecture.
+The whole approach is appropriate  for SIMD (Single Instruction, Multiple Data) hybrid architectures of CPUs and GPUs, since only the initial brick setup of boundary matrix and image slices, as well the final collection of computed surface portions, require inter-process communication.

### 2020-04-26 src/examples.tex
submission

@@
 \centering
-\includegraphics[width=0.45\textwidth]{src/figs/ircad01_segmentation_65.png}
-\includegraphics[width=0.45\textwidth]{src/figs/ircad01_liver_tricolore_01.png}
-\includegraphics[width=0.45\textwidth]{src/figs/ircad01_porta_blue_01.png}
-\includegraphics[width=0.45\textwidth]{src/figs/nrn10_200_pink_02.png}
+\includegraphics[width=0.40\textwidth]{src/figs/ircad01_segmentation_65.png}
+\includegraphics[width=0.40\textwidth]{src/figs/ircad01_liver_tricolore_01.png}
+\includegraphics[width=0.40\textwidth]{src/figs/ircad01_porta_blue_01.png}
+\includegraphics[width=0.40\textwidth]{src/figs/nrn10_200_pink_02.png}
 % \includegraphics[width=0.4\textwidth]{figs/liver_01_red_3.png} 
@@
 \caption{
-Liver structures. The upper left image shows one slice  with segmented organs from  the Ircad dataset \cite{ircad}.
-The other images show the triangulated isosurfaces of the macroscopic and microscopic strucures of the liver extracted with \textsc{lar-surf} algorithm. 
-On the upper right image can be seen the wire frame model of the human liver along with portal vein and hepatic artery. 
-% Both right images show the Portal Vein from the same Ircad dataset \cite{ircad} 
+Liver structures. The upper left image shows one slice  with segmented organs from  the Ircad dataset \cite{ircadb}.
+The other images show the triangulated isosurfaces of the macroscopic and microscopic structures of the liver extracted with \textsc{lar-surf} algorithm. 
+The wireframe model of the human liver along with the portal vein and hepatic artery can be seen on the upper right image. 
 The left bottom image shows the detail of the portal vein  
@@
 On the bottom right image is shown the microvasculature of a pig liver based on corrosion cast prepared by Eberlova
-\cite{eberlova2017use}. The size of the speciment is 0.936 $[mm]$ along each axis and the resolution of the Micro-CT data is 4.682 $\mu{}m$.
+\cite{eberlova2017use}. The size of the specimen is 0.936 $[mm]$ along each axis and the resolution of the Micro-CT data is 4.682 $\mu{}m$.
 } \label{fig:example_liver_macro_micro}
 \end{figure}
+% Both right images show the Portal Vein from the same Ircad dataset \cite{ircad}
 
@@
 
-The implementation of our algorithm is available in LarSurf package. 
-Use of this package can be seen on listings \ref{lst:example1} where the liver segmentation with 2865131 voxels from
+% The implementation of our algorithm is available in LarSurf package. 
+Use of the \texttt{LarSurf.jl} package can be seen on listings \ref{lst:example1} where the liver segmentation with 2865131 voxels from the
 Ircad dataset is used as an input for our surface extraction algorithm. The size of 3D volumetric 
@@
 and the voxel resolution is $1.6\times0.57\times0.57$ [mm]. 
-The output surface model created by  182124 triangles and the number of vertices is 90822. The visualization can 
-be seen on figure \ref{fig:example_liver_macro_micro}. 
-The portal vein surface extraction can be performed with a small change of input path in code.
-The The 3D image resolution is the same. 
-It can be seen in the same figure. 
-The number of input voxels is 103533. 
-The output surface is created by 90822 vertices and 182124 triangles.
+The output liver surface model is formed by  182124 triangles and the number of vertices is 90822. 
+The visualization can 
+be seen in Fig. \ref{fig:example_liver_macro_micro}.
 
@@
 
-The right image of figure \ref{fig:example_liver_macro_micro} is surface of a microvasculature of pig liver. 
-Volumetric image is based on Micro-CT data of corrosion casts of pig liver.
+The portal vein surface extraction can be performed with a small change of input path in code.
+The 3D image resolution is the same. 
+% It can be seen in the same figure. 
+The number of input voxels is 103533. 
+The output surface is created by 90822 vertices and 182124 triangles.
+
+The right bottom image of Fig. \ref{fig:example_liver_macro_micro} is the surface of the microvasculature of pig liver. 
+The volumetric image is based on Micro-CT data of corrosion casts of pig liver.
 \cite{eberlova2017use}.
-The size of the visualized data is $100\times100\times100$ voxels and size of voxel is 4.682 $\mu{}m$.
-The number of of triangles is 
-544784 and number of vertices is 272826.
+The size of the visualized data is $100\times100\times100$ voxels and the size of the voxel is 4.682 $\mu{}m$.
+The number of triangles is 
+544784 and the number of vertices is 272826.
 

### 2020-04-26 src/examples_appendix.tex
submission

\vspace{10pt}
\noindent\underline{
Boundary matrices for grids of cubes:
}\vspace{0.2em}\newline 
% NOTE this is new 
We give here the full Julia code for the algebraic computation of $\partial_3$ matrix, for a very little grid of unit
3-cubes. Due to the simplicity of the cells (voxels = cubes), a sufficient (geom,top) pair is given below
as (\texttt{V,CV}), where \texttt{CV} is an array of arrays of \texttt{Float64} indices of grid cubes.

% julia> V
% 3x24 Array{Float64,2}:
% julia> CV
% 6-element Array{Array{Int64,1},1}:
\begin{Verbatim}[fontsize=\footnotesize]
julia> V, CV = Lar.cuboidGrid([3,2,1])
julia> V
0.0 0.0 0.0 0.0 0.0 0.0 1.0 1.0 1.0 1.0 1.0 1.0 2.0 2.0 2.0 2.0 2.0 2.0 3.0 3.0 3.0 3.0 3.0 3.0
0.0 0.0 1.0 1.0 2.0 2.0 0.0 0.0 1.0 1.0 2.0 2.0 0.0 0.0 1.0 1.0 2.0 2.0 0.0 0.0 1.0 1.0 2.0 2.0
0.0 1.0 0.0 1.0 0.0 1.0 0.0 1.0 0.0 1.0 0.0 1.0 0.0 1.0 0.0 1.0 0.0 1.0 0.0 1.0 0.0 1.0 0.0 1.0
julia> CV
[[ 1, 2, 3, 4, 7, 8, 9,10], [ 3, 4, 5, 6, 9,10,11,12], [ 7, 8, 9,10,13,14,15,16],
 [ 9,10,11,12,15,16,17,18], [13,14,15,16,19,20,21,22], [15,16,17,18,21,22,23,24]
\end{Verbatim}

\vspace{10pt}
\noindent\underline{
Face and Edge Data generation:
}\vspace{0.2em}\newline 
%TODO describe CV2FV and CV2EV
In the following, we provide the functions for generating the face data \texttt{FV} (vertex indices in faces) with function  \texttt{CV2FV} and edge data \texttt{EV} (vertex indices in edges) with function \texttt{CV2EV} from cell data \texttt{CV}. 
% In particular,  and  functions apply to all the 3-cells in \texttt{CV} the pattern of reference to vertices used by faces and edges of the single 3-cube:

% \footnotesize
%\small]
\begin{Verbatim}[fontsize=\footnotesize]
function CV2FV( v:: Array { Int64 } )
return faces = [[v[1], v[2], v[3], v[4]], [v[5], v[6], v[7], v[8]],
                [v[1], v[2], v[5], v[6]], [v[3], v[4], v[7], v[8]],
                [v[1], v[3], v[5], v[7]], [v[2], v[4], v[6], v[8]]]
end
function CV2EV( v:: Array { Int64 } )
return edges = [[v[1],v[2]], [v[3],v[4]], [v[5],v[6]], [v[7],v[8]], [v[1],v[3]], [v[2],v[4]],
                [v[5],v[7]], [v[6],v[8]], [v[1],v[5]], [v[2],v[6]], [v[3],v[7]], [v[4],v[8]]]
end
\end{Verbatim}

\vspace{10pt}
\noindent\underline{
Characteristic matrices:
}\vspace{0.2em}\newline 
The function \texttt{K} transforms an array of arrays ($\mathtt{VV, EV, FV, CV}$) into a sparse binary characteristic matrix
($\mathtt{M_0, M_1, M_2, M_3}$). A Julia sparse matrix needs three arrays I, J, Vals of rows, columns, values of non-zeros:

% VV = [[v] for v=1:size(V, 2)]
% [1],[2],[3],[4],[5],[6],[7],[8],[9],[10],[11],[12],[13],[14],[15], [16],[17],
% [18],[19],[20],[21],[22],[23],[24]
% [1],[2],[3],[4],[5],[6],[7],[8],[9],[10],[11],[12],[13],[14],[15], [16],[17],
% [18],[19],[20],[21],[22],[23],[24]
\begin{Verbatim}[fontsize=\footnotesize]
VV = [[v] for v=1:size(V, 2)];
FV = collect(Set{Array{Int64,1}}(cat(map(CV2FV, CV))))
[[13,15,19,21], [1,2,3,4], [7,9,13,15], [13,14,15,16], [7,8,13,14], [1,2,7,8], [2,4,8,10], [7,8,9,10], 
 [3,5,9,11], [8,10,14,16], [15,16,21,22], [9,11,15,17], [3,4,5,6], [17,18,23,24], [11,12,17,18], 
 [1,3,7,9], [3,4,9,10], [9,10,15,16], [4,6,10,12], [13,14,19,20], [9,10,11,12], [15,16,17,18], 
 [19,20,21,22], [15,17,21,23], [16,18,22,24], [21,22,23,24], [10,12,16,18], [5,6,11,12], [14,16,20,22]]

EV = collect(Set{Array{Int64,1}}(cat(map(CV2EV, CV))))
[[15,17], [16,22], [6,12], [17,23], [18,24], [4,10], [3,4], [13,15], [11,12], [9,15], [13,19],
 [1,7], [5,11], [5,6], [12,18], [8,14], [15,21], [17,18], [1,3], [2,4], [16,18], [2,8], [21,23],
 [20,22], [1,2], [14,16], [10,16], [13,14], [19,21], [7,13], [9,10], [23,24], [11,17], [21,22],
 [3,9], [3,5], [9,11], [7,9], [14,20], [7,8], [22,24], [19,20], [8,10], [15,16], [10,12], [4,6]]

function K(CV)
    I = vcat ( [ [k for h in CV[k]] for k =1: length (CV) ]...)
    J = vcat (CV ...)
    Vals = Int8 [1 for k=1: length (I)]
    return SparseArrays . sparse (I,J,Vals)
end
M0 = K(VV); M1 = K(EV); M2 = K(FV); M3 = K(CV)
\end{Verbatim}

\vspace{10pt}
\noindent\underline{
Boundary matrices:
}\vspace{0.2em}\newline 
The boundary matrices between non-oriented chain spaces are computed by sparse matrix multiplication
followed by matrix filtering, produced in Julia by the broadcast of vectorized integer division ($.\div$):

% TODO this is strange the M_1 is M_1 after multiplication?
$\partial_1 =  \mathtt{M_0} * \mathtt{M'_1} = \mathtt{M'_1}$

$\partial_2 =  \left(\mathtt{M_1} * \mathtt{M'_2}\right) .\div \mathtt{sum(M_1, dims=2)}$

$\partial_3 =  \left(\mathtt{M_2} * \mathtt{M'_3}\right) .\div \mathtt{sum(M_2, dims=2)}$

### 2020-04-26 src/implementation.tex
submission

@@
 The computer code is implemented in Julia language~\cite{BEKS14} according to the workflow described below, whose stages are parallelized and/or optimized in various ways.
+The workflow scheme can be seen on Fig.
+\ref{fig:schema}.
+The implementation is available on GitHub \cite{larsurf-github} and our \texttt{LarSurf.jl} package can be installed using a standard Julia package register.
 
@@
 
-\item input of 3D medical image $\mathcal{I}$ with \emph{shape} $(\ell_1, \ell_2, \ell_3)$, such that: $\mathcal{I} = [\ell_1]\times[\ell_2]\times[\ell_3]$, where $[\ell_k] = [1,2,\ldots,\ell_k]$;
+\item the input of 3D medical image $\mathcal{I}$ with \emph{shape} $(\ell_1, \ell_2, \ell_3)$, such that: $\mathcal{I} = [\ell_1]\times[\ell_2]\times[\ell_3]$, where $[\ell_k] = [1,2,\ldots,\ell_k]$;
 
@@
 
-\item depending on the above, best decision for \emph{size} of 3D image block (or brick) $\mathcal{B}$.
-With defaults $\emph{size}=64$,  the \emph{number of bricks} will be $n=\ceil{\ell_1/size}\times\ceil{\ell_2/size}\times\ceil{\ell_3/size}$.
+\item depending on the above, best decision for the \emph{size} of 3D image brick $\mathcal{B}$.
+With default $\emph{size}=64$,  the \emph{number of bricks} will be $n=\ceil{\ell_1/size}\times\ceil{\ell_2/size}\times\ceil{\ell_3/size}$.
 Hence the default number of bricks is $n=256$, for standard images $512\times 512\times 256$;
@@
 \sloppy{
-\item computation of the Julia's sparse boundary matrix $[\partial_B]$, returning a value of type \texttt{SparseMatrixCSC\{Int8\}\{Int64\}}, where \texttt{Int8} and \texttt{Int64} are the types for values and indices, respectively, stored by Compressed Sparse Column (CSC) format; the storage of $[\partial_B]$ (for $\emph{size}$ = 64) requires about 45 MB;
+\item computation of Julia's sparse boundary matrix $[\partial_B]$, returning a value of type \texttt{SparseMatrixCSC\{Int8\}\{Int64\}}, where \texttt{Int8} and \texttt{Int64} are the types for values and indices, respectively, stored by Compressed Sparse Column (CSC) format; the storage of $[\partial_B]$ (for $\emph{size}$ = 64) requires about 45 MB;
 }
@@
 
-\item distribution of matrix $[\partial_B]$, of default size $45$ MB, to all available nodes/cores (Julia workers) using the Julia macro \texttt{@eveywhere}.
+\item distribution of matrix $[\partial_B]$, of default size $45$ MB, to all available nodes/cores (Julia workers), using the Julia macro \texttt{@eveywhere}.
 
@@
 
-\item enqueing the job (as a sequence of integer positions for the non-zeros image elements aligned in a memory buffer of proper \texttt{Channel} type).
+\item enqueuing the job (as a sequence of integer positions for the non-zeros image elements aligned in a memory buffer of proper \texttt{Channel} type).
 \end{enumerate}
@@
 
-In coordinates local to $\mathcal{B}$, once fixed an ordering from Cartesian to linear coords, this vector is represented by a \emph{binary array} of length $size^3$.
+In coordinates local to $\mathcal{B}$, once fixed an ordering from Cartesian to linear coordinates, this vector is represented by a \emph{binary array} of length $size^3$.
 With $size=64$, we have $64^3=262144$,  with a non-zero value (i.e.~$1$) for each foreground voxel in $\mathcal{S}(B)$.
-Therefore, the coded segment portion $\mathcal{S}(B)$, results with a space occupancy of about $262$ KB if encoded as a full array (i.e.~including the zero values).
+Therefore, the coded segment portion $\mathcal{S}(B)$ results with a space occupancy of about $262$ KB if encoded as a full array (i.e.~including the zero values).
 Whether encoded as a sparse vector, its space occupancy will correspondingly decrease.
@@
 \item prepare subsequences of such data vectors (non-zero linear row indices), in order to feed efficiently the available processor threads.
-In case of presence of one/more GPUs, a smaller size of the block---and hence of the boundary matrix and the encoded 3-chain vectors---and then much higher vector numbers, are preferable for speed.
+In case of the presence of one/more GPUs, a smaller size of the block---and hence of the boundary matrix and the encoded 3-chain vectors---and then much higher vector numbers, are preferable for speed.
 
@@
 \subsubsection*{SpMM Multiplication}\label{SpMM-multiplication}
-According to the current literature~\cite{} it is more convenient to execute SpMV (sparse matrix-vector) multiplications than SpMSpV (sparse matrix-sparse vector) multiplications.
-Since we have 256 such jobs (one multiplication per block) to perform in the standard setting of the algorithm\footnote{
-Size of block $64^3$; size of image $512^2\times 256$.
-}, or more in case of either smaller blocks or image greater than the standard one, this stage must be evidently parallelized and carefully tuned, possibly by using the GPU, if available.
+According to the current literature~
+\cite{Buluc2017} it is more convenient to execute SpMV (sparse matrix-vector) multiplications than SpMSpV (sparse matrix-sparse vector) multiplications.
+Since we have 256 such jobs (one multiplication per block) to perform in the default setting of the algorithm (
+thize of the block $64^3$; the size of the image $512^2\times 256$), or more in case of either smaller blocks or image greater than the standard one, this stage must be evidently parallelized and carefully tuned, possibly by using the GPU, if available.
 \begin{enumerate}
@@
 
-\item conversion from position of ones (or non-zeros) in the 2-chain to linear indices of rows;
+\item conversion from the position of ones (or non-zeros) in the 2-chain to linear indices of rows;
 
@@
 
-A Julia's vectorized pipeline dataflow seems the more appropriate implementation model for the job of each worker.
+Julia's vectorized pipeline dataflow seems the more appropriate implementation model for the job of each worker.
 
@@
 In other words, \emph{all quads} of \textbf{each job} are now given in the \textbf{same} \emph{local coordinates}.
-Besides to put each partial surface $\mathcal{S}(B) = (\texttt{V}_B, \texttt{FV}_\sigma)$ in the global coordinate system of the image, the present stage must eliminate the redundant boundary features possibly generated at the edges of the partial surface $\mathcal{S}(B)$ within each block $\mathcal{B}$ such that $B \cap \mathcal{I} \not= \emptyset$:
+Besides putting each partial surface $\mathcal{S}(B) = (\texttt{V}_B, \texttt{FV}_\sigma)$ in the global coordinate system of the image, the present stage must eliminate the redundant boundary features possibly generated at the edges of the partial surface $\mathcal{S}(B)$ within each block $\mathcal{B}$ such that $B \cap \mathcal{I} \not= \emptyset$:
 
@@
 \item remove both instances of \emph{double quads} generated by \texttt{Lar} software at the block boundaries 
-(see Fig.~\ref{fig:blockboundary}). 
+(see Fig.~\ref{fig:brickboundary}). 
 % TODO fix reference probably uncomment next line to link it to Figure 3
@@
 a smart strategy of removal of such artifacts may be used, which does not require any sorting nor searching on the assembled array of quads.
-It will consist in arranging each block with all three dimensions decreased-increased by one, so that each 2-adjacent pair of blocks will be covering each other for a full side extent of blocks of depth one.
+It will consist in arranging each block with all three dimensions decreased-increased by one so that each 2-adjacent pair of blocks will be covering each other for a full side extent of blocks of depth one.
 % The details of this \emph{artifact filtering} are elucidated in  section~\ref{sec:covering}.
@@
 \subsubsection*{Smoothing}\label{sec:smoothing}
-The final smoothing of the generated surfaces cannot be performed block-wise, since this would introduce smoothing artifacts at the block boundaries.
-Anyway, the Taubin smoothing~\cite{Taubin1995} can be performed in parallel, since for each vertex in the final surface (except eventually the ones on the image $\mathcal{I}$ boundaries) it essentially consist in computing a new position as a proper average of its neighborhood vertices, i.e.~by applying a discrete Laplacian operator.
-Some appropriate set of workers may so be assigned the task of generating iteratively a new position for the vertices they take cure of.
+The final smoothing of the generated surfaces cannot be performed block-wise since this would introduce smoothing artifacts at the block boundaries.
+Anyway, the Taubin smoothing~\cite{Taubin1995} can be performed in parallel, since for each vertex in the final surface (except eventually the ones on the image $\mathcal{I}$ boundaries) it essentially consists in computing a new position as a proper average of its neighborhood vertices, i.e.~by applying a discrete Laplacian operator.
+Some appropriate sets of workers may so be assigned the task of generating iteratively a new position for the vertices they take cure of.
 In particular, we have:
@@
 
-The size of boundary matrix is critical parameter of LAR-SURF method.
-To determine optimal size of boundary matrix the experiment on artifical data was performed (Fig.
+\subsubsection*{Boundary matrix size}
+The size of the boundary matrix is a critical parameter of the \textsc{lar-surf} method.
+To determine optimal size of boundary matrix the experiment on artificial data was performed (Fig.
 \ref{fig:bm_size_tesla}).
-Size of experimental data is set to $512\times512\times512$ and it is derived from typical size of Computed Tomography medical images.
-Computation is done on Tesla DGX-1 machine.
+The size of experimental data is set to $512\times512\times512$ and it is derived from a typical size of Computed Tomography medical images.
+Computation is done on the Tesla DGX-1 machine.
+
+According to the experiment the fastest computation is with the boundary matrix with size $64\times64\times64$.
+This is an expected result.
+The larger boundary matrix is too big to fit in CPU's cache memory.
 
@@
 %\includegraphics[scale=1]{input/ircad_comparison.pdf} 
-\subcaption{Time requirements of the \textsc{lar-surf} filter used on artificial volumetric data with various size of boundary matrix}
+\subcaption{Time requirements of the \textsc{lar-surf} filter used on artificial volumetric data with various sizes of boundary matrix}
 % \caption{Time requirements of LAR-SURF filter used on artificial with different size of boundary matrix}
@@
 %\includegraphics[scale=1]{input/ircad_comparison.pdf} 
-\subcaption{Time requirements of the LAR-SURF filter and Marching Cubes on Ircad dataset. Error bars shows the 95\% confidence interval}
+\subcaption{Time requirements of the LAR-SURF filter and Marching Cubes on the Ircad dataset. Error bars show the 95\% confidence interval}
+\label{fig:ircad_comparison}
 \end{subfigure}
@@
 
-To compare time requirements of LAR-SURF with Marching Cubes implemented in Python we performed experiment on Ircadb dataset \cite{ircadb}.
+\subsubsection*{Comparison with the Marching Cubes algorithm}
+To compare the time requirements of \textsc{lar-surf} with Marching Cubes implemented in Python we performed an experiment on Ircadb dataset \cite{ircadb}.
 Dataset contain 20 Computed Tomography images (see table \ref{tab:ircad2}) with xy-resolution from 0.56 mm to 0.87 mm and z-resolution from 1.0 mm to 4.0 mm.
-The number of slices is each series varies from 74 to 260 and size of each slice is $512\times512$.
-Dataset contain manually segmented liver, portal vein and other structures.
-We performed surface extraction of liver with Marching Cubes and LAR-SURF.
-Time required for computation can be seen on figure \ref{fig:ircad_comparison}.
+The number of slices is each series varies from 74 to 260 and the size of each slice is $512\times512$. 
+The dataset contains manually segmented liver, portal vein, and other structures.
+We performed surface extraction of the liver with Marching Cubes and LAR-SURF.
+The time required for computation can be seen in Fig.
+\ref{fig:ircad_comparison}.
 
-Based on t-test with $\alpha=0.99$, $p=\input{input/ircad_pvalue}$ and 
+Based on the t-test with $\alpha=0.99$, $p=\input{input/ircad_pvalue}$ and 
 $s=\input{input/ircad_statistics}$ it can be shown that the mean of time consumed by LAR-SURF is significantly lower from time consumed by Marching Cubes.

### 2020-04-26 src/intro.tex
submission

@@
 A completely data-parallel algorithm, implemented in OpenCL that runs entirely on the GPU is presented in~\cite{Smistad12}. 
-A Linear Algebraic Representation approach, parallelized using the OpenCL framework on Linux, was discussed in~\cite{paodcvjcadanda2015}.
+A Linear Algebraic Representation approach, parallelized using the OpenCL framework on Linux, was discussed in~\cite{Paoluzzi2016}. % \cite{paodcvjcadanda2015}.
 
@@
 Our \textsc{lar-surf} (Linear Algebraic Representation Surface extraction) filter is based on basic algebraic topology and linear algebra, using linear spaces $C_p$ of chains (of cells) of dimension $0 \leq p \leq 3$ and the boundary matrix $[\partial_3] : C_3 \to C_2$.
+
+Input volumetric data are represented by a 3D voxel array and can be generated by segmentation computed tomography (upper left image on Fig.
+\ref{fig:example_liver_macro_micro}).
 A decomposition of the input volumetric data into small submatrices called \emph{bricks} is performed, then the binary coordinate vector of each interesting chain of voxels is generated, and its boundary is computed by matrix multiplication times the boundary matrix producing the binary representation of the boundary surface (the surface which defines the boundary).
